---
title: "clustering"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Hierarchical clustering of expression data observation from combined-experiment clustering: each experiment only has a small number of distinct "shapes" to the expression a happy medium between experiment-specific clustering and combined-experiment clustering is to describe each gene by its combination of expression "shapes" in each experiment

```{r}
options(stringsAsFactors = FALSE)
source("functions_for_figure_scripts.R")
load("data_files/Clustering_Counts.RData")
set.seed(23)

# Method:
# 1) Filter out genes with low var/mean ratio (low dispersion)
# 2) Correlation cluster remaining genes

# initial values for low var filter applied in corCluster (decided in following section)
var_thresh <- 3
```

## Clustering functions and unit tests

Bootstrapping function - runs clustering repeatedly on subsets of full clustering data and labels each gene by the cluster it is most frequently in (will be tested below corCluster)

```{r bootstrapClustering}
# helper function for corCluster
# runs nIter clusterings on random subsets of genes 
# returns the majority vote for each gene's cluster
bootstrapClustering <- function(.filtered_cts, .nClust,
                                .nIter = 100, .frac = 0.75) {
  clusterSubset <- function(.cts) {
    gene_idxs <- sample(c(1:nrow(.cts)), size = nrow(.cts)*.frac, replace = FALSE)
    frac_cts <- .cts[gene_idxs,]
    cor_mat <- frac_cts |> t() |> cor(use = "pairwise.complete.obs")
    tree <- hclust(as.dist(-cor_mat), method = "average")
    topBranchHeight <- sort(tree$height, decreasing = TRUE)[.nClust]
    tree_labels <- cutreeStatic(tree, cutHeight = topBranchHeight,
                                minSize = 1)
    output <- rep(NA, nrow(.cts))
    output[gene_idxs] <- tree_labels
    return(output)
  }
  label_list <- vector(mode = "list", length = .nIter)
  for (i in c(1:.nIter)) {
    label_list[[i]] <- clusterSubset(.filtered_cts)
  }
  labeldf <- purrr::reduce(label_list, .f = cbind)
  cat("finished iterations, nGenes x nIter:", dim(labeldf), "\n")
  rownames(labeldf) <- rownames(.filtered_cts)
  # aligning labels between iterations
  matched <- apply(labeldf, 2, matchLabels, reference = labeldf[,1],
                   ignoreLabels = NA)
  matched[matched > .nClust] <- NA # matchLabels adds new labels to any gene over pThreshold
  labels <- apply(matched, 1, \(x) {
    label <- names(sort(table(x, useNA = "no"), decreasing = TRUE))[1]
    if (length(label) == 0) {
      return(NA)
    }
    return(as.numeric(label))
  })
  return(labels)
}
```

Clustering function
given a count matrix where rows are genes and columns are samples/conditions, returns heirarchical cluster assignments as vector of length nGenes labels follow convention:
  1) NA = too lowly expressed (mean expression < .min_expr)
  2) 0 = too low variance (mean expression >= .min_expr & var expression < .min_var)
  3) 1 - .nClust = clusters of co-varying genes

```{r}
corCluster <- function(.cts, .nClust, .min_var = var_thresh,
                       .tree_too = FALSE, .gap_stat = FALSE, 
                       .just_counts = FALSE, .bootstrapIter = 100,
                       .bootstrapFrac = 0.75) {
  gene_idxs <- rownames(.cts)
  gene_means <- apply(.cts, 1, mean, na.rm = TRUE)
  gene_disps <- apply(.cts, 1, \(x) {var(x, na.rm = TRUE)/abs(mean(x, na.rm = TRUE))})
  gene_disps[gene_means == 0] <- 0 # avoiding divide by 0 NaN if gene is fully not expressed
  # identifying low var and low expr genes
  low_var_genes <- gene_idxs[gene_disps < .min_var]
  cat(length(low_var_genes), "low variance genes, assigned to 0 label\n")
  good_clustering_genes <- setdiff(x = gene_idxs, y = low_var_genes)
  filtered_cts <- .cts[good_clustering_genes,]
  if (.gap_stat) {
    cluster_fun <- function(x, k) {
      list(cluster = cutree(hclust(d = as.dist(-cor(t(x), use = "pairwise.complete.obs")), 
                                   method = "average"), 
                            k = k))}
    gap_stat <- clusGap(filtered_cts, FUNcluster = cluster_fun, K.max = 10, B = 5, method = "firstmax")
    return(gap_stat)
  }
  if (.just_counts) {
    return(filtered_cts)
  }
  # clustering
  if (sum(apply(filtered_cts, 1, \(x) {all(is.na(x))})) != 0) { # if any rows (genes) have all NA values, they will cause cor to fail below
    cat("NA genes in counts matrix, returning counts matrix only\n")
    return(cor_mat)
  }
  if (.bootstrapIter > 1) {
    tree_labels <- bootstrapClustering(filtered_cts, .nClust = .nClust,
                        .nIter = .bootstrapIter, .frac = .bootstrapFrac)
    labeldf <- tibble(gene_name = rownames(filtered_cts),
                      label = tree_labels) |> 
      bind_rows(tibble(gene_name = low_var_genes,
                       label = 0))
    # making single tree for .treeToo = TRUE
    cor_mat <- filtered_cts |> t() |> cor(use = "pairwise.complete.obs")
    tree <- hclust(as.dist(-cor_mat), method = "average")
  }
  if (.bootstrapIter <= 1) {
    cor_mat <- filtered_cts |> t() |> cor(use = "pairwise.complete.obs")
    tree <- hclust(as.dist(-cor_mat), method = "average") # negative because hclust expects distance mat --- the higher the ij value the LESS similar genes i and j are
    topBranchHeight <- sort(tree$height, decreasing = TRUE)[.nClust]
    tree_labels <- cutreeStatic(tree, cutHeight = topBranchHeight,
                                minSize = 1) # guaranteeing no 0 class
    cat("cutting tree at height", topBranchHeight, "into", length(unique(tree_labels)), "clusters\n")
    labeldf <- tibble(gene_name = tree$labels,
                      label = tree_labels) |> 
      bind_rows(tibble(gene_name = low_var_genes,
                       label = 0))
  }
  # matching each label to each gene
  # rearranging genes into their original order
  outdf <- left_join(tibble(gene_name = rownames(.cts)),
                     labeldf, by = "gene_name")
  if (!.tree_too) {
    return(outdf)
  }
  if (.tree_too) {
    return(list(tree = tree, df = outdf))
  }
}
```

Unit tests for boostrapClustering and corCluster
```{r}
### tests for corCluster/bootstrapClustering
# toy genes
# known co-expressed genes in LowN
# should split these into genes highest at TP2 (^ genes)
# and genes lowest at TP2 (v genes)
toy_idxs <- c("YBR172C", "YML015C", # v genes
              "YBR162W-A", "YKL196C", "YBR171W") # ^ genes (the excess of YBRs are coincidental, as I was just scrolling through that part of the module --- although the 171 172 W/C gene pair is probably overlapping)
toy_mat <- counts_list$par_LowN[toy_idxs,]
toydf <- makeDf(toy_mat, info, .join_by = "condition")
toyoutput <- corCluster(toy_mat, .nClust = 2, .tree_too = TRUE, 
                        .bootstrapIter = 0)
toydf <- left_join(toydf, toyoutput$df, by = "gene_name")
ggplot(toydf, aes(x = time_point_str, y = log2(expr + 1))) + 
  geom_line(aes(group = gene_name,
                color = gene_name)) +
  facet_wrap(~label)

# full dataset
toy_mat <- counts_list$cer_LowN
toydf <- makeDf(toy_mat, info, .join_by = "condition")
# no bootstrap
test_labels <- corCluster(toy_mat, .nClust = 4, .min_var = 2,
                          .bootstrapIter = 0)
# yes bootstrap
test_labels <- corCluster(toy_mat, .nClust = 4, .min_var = 2,
                          .bootstrapIter = 10)
plotdf <- toydf |> 
  group_by(gene_name, time_point_str) |> 
  summarise(expr = mean(expr)) |> 
  ungroup() |> 
  reframe(mean_expr = mean(expr), 
          sd_expr = sd(expr),
          expr = expr,
          time_point_str = time_point_str,
          .by = "gene_name") |> 
  mutate(scaled_expr = (expr - mean_expr)/sd_expr) |> 
  left_join(y = test_labels, by = "gene_name")
ggplot(plotdf, aes(x = time_point_str, y = scaled_expr)) + 
  geom_line(aes(group = gene_name)) +
  facet_wrap(~label)
```

Wrapper function for corCluster, repeatedly clusters list of environmental datasets

```{r}
# Wrapper function that converts named list of counts (from 1 experiment) into
# tidy counts data frame with nGenes * nConditions * length(counts list) number of rows
clusterCountsList <- function(.cts_list, .nClust = 4,
                              .bootstrapIter = 100,
                              .bootstrapFrac = 0.75,
                              .min_var = var_thresh,
                              .tree_too = FALSE,
                              .gap_stat = FALSE,
                              .just_counts = FALSE) {
  # giving each ortholog a unique name: YGR192C becomes YGR192C_cer
  cts <- map2(.cts_list, names(.cts_list), .f = \(x, nm) {
    nm <- gsub("_.*", "", nm)
    rownames(x) <- paste(nm, rownames(x), sep = "_")
    return(x)
  })
  dfs <- map(cts, makeDf, .info = info, .join_by = "condition")
  outdf <- purrr::reduce(dfs, bind_rows)
  if (all(unlist(map(cts, \(x) {all(colnames(x) == colnames(cts[[1]]))})))) {
    cat("counts are in same order, joining counts into single matrix\n")
    cts <- purrr::reduce(cts, .f = rbind)
  }
  else {
    cat("counts colnames don't match, returning counts list \n")
    return(cts)
  }
  # after collapsing replicates, samples should be in the same condition order
  # and able to be rbound
  if (.just_counts) {
    filtered_cts <- corCluster(cts, .nClust = .nClust, .min_var = .min_var,
                               .just_counts = TRUE, .bootstrapIter = .bootstrapIter,
                               .bootstrapFrac = .bootstrapFrac)
    return(filtered_cts)
  }
  if (!.tree_too) {
    labeldf <- corCluster(cts, .nClust = .nClust, .min_var = .min_var,
                          .bootstrapIter = .bootstrapIter,
                          .bootstrapFrac = .bootstrapFrac)
    outdf <- left_join(outdf, labeldf, by = "gene_name")
    return(outdf)
  }
  if (.tree_too) {
    output <- corCluster(cts, .nClust = .nClust, .min_var = .min_var,
                         .bootstrapIter = .bootstrapIter,
                         .bootstrapFrac = .bootstrapFrac,
                         .tree_too = TRUE)
    output$df <- left_join(outdf, output$df, by = "gene_name")
    return(output)
  }
}
# tests for clusterCountsList
# par is missing 1 condition versus cer
testout <- clusterCountsList(list("cer_LowN" = counts_list$cer_LowN,
                                  "par_LowN" = counts_list$par_LowN),
                             .bootstrapIter = 10, .min_var = 3)
testout$gene_name[1:10]
```
Visualize clustering

```{r plotClusters}
# given a dataframe with labels for each gene, returns a facet plot
# showing expression of each gene in each cluster (usually randomly 
# downsample option to save plotting computation)
plotClusters <- function(.df, .nDownsample = 0, .normalization = "scale",
                         .showProblem = FALSE) {
  if (.nDownsample != 0) {
    gene_idxs <- .df$gene_name |> sample(size = .nDownsample, replace = FALSE)
    .df <- .df |> filter(gene_name %in% gene_idxs)
  }
  if (.normalization == "scale") {
    plotdf <- .df |>
      reframe(mean_expr = mean(expr, na.rm = TRUE), 
              sd_expr = sd(expr, na.rm = TRUE),
              expr = expr,
              time_point_num = time_point_num,
              label = label,
              .by = "gene_name") |> 
      mutate(plot_expr = (expr - mean_expr)/sd_expr)
  }
  if (.normalization == "log2") {
    plotdf <- .df |> 
      mutate(plot_expr = log2(expr + 1))
  }
  p <- ggplot(plotdf, aes(x = time_point_num, y = plot_expr)) + 
    geom_line(aes(group = gene_name)) +
    # adding mean expr line for each cluster:
    geom_line(data = summarise(group_by(plotdf, time_point_num, label),
                                mean_expr = mean(plot_expr, na.rm = TRUE)),
              aes(x = time_point_num, y = mean_expr),
              color = "gold") +
    # adding mean expr for all low expressed genes (group_by won't create an NA group):
    geom_line(data = summarise(group_by(filter(plotdf, is.na(label)),
                                        time_point_num),
                               mean_expr = mean(plot_expr, na.rm = TRUE),
                               label = NA),
              aes(x = time_point_num, y = mean_expr),
              color = "gold") +
    facet_wrap(~ label)
  return(p)
}
```

## Choosing a low variance threshold

Blue line in below plots represents dispersion threshold. Genes below the line will be assigned to lowly varying (static) cluster. Yellow line is best fit for mean/var relationship

```{r var-thresh-exploration, cache=TRUE}
var_thresh <- 3

getCollapsedCountsByExperiment <- function(.experiment) {
  name_cer <- paste("cer", .experiment, sep = "_")
  name_par <- paste("par", .experiment, sep = "_")
  .cts_list <- list(name_cer = counts_list[[name_cer]],
                    name_par = counts_list[[name_par]])
  cts <- map2(.cts_list, names(.cts_list), .f = \(x, nm) {
    nm <- gsub("_.*", "", nm)
    rownames(x) <- paste(nm, rownames(x), sep = "_")
    return(x)
  }) 
  dfs <- map(cts, makeDf, .info = info, .join_by = "condition")
  outdf <- purrr::reduce(dfs, bind_rows)
  # after collapsing replicates, samples should be in the same condition order
  # and able to be rbound
  cts <- cts |> 
    purrr::reduce(rbind)
  return(cts)
}
# HAP4
toy_mat_collapsed <- getCollapsedCountsByExperiment("HAP4")
# note that it is important to specify whether we're thresholding based on log2(mean(expr))
# versus mean(log2(expr)):
plot(rowMeans(log2(toy_mat_collapsed)), 
     log2(rowMeans(toy_mat_collapsed)))
abline(a = 0, b = 1, col = "gold") 
# they're mostly a similar measure, 
# but for genes with high variance,
# log2(mean(expr)) tends to be higher 
# (hence those spikes of genes well above the y=x)
# we are thresholding by log2(mean(expr)) because it makes more intuitive sense
plot(log2(rowMeans(toy_mat_collapsed)), log2(rowVars(toy_mat_collapsed)), main = "Diauxic Shift")
mod <- lm(log2(rowVars(toy_mat_collapsed) + 1) ~ log2(rowMeans(toy_mat_collapsed) + 1))
abline(a = mod$coefficients[1], 
       b = mod$coefficients[2], col = "gold")
# the point where the low expr genes "lift off" this mean-var relationship line is where 
# lowly expressed genes have an abnormally high variance, variance that is more likely
# due to noise than response to the environment, and should therefore be in the low expr group
# low var filtering:
# var thresh is actually a dispersion thresh --- var/mean
# to filter out lowly varying genes, we apply disp threshold after the expr threshold
# the higher you're expressed, the more you need to vary in order to not be put in low var category
abline(a = 0, b = log2(var_thresh), col = "blue") 
# HU Shock
toy_mat_collapsed <- getCollapsedCountsByExperiment("CC")
plot(log2(rowMeans(toy_mat_collapsed)), log2(rowVars(toy_mat_collapsed)), main = "HU Shock")
mod <- lm(log2(rowVars(toy_mat_collapsed) + 1) ~ log2(rowMeans(toy_mat_collapsed) + 1))
abline(a = mod$coefficients[1], 
       b = mod$coefficients[2], col = "gold") 
abline(a = 0, b = log2(var_thresh), col = "blue")
# LowN
toy_mat_collapsed <- getCollapsedCountsByExperiment("LowN")
plot(log2(rowMeans(toy_mat_collapsed)), log2(rowVars(toy_mat_collapsed)), main = "LowN")
mod <- lm(log2(rowVars(toy_mat_collapsed) + 1) ~ log2(rowMeans(toy_mat_collapsed) + 1))
abline(a = mod$coefficients[1], 
       b = mod$coefficients[2], col = "gold") 
abline(a = 0, b = log2(var_thresh), col = "blue")
# LowPi
toy_mat_collapsed <- getCollapsedCountsByExperiment("LowPi")
plot(log2(rowMeans(toy_mat_collapsed)), log2(rowVars(toy_mat_collapsed)), main = "LowPi")
mod <- lm(log2(rowVars(toy_mat_collapsed) + 1) ~ log2(rowMeans(toy_mat_collapsed) + 1))
abline(a = mod$coefficients[1], 
       b = mod$coefficients[2], col = "gold") 
abline(a = 0, b = log2(var_thresh), col = "blue")
# Heat
toy_mat_collapsed <- getCollapsedCountsByExperiment("Heat")
plot(log2(rowMeans(toy_mat_collapsed)), log2(rowVars(toy_mat_collapsed)), main = "Heat")
mod <- lm(log2(rowVars(toy_mat_collapsed) + 1) ~ log2(rowMeans(toy_mat_collapsed) + 1))
abline(a = mod$coefficients[1], 
       b = mod$coefficients[2], col = "gold") 
abline(a = 0, b = log2(var_thresh), col = "blue")
# Cold
toy_mat_collapsed <- getCollapsedCountsByExperiment("Cold")
plot(log2(rowMeans(toy_mat_collapsed)), log2(rowVars(toy_mat_collapsed)), main = "Cold")
mod <- lm(log2(rowVars(toy_mat_collapsed) + 1) ~ log2(rowMeans(toy_mat_collapsed) + 1))
abline(a = mod$coefficients[1], 
       b = mod$coefficients[2], col = "gold") 
abline(a = 0, b = log2(var_thresh), col = "blue")
```

## Clustering

Running the heirarchical clustering using 2, 3, or 4 clusters per environment and a dispersion threshold of 1, 3, or 5 (using nClust=2 when modulating dispersion threshold)

```{r clustering, cache=TRUE}
# setting dispersion threshold
var_thresh <- 3

# clustering with nClust=2, disp=3
clusterdf_list <- vector(mode = "list", length = 0)
for (e in c("CC", "HAP4", "LowN", "LowPi", "Heat", "Cold")) {
  cat("*********** working on 2 clusters disp=", var_thresh, "in", e, "*********** \n")
  output <- clusterCountsList(counts_list[grepl(e, names(counts_list))], 
                          .nClust = 2, 
                          .tree_too = TRUE,
                          .min_var = var_thresh)
  clusterdf_list[[paste(e, 2, sep = "_")]] <- output
}

# converting cluster lists to a dataframe
getClusterCombination <- function(.clust_list) {
  cluster_comb <- tibble()
  for (nm in names(.clust_list)) {
    e <- gsub("_.*", "", nm)
    nclust <- gsub(".*_", "", nm)
    e_clust <- .clust_list[[paste(e, nclust, sep = "_")]]$df |> 
      select(label, gene_name) |> 
      unique()
    e_clust$gene_ID <- map(e_clust$gene_name, \(.g) {
      return(gsub(".*_", "", .g))
    }) |> unlist()
    e_clust$species <- map(e_clust$gene_name, \(.g) {
      return(gsub("_.*", "", .g))
    }) |> unlist()
    e_clust <- e_clust |> 
      select(gene_ID, species, label) |> 
      pivot_wider(id_cols = gene_ID, names_from = species, values_from = label) |> 
      mutate(experiment = e)
    cluster_comb <- bind_rows(cluster_comb, e_clust)
  }
  return(cluster_comb)
}
clusterdf <- getClusterCombination(clusterdf_list)

# clustering with nClust=2, disp=1
clusterdf_list_1disp <- vector(mode = "list", length = 0)
for (e in c("CC", "HAP4", "LowN", "LowPi", "Heat", "Cold")) {
  cat("*********** working on 2 clusters disp=1 in", e, "*********** \n")
  output <- clusterCountsList(counts_list[grepl(e, names(counts_list))], 
                          .nClust = 2, 
                          .tree_too = TRUE,
                          .min_var = 1) # changed dispersion threshold for this run
  clusterdf_list_1disp[[paste(e, 2, sep = "_")]] <- output
}
clusterdf_1disp <- getClusterCombination(clusterdf_list_1disp)

# clustering with nClust=2, disp=5
clusterdf_list_5disp <- vector(mode = "list", length = 0)
for (e in c("CC", "HAP4", "LowN", "LowPi", "Heat", "Cold")) {
  cat("*********** working on 2 clusters disp=5 in", e, "*********** \n")
  output <- clusterCountsList(counts_list[grepl(e, names(counts_list))], 
                          .nClust = 2, 
                          .tree_too = TRUE,
                          .min_var = 5) # changed dispersion threshold for this run
  clusterdf_list_5disp[[paste(e, 3, sep = "_")]] <- output
}
clusterdf_5disp <- getClusterCombination(clusterdf_list_5disp)

# clustering with nClust=3, disp=3
clusterdf_list3 <- vector(mode = "list", length = 0)
for (e in c("CC", "HAP4", "LowN", "LowPi", "Heat", "Cold")) {
  cat("*********** working on 3 clusters disp=", var_thresh, "in", e, "*********** \n")
  output <- clusterCountsList(counts_list[grepl(e, names(counts_list))],
                              .nClust = 3, .tree_too = TRUE,
                              .min_var = var_thresh)
  clusterdf_list3[[paste(e, 3, sep = "_")]] <- output
}
clusterdf3 <- getClusterCombination(clusterdf_list3)

# clustering with nClust=4, disp=3
clusterdf_list4 <- vector(mode = "list", length = 0)
for (e in c("CC", "HAP4", "LowN", "LowPi", "Heat", "Cold")) {
  cat("*********** working on 4 clusters disp=", var_thresh, "in", e, "*********** \n")
  output <- clusterCountsList(counts_list[grepl(e, names(counts_list))],
                              .nClust = 4, .tree_too = TRUE,
                              .min_var = var_thresh)
  clusterdf_list4[[paste(e, 4, sep = "_")]] <- output
}
clusterdf4 <- getClusterCombination(clusterdf_list4)
```

## Saving

```{r saving}
# primary clustering set we'll use:
save(clusterdf, clusterdf_list, file = "data_files/CorrelationClustering3Disp.RData")
# QC changing nClust or dispersion threshold
save(clusterdf, clusterdf_list,
     clusterdf_1disp, clusterdf_list_1disp,
     clusterdf_5disp, clusterdf_list_5disp,
     clusterdf3, clusterdf_list3,
     clusterdf4, clusterdf_list4, file = "data_files/QC_CorrelationClustering.RData")
```

