---
title: "Heiarchical Clustering"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Hierarchical clustering of expression data observation from combined-experiment clustering: each experiment only has a small number of distinct "shapes" to the expression a happy medium between experiment-specific clustering and combined-experiment clustering is to describe each gene by its combination of expression "shapes" in each experiment

```{r}
options(stringsAsFactors = FALSE)
source("functions_for_figure_scripts.R")
library(ggridges)

# for heiarchical clustering, we used counts 
# that have been "collapsed", averaged across replicates
# at each timepoint
load("data_files/Clustering_Counts.RData")
# for deciding low-varying genes we use the signal to noise ratio of
# time-dependent expression variability versus noise between replicates,
# so we need non-collapsed counts for this:
load("data_files/Cleaned_Count_Data.RData")
degree_val <- 3
snr_thresh <- 1
set.seed(23)

# Method:
# 1) Filter out genes with low signal-to-noise ratio
# 2) Correlation cluster remaining genes

# initial values for low var filter applied in corCluster (decided in following section)
var_thresh <- 3
```

## Clustering functions and unit tests

Bootstrapping function - runs clustering repeatedly on subsets of full clustering data and labels each gene by the cluster it is most frequently in (will be tested below corCluster)

```{r bootstrapClustering}
# helper function for corCluster
# runs nIter clusterings on random subsets of genes 
# returns the majority vote for each gene's cluster
bootstrapClustering <- function(.filtered_cts, .nClust,
                                .nIter = 100, .frac = 0.75) {
  clusterSubset <- function(.cts) {
    gene_idxs <- sample(c(1:nrow(.cts)), size = nrow(.cts)*.frac, replace = FALSE)
    frac_cts <- .cts[gene_idxs,]
    cor_mat <- frac_cts |> t() |> cor(use = "pairwise.complete.obs")
    tree <- hclust(as.dist(-cor_mat), method = "average")
    topBranchHeight <- sort(tree$height, decreasing = TRUE)[.nClust]
    tree_labels <- cutreeStatic(tree, cutHeight = topBranchHeight,
                                minSize = 1)
    output <- rep(NA, nrow(.cts))
    output[gene_idxs] <- tree_labels
    return(output)
  }
  label_list <- vector(mode = "list", length = .nIter)
  for (i in c(1:.nIter)) {
    label_list[[i]] <- clusterSubset(.filtered_cts)
  }
  labeldf <- purrr::reduce(label_list, .f = cbind)
  cat("finished iterations, nGenes x nIter:", dim(labeldf), "\n")
  rownames(labeldf) <- rownames(.filtered_cts)
  # aligning labels between iterations
  matched <- apply(labeldf, 2, matchLabels, reference = labeldf[,1],
                   ignoreLabels = NA)
  matched[matched > .nClust] <- NA # matchLabels adds new labels to any gene over pThreshold
  labels <- apply(matched, 1, \(x) {
    label <- names(sort(table(x, useNA = "no"), decreasing = TRUE))[1]
    if (length(label) == 0) {
      return(NA)
    }
    return(as.numeric(label))
  })
  return(labels)
}
```

```{r getData}
### Helper for signal-to-noise modeling, get the counts and time data for the model from
# the gene name, environment, and organism (cer or par)
getData <- function(.gene, .environment, .organism, .cts, .info) {
  .cts_vec <- .cts[.gene, .info$experiment == .environment & .info$organism == .organism]
  ctsdf <- tibble(counts = .cts_vec,
                  sample_name = names(.cts_vec)) |> 
    left_join(y = select(.info, sample_name, time_point_num), by = "sample_name")
  return(list(counts = as.numeric(ctsdf$counts), 
              timepoint = as.numeric(ctsdf$time_point_num)))
}
```

```{r fitModel}
### Fit our polynomial linear model
fitModel <- function(.counts_vec, .time_vec, .degree = 3) {
  x <- data.frame(count = .counts_vec,
                  timepoint = .time_vec)
  mod <- lm(count ~ poly(timepoint, degree = .degree), data = x)
  return(mod)
}
```

```{r getSNR}
getSNR <- function(.mod) {
  mod_anova <- anova(.mod)
  snr <- mod_anova[1,"Sum Sq"]/mod_anova["Residuals", "Sum Sq"]
  return(snr)
}
```

```{r test-getSNR}
# tests for fitModel and getData and getSNR
test_data <- getData(.gene = "YGR192C", .environment = "CC", .organism = "par", 
                     .cts = counts, .info = sample_info)
test_mod <- fitModel(.counts_vec = test_data$counts, 
                     .time_vec = test_data$timepoint)
getSNR(test_mod)
```
```{r getting-SNRs, cache=TRUE}
# Given an environment name and a list of gene idxs, 
# collect a vector with all the SNRs in that environment
griddf <- expand_grid(environment = unique(sample_info$experiment),
                      organism = c("cer", "par"))
snrdf <- map2(griddf$environment, griddf$organism, \(e, org) {
  cat("************", e, org, "************\n")
  modeldf <- map(rownames(counts),
                \(g) {
                  #cat(which(g == rownames(counts)), "/", nrow(counts), "\n")
                  e_degree <- if_else(e == "LowN",
                                      true = 2, false = degree_val)
                  g_data <- getData(.gene = g, .cts = counts,
                                    .info = sample_info, .environment = e, .organism = org)
                  mod <- fitModel(.counts_vec = g_data$counts,
                                  .time_vec = g_data$timepoint,
                                  .degree = e_degree)
                  snr <- getSNR(mod)
                  outdf <- tibble("gene_name" = g, "environment" = e,
                                  "organism" = org,
                                  "signal_to_noise" = snr)
                  return(outdf)
                }) |> purrr::reduce(.f = bind_rows)
  return(modeldf)
}) |> purrr::reduce(.f = bind_rows)
```

Clustering function
given a count matrix where rows are genes and columns are samples/conditions, returns heirarchical cluster assignments as vector of length nGenes labels follow convention:
  1) NA = too lowly expressed (mean expression < .min_expr)
  2) 0 = too low variance (mean expression >= .min_expr & var expression < .min_var)
  3) 1 - .nClust = clusters of co-varying genes

```{r corCluster}
corCluster <- function(.cts, .nClust, .is_low_var_vec,
                       .tree_too = FALSE, .gap_stat = FALSE, 
                       .just_counts = FALSE, .bootstrapIter = 100,
                       .bootstrapFrac = 0.75) {
  gene_idxs <- rownames(.cts)
  low_var_genes <- gene_idxs[.is_low_var_vec]
  cat(length(low_var_genes), "low variance genes, assigned to 0 label\n")
  good_clustering_genes <- setdiff(x = gene_idxs, y = low_var_genes)
  filtered_cts <- .cts[good_clustering_genes,]
  if (.gap_stat) {
    cluster_fun <- function(x, k) {
      list(cluster = cutree(hclust(d = as.dist(-cor(t(x), use = "pairwise.complete.obs")), 
                                   method = "average"), 
                            k = k))}
    gap_stat <- clusGap(filtered_cts, FUNcluster = cluster_fun, K.max = 10, B = 5, method = "firstmax")
    return(gap_stat)
  }
  if (.just_counts) {
    return(filtered_cts)
  }
  # clustering
  if (sum(apply(filtered_cts, 1, \(x) {all(is.na(x))})) != 0) { # if any rows (genes) have all NA values, they will cause cor to fail below
    cat("NA genes in counts matrix, returning counts matrix only\n")
    return(cor_mat)
  }
  if (.bootstrapIter > 1) {
    tree_labels <- bootstrapClustering(filtered_cts, .nClust = .nClust,
                        .nIter = .bootstrapIter, .frac = .bootstrapFrac)
    labeldf <- tibble(gene_name = rownames(filtered_cts),
                      label = tree_labels) |> 
      bind_rows(tibble(gene_name = low_var_genes,
                       label = 0))
    # making single tree for .treeToo = TRUE
    cor_mat <- filtered_cts |> t() |> cor(use = "pairwise.complete.obs")
    tree <- hclust(as.dist(-cor_mat), method = "average")
  }
  if (.bootstrapIter <= 1) {
    cor_mat <- filtered_cts |> t() |> cor(use = "pairwise.complete.obs")
    tree <- hclust(as.dist(-cor_mat), method = "average") # negative because hclust expects distance mat --- the higher the ij value the LESS similar genes i and j are
    topBranchHeight <- sort(tree$height, decreasing = TRUE)[.nClust]
    tree_labels <- cutreeStatic(tree, cutHeight = topBranchHeight,
                                minSize = 1) # guaranteeing no 0 class
    cat("cutting tree at height", topBranchHeight, "into", length(unique(tree_labels)), "clusters\n")
    labeldf <- tibble(gene_name = tree$labels,
                      label = tree_labels) |> 
      bind_rows(tibble(gene_name = low_var_genes,
                       label = 0))
  }
  # matching each label to each gene
  # rearranging genes into their original order
  outdf <- left_join(tibble(gene_name = rownames(.cts)),
                     labeldf, by = "gene_name")
  if (!.tree_too) {
    return(outdf)
  }
  if (.tree_too) {
    return(list(tree = tree, df = outdf))
  }
}
```

Unit tests for boostrapClustering and corCluster
```{r tests-corCluster}
### tests for corCluster/bootstrapClustering
# toy genes
# known co-expressed genes in LowN
# should split these into genes highest at TP2 (^ genes)
# and genes lowest at TP2 (v genes)
toy_idxs <- c("YBR172C", "YML015C", # v genes
              "YBR162W-A", "YKL196C", "YBR171W") # ^ genes (the excess of YBRs are coincidental, as I was just scrolling through that part of the module --- although the 171 172 W/C gene pair is probably overlapping)
toy_mat <- counts_list$par_LowN[toy_idxs,]
toydf <- makeDf(toy_mat, info, .join_by = "condition")
toyoutput <- corCluster(toy_mat, .nClust = 2, 
                        .is_low_var_vec = rep(FALSE, nrow(toy_mat)), 
                        .tree_too = TRUE, 
                        .bootstrapIter = 0)
toydf <- left_join(toydf, toyoutput$df, by = "gene_name")
ggplot(toydf, aes(x = time_point_str, y = log2(expr + 1))) + 
  geom_line(aes(group = gene_name,
                color = gene_name)) +
  facet_wrap(~label)

# full dataset
toy_mat <- counts_list$cer_LowN
lowVarGenes <- snrdf |> filter(environment == "LowN" & organism == "cer" &
                                 (is.na(signal_to_noise) | signal_to_noise < snr_thresh)) |> 
  select(gene_name) |> pull()
isLowVar <- rownames(toy_mat) %in% lowVarGenes
toydf <- makeDf(toy_mat, info, .join_by = "condition")
# no bootstrap
test_labels <- corCluster(toy_mat, .nClust = 4, 
                          .is_low_var_vec = isLowVar,
                          .bootstrapIter = 0)
# yes bootstrap
test_labels <- corCluster(toy_mat, .nClust = 4, 
                          .is_low_var_vec = isLowVar,
                          .bootstrapIter = 10)
plotdf <- toydf |> 
  group_by(gene_name, time_point_str) |> 
  summarise(expr = mean(expr)) |> 
  ungroup() |> 
  reframe(mean_expr = mean(expr), 
          sd_expr = sd(expr),
          expr = expr,
          time_point_str = time_point_str,
          .by = "gene_name") |> 
  mutate(scaled_expr = (expr - mean_expr)/sd_expr) |> 
  left_join(y = test_labels, by = "gene_name")
ggplot(plotdf, aes(x = time_point_str, y = scaled_expr)) + 
  geom_line(aes(group = gene_name)) +
  facet_wrap(~label)
```

Wrapper function for corCluster, repeatedly clusters list of environmental datasets

```{r clusterCountsList}
# Wrapper function that converts named list of counts (from 1 experiment) into
# tidy counts data frame with nGenes * nConditions * length(counts list) number of rows
clusterCountsList <- function(.cts_list, .nClust = 4,
                              .environment,
                              .bootstrapIter = 10,
                              .bootstrapFrac = 0.75,
                              .is_low_var_df, 
                              .snr_thresh = NULL,
                              .snr_frac = NULL,
                              .tree_too = FALSE,
                              .gap_stat = FALSE,
                              .just_counts = FALSE) {
  # giving each ortholog a unique name: YGR192C becomes YGR192C_cer
  cts <- map2(.cts_list, names(.cts_list), .f = \(x, nm) {
    nm <- gsub("_.*", "", nm)
    rownames(x) <- paste(rownames(x), nm, sep = "_")
    return(x)
  })
  # adjusting this environment's snrdf to name orthologs uniquely: YGR192C becomes YGR192C_cer
  low_var_df_e <- .is_low_var_df |> 
    filter(environment == .environment) |> 
    mutate(gene_name = paste(gene_name, organism, sep = "_"))
  # filtering lowly-varying genes
  if (!is.null(.snr_frac) & !is.null(.snr_thresh)) {
    stop("cannot specify both a threshold and a fraction for SNR filtering\n")
  }
  # if a hard cutoff was given, filtering low varying gens based on it
  if (!is.null(.snr_thresh)) {
    # filtering low varying genes based on given threshold
    low_var_genes <- filter(low_var_df_e, is.na(signal_to_noise) | signal_to_noise < .snr_thresh) |> 
      select(gene_name) |> pull()
  }
  if (!is.null(.snr_frac)) {
    # filtering low varying genes based on quantile
    snr_thresh_cer <- filter(low_var_df_e, organism == "cer") |> 
      drop_na() |> select(signal_to_noise) |> pull() |> quantile(probs = .snr_frac)
    snr_thresh_par <- filter(low_var_df_e, organism == "par") |> 
      drop_na() |> select(signal_to_noise) |> pull() |> quantile(probs = .snr_frac)
    low_var_genes <- filter(low_var_df_e, is.na(signal_to_noise) |
                            (organism == "cer" & signal_to_noise < snr_thresh_cer) | 
                              (organism == "par" & signal_to_noise < snr_thresh_par)) |>
      select(gene_name) |> pull()
  }
  
  dfs <- map(cts, makeDf, .info = info, .join_by = "condition")
  outdf <- purrr::reduce(dfs, bind_rows)
  if (all(unlist(map(cts, \(x) {all(colnames(x) == colnames(cts[[1]]))})))) {
    cat("counts are in same order, joining counts into single matrix\n")
    cts <- purrr::reduce(cts, .f = rbind)
  }
  else {
    cat("counts colnames don't match, returning counts list \n")
    return(cts)
  }
  # after collapsing replicates, samples should be in the same condition order
  # and able to be rbound
  
  # converting low var genes list to a TRUE/FALSE vector the same lenght as cts
  is_low_var_vec <- rownames(cts) %in% low_var_genes

  if (.just_counts) {
    filtered_cts <- corCluster(cts, .nClust = .nClust, 
                               .is_low_var_vec = is_low_var_vec,
                               .just_counts = TRUE, .bootstrapIter = .bootstrapIter,
                               .bootstrapFrac = .bootstrapFrac)
    return(filtered_cts)
  }
  if (!.tree_too) {
    labeldf <- corCluster(cts, .nClust = .nClust, 
                          .is_low_var_vec = is_low_var_vec,
                          .bootstrapIter = .bootstrapIter,
                          .bootstrapFrac = .bootstrapFrac)
    outdf <- left_join(outdf, labeldf, by = "gene_name")
    return(outdf)
  }
  if (.tree_too) {
    output <- corCluster(cts, .nClust = .nClust, 
                         .is_low_var_vec = is_low_var_vec,
                         .bootstrapIter = .bootstrapIter,
                         .bootstrapFrac = .bootstrapFrac,
                         .tree_too = TRUE)
    output$df <- left_join(outdf, output$df, by = "gene_name")
    return(output)
  }
}
# tests for clusterCountsList
# par is missing 1 condition versus cer
testout <- clusterCountsList(list("cer_LowN" = counts_list$cer_LowN,
                                  "par_LowN" = counts_list$par_LowN),
                             .environment = "LowN",
                             .bootstrapIter = 10, 
                             .is_low_var_df = snrdf,
                             .snr_thresh = 1)
testout$gene_name[1:10]
```
Visualize clustering

```{r plotClusters}
# given a dataframe with labels for each gene, returns a facet plot
# showing expression of each gene in each cluster (usually randomly 
# downsample option to save plotting computation)
plotClusters <- function(.df, .nDownsample = 0, .normalization = "scale",
                         .showProblem = FALSE) {
  if (.nDownsample != 0) {
    gene_idxs <- .df$gene_name |> sample(size = .nDownsample, replace = FALSE)
    .df <- .df |> filter(gene_name %in% gene_idxs)
  }
  if (.normalization == "scale") {
    plotdf <- .df |>
      reframe(mean_expr = mean(expr, na.rm = TRUE), 
              sd_expr = sd(expr, na.rm = TRUE),
              expr = expr,
              time_point_num = time_point_num,
              label = label,
              .by = "gene_name") |> 
      mutate(plot_expr = (expr - mean_expr)/sd_expr)
  }
  if (.normalization == "log2") {
    plotdf <- .df |> 
      mutate(plot_expr = log2(expr + 1))
  }
  p <- ggplot(plotdf, aes(x = time_point_num, y = plot_expr)) + 
    geom_line(aes(group = gene_name)) +
    # adding mean expr line for each cluster:
    geom_line(data = summarise(group_by(plotdf, time_point_num, label),
                                mean_expr = mean(plot_expr, na.rm = TRUE)),
              aes(x = time_point_num, y = mean_expr),
              color = "gold") +
    # adding mean expr for all low expressed genes (group_by won't create an NA group):
    geom_line(data = summarise(group_by(filter(plotdf, is.na(label)),
                                        time_point_num),
                               mean_expr = mean(plot_expr, na.rm = TRUE),
                               label = NA),
              aes(x = time_point_num, y = mean_expr),
              color = "gold") +
    facet_wrap(~ label)
  return(p)
}
plotClusters(testout)
```

## Choosing a low variance threshold

First checking all the instances where signal-to-noise wasn't able to be calculated (returned NA):

```{r snr-NA-QC}
mean_expr_df <- map2(counts_list, names(counts_list), \(mat, nm) {
  outdf <- tibble(gene_name = rownames(mat),
                  mean_expr = rowMeans(mat),
                  mat_name = nm)
  return(outdf)
}) |> purrr::reduce(.f = bind_rows) |> 
  mutate(organism = unlist(tstrsplit(mat_name, split = "_", keep = 1)),
         environment = unlist(tstrsplit(mat_name, split = "_", keep = 2))) |> 
  dplyr::select(-mat_name)
plotdf <- snrdf |> mutate(hasNA_SNR = is.na(signal_to_noise)) |> 
  left_join(mean_expr_df, by = c("gene_name", "organism", "environment"))
# If the gene had an NA for its SNR value, it either has has a mean expr of 0:
filter(plotdf, hasNA_SNR) |> select(mean_expr) |> pull() |> max(na.rm = TRUE) # should output 0
# Or it was below the low-expr threshold in both species (mean expr NA b/c it was missing from counts list in that environment):
filter(plotdf, is.na(mean_expr)) |> select(hasNA_SNR) |> table() # should have some "TRUE" values
```

Conclusion: all the instances of SNR = NA were either genes with mean expression of 0 in one species or genes that were excluded for low expression because they were below our cutoff in both species.

```{r snr-distributions}
# visualzing if SNR distributions vary substantially between environments
# or if we can use the same cutoff for all of them (SNR = 1 probably)
plotdf <- snrdf |> filter(signal_to_noise > 0)
p <- ggplot(data = plotdf, aes(x = log2(signal_to_noise), 
                               y = paste0(environment, organism, sep = "_"), 
                                          fill = environment)) +
  geom_density_ridges(data = plotdf) +
  geom_vline(xintercept = 1) +
  theme_classic() +
  theme(legend.position = "none") +
  ylab("") +
  xlab("log2 SNR")
p
```

Mostly similar distributions between species and experiments, the exception is Heat and Cold, where both species have higher SNR but Spar has higher SNR than Scer. Interestingly, 71% of genes that were diverging in level in Cold and 65% in Heat had higher expression in Scer, i.e. it was not symmetrical. These could be real differences in biology, but they could also be differences in sequencing techniques

Conclusion: A common SNR threshold that removes about 5-10% of genes for being lowly varying in the non-temperature environments would likely work well in most cases. Fewer genes would be eliminated from temperature, but we saw that previously.

...of course an SNR of 1 would preserve some genes in all environments and it is intuitive. Ideally we try both a hard threshold and a percentile-based threshold

## Clustering

Running the heirarchical clustering using 2, 3, or 4 clusters per environment and a dispersion threshold of 1, 3, or 5 (using nClust=2 when modulating dispersion threshold)

```{r clustering, cache=TRUE}

### SNR cutoff is hard threshold
# clustering with nClust=2
clusterdf_list <- vector(mode = "list", length = 0)
for (e in c("CC", "HAP4", "LowN", "LowPi", "Heat", "Cold")) {
  cat("*********** working on 2 clusters, SNR > 1", "in", e, "*********** \n")
  output <- clusterCountsList(counts_list[grepl(e, names(counts_list))], 
                          .nClust = 2, 
                          .environment = e,
                          .snr_thresh = 1,
                          .is_low_var_df = snrdf,
                          .tree_too = TRUE)
  clusterdf_list[[paste(e, 2, sep = "_")]] <- output
}

# converting cluster lists to a dataframe
getClusterCombination <- function(.clust_list) {
  cluster_comb <- tibble()
  for (nm in names(.clust_list)) {
    e <- gsub("_.*", "", nm)
    nclust <- gsub(".*_", "", nm)
    e_clust <- .clust_list[[paste(e, nclust, sep = "_")]]$df |> 
      select(label, gene_name) |> 
      unique()
    e_clust$gene_ID <- map(e_clust$gene_name, \(.g) {
      return(gsub("_.*", "", .g))
    }) |> unlist()
    e_clust$species <- map(e_clust$gene_name, \(.g) {
      return(gsub(".*_", "", .g))
    }) |> unlist()
    e_clust <- e_clust |> 
      select(gene_ID, species, label) |> 
      pivot_wider(id_cols = gene_ID, names_from = species, values_from = label) |> 
      mutate(experiment = e)
    cluster_comb <- bind_rows(cluster_comb, e_clust)
  }
  return(cluster_comb)
}
clusterdf <- getClusterCombination(clusterdf_list)

# clustering with nClust=2, snr = 0.5
clusterdf_list_snr0.5 <- vector(mode = "list", length = 0)
for (e in c("CC", "HAP4", "LowN", "LowPi", "Heat", "Cold")) {
  cat("*********** working on 2 clusters, SNR > 0.5 in", e, "*********** \n")
  output <- clusterCountsList(counts_list[grepl(e, names(counts_list))], 
                          .nClust = 2, 
                          .environment = e,
                          .is_low_var_df = snrdf,
                          .snr_thresh = 0.5,
                          .tree_too = TRUE) # changed dispersion threshold for this run
  clusterdf_list_snr0.5[[paste(e, 2, sep = "_")]] <- output
}
clusterdf_snr0.5 <- getClusterCombination(clusterdf_list_snr0.5)

# clustering with nClust=2, SNR = 0.1
clusterdf_list_snr0.1 <- vector(mode = "list", length = 0)
for (e in c("CC", "HAP4", "LowN", "LowPi", "Heat", "Cold")) {
  cat("*********** working on 2 clusters, SNR > 0.1 in", e, "*********** \n")
  output <- clusterCountsList(counts_list[grepl(e, names(counts_list))], 
                          .nClust = 2, 
                          .environment = e,
                          .is_low_var_df = snrdf,
                          .snr_thresh = 0.1,
                          .tree_too = TRUE)
  clusterdf_list_snr0.1[[paste(e, 2, sep = "_")]] <- output
}
clusterdf_snr0.1 <- getClusterCombination(clusterdf_list_snr0.1)

### SNR cutoff as percentile instead of solid threshold
# clustering with nClust = 2

# 30% lowest SNR per species
clusterdf_list_snr30pct <- vector(mode = "list", length = 0)
for (e in c("CC", "HAP4", "LowN", "LowPi", "Heat", "Cold")) {
  cat("*********** working on 2 clusters, SNR @ 30%", "in", e, "*********** \n")
  output <- clusterCountsList(counts_list[grepl(e, names(counts_list))], 
                          .nClust = 2, 
                          .environment = e,
                          .snr_frac = 0.3,
                          .is_low_var_df = snrdf,
                          .tree_too = TRUE)
  clusterdf_list_snr30pct[[paste(e, 2, sep = "_")]] <- output
}
clusterdf_snr30pct <- getClusterCombination(clusterdf_list_snr30pct)

# 10% lowest SNR per species
clusterdf_list_snr10pct <- vector(mode = "list", length = 0)
for (e in c("CC", "HAP4", "LowN", "LowPi", "Heat", "Cold")) {
  cat("*********** working on 2 clusters, SNR @ 10%", "in", e, "*********** \n")
  output <- clusterCountsList(counts_list[grepl(e, names(counts_list))], 
                          .nClust = 2, 
                          .environment = e,
                          .snr_frac = 0.1,
                          .is_low_var_df = snrdf,
                          .tree_too = TRUE)
  clusterdf_list_snr10pct[[paste(e, 2, sep = "_")]] <- output
}
clusterdf_snr10pct <- getClusterCombination(clusterdf_list_snr10pct)

# 5% lowest SNR per species
clusterdf_list_snr5pct <- vector(mode = "list", length = 0)
for (e in c("CC", "HAP4", "LowN", "LowPi", "Heat", "Cold")) {
  cat("*********** working on 2 clusters, SNR @ 5%", "in", e, "*********** \n")
  output <- clusterCountsList(counts_list[grepl(e, names(counts_list))], 
                          .nClust = 2, 
                          .environment = e,
                          .snr_frac = 0.05,
                          .is_low_var_df = snrdf,
                          .tree_too = TRUE)
  clusterdf_list_snr5pct[[paste(e, 2, sep = "_")]] <- output
}
clusterdf_snr5pct <- getClusterCombination(clusterdf_list_snr5pct)

### changing nClust
# clustering with nClust=3, SNR = 0.05%
clusterdf_list3 <- vector(mode = "list", length = 0)
for (e in c("CC", "HAP4", "LowN", "LowPi", "Heat", "Cold")) {
  cat("*********** working on 3 clusters SNR < 0.5 in", e, "*********** \n")
  output <- clusterCountsList(counts_list[grepl(e, names(counts_list))],
                              .nClust = 3, 
                              .environment = e,
                              .tree_too = TRUE,
                              .snr_frac = 0.05,
                              .is_low_var_df = snrdf)
  clusterdf_list3[[paste(e, 3, sep = "_")]] <- output
}
clusterdf3 <- getClusterCombination(clusterdf_list3)

# clustering with nClust=4, SNR = 0.05%
clusterdf_list4 <- vector(mode = "list", length = 0)
for (e in c("CC", "HAP4", "LowN", "LowPi", "Heat", "Cold")) {
  cat("*********** working on 4 clusters SNR < 0.5 in", e, "*********** \n")
  output <- clusterCountsList(counts_list[grepl(e, names(counts_list))],
                              .nClust = 4, 
                              .environment = e,
                              .tree_too = TRUE,
                              .snr_frac = 0.05,
                              .is_low_var_df = snrdf)
  clusterdf_list4[[paste(e, 4, sep = "_")]] <- output
}
clusterdf4 <- getClusterCombination(clusterdf_list4)
```

## Saving

```{r saving}
# primary clustering set we'll use:
save(clusterdf, clusterdf_list, file = "data_files/CorrelationClustering_SNR1.RData")
# QC changing nClust or dispersion threshold
save(clusterdf, clusterdf_list,
     clusterdf_snr0.5, clusterdf_list_snr0.5,
     clusterdf_snr0.1, clusterdf_list_snr0.1,
     clusterdf_snr10pct, clusterdf_list_snr10pct,
     clusterdf_snr5pct, clusterdf_list_snr5pct,
     clusterdf_snr30pct, clusterdf_list_snr30pct,
     clusterdf3, clusterdf_list3,
     clusterdf4, clusterdf_list4, file = "data_files/QC_CorrelationClustering_SignalToNoise.RData")
```

