---
title: "Level and plasticity divergence in a common trans-regulatory environment"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r package-loading}
source("functions_for_figure_scripts.R")
library(ggExtra)
library(ggVennDiagram)
library(ggalluvial)
load("data_files/FinalDataframe3Disp.RData")
load("data_files/Cleaned_Counts.RData")
load("data_files/Cleaned_Counts_Allele.RData")

cor_thresh <- 0.8
eff_thresh <- log2(1.5)
# extreme cor and log2 fold change values to extract the most
# cis plasticity divergers and most trans level divergers for
# avg expression figures
cis_plast_thresh <- -0.25 
trans_lev_thresh <- log2(2)
```

## Discrete heatmap of hybrid vs parent plasticity

```{r discrete-heatmap}
# HAP4
plot_mat <- finaldf |> filter(experiment == "HAP4") |> 
  dplyr::select(cer, par, hyc, hyp) |> 
  drop_na() |> 
  mutate(parent_code = paste0(cer, par),
         hybrid_code = paste0(hyc, hyp)) |> 
  dplyr::select(parent_code, hybrid_code) |> 
    as.matrix() |> t()
ordered_plot_mat <- orderGenesByGroup(plot_mat, .labels = c("00", "11", "22", "01", "02", "10", "12", "20", "21"))
plotDiscreteHeatmap(ordered_plot_mat, .row_names = c("parents", "hybrid"))
```

### Alluvial

```{r alluvial}
plotdf_counts <- finaldf |> 
  #filter(cer != 0 & par != 0 & hyc != 0 & hyp != 0) |> 
  dplyr::select(experiment, 
                cer, par, hyc, hyp) |> 
  drop_na() |> 
  mutate(parents = if_else(cer == par,
                           true = "parents_same",
                           false = "parents_different"),
         hybrids = if_else(hyc == hyp,
                           true = "hybrids_same",
                           false = "hybrids_different"),
         `hybrid allele matches parental` = cer == hyc & par == hyp) |>
  group_by(experiment, parents, hybrids, `hybrid allele matches parental`) |> 
  summarise(ngenes = n()) |> 
  ungroup()
# frequency of matching among genes with plasticity 
# divergence in parents and hybrid (20% would be random chance)
plotdf_counts |> filter(parents == "parents_different" &
                            hybrids == "hybrids_different") |> select(experiment, `hybrid allele matches parental`, ngenes) |> pivot_wider(id_cols = "experiment", names_from = "hybrid allele matches parental", values_from = "ngenes") |> mutate(prop_match = `TRUE`/(`TRUE`+`FALSE`))
# plotting alluvial
p <- ggplot(plotdf_counts, aes(y = ngenes, axis1 = parents, axis2 = hybrids)) +
  geom_alluvium(aes(fill = `hybrid allele matches parental`), color = "black", width = 1/12) +
  #scale_fill_discrete(type = c(TRUE = "red", FALSE = "black")) +
  geom_stratum(width = 1/12, fill = "white", color = "black") +
  geom_label(stat = "stratum", aes(label = after_stat(stratum)),
             angle = 90, size = 1) +
  theme_classic() +
  facet_wrap(~factor(experiment, levels = ExperimentNames, 
                     labels = LongExperimentNames)) +
  xlab("") +
  ylab("number of genes") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.position = "top")
p
```

Saving for figure:

```{r alluvial-pdf}
ggsave("paper_figures/CisTrans/alluvial.pdf", width = 5, height = 5)
```

Plotting average expression for random cis group and random trans group in each environment, provided parents are 1 or 2

```{r cis-trans-avg-expr}
plotlist_cis <- purrr::map(ExperimentNames, \(e) {
  scer_clust <- sample(c(1, 2), size = 1)
  spar_clust <- setdiff(c(1, 2), scer_clust)
  gene_idxs <- finaldf |> 
    dplyr::filter(cer == scer_clust & 
                    par == spar_clust &
                    hyc == scer_clust &
                    hyp == spar_clust &
                    experiment == e) |> 
    select(gene_name) |> pull()
  p <- plotGenes(gene_idxs, .normalization = "scale", .quartet = TRUE, .experiment_name = e, .plot_titles = paste0(e, " ", length(gene_idxs), " genes"))
  return(p)
})
ggarrange(plotlist = plotlist_cis,
          nrow = 1, ncol = 6)
#ggsave("paper_figures/CisTrans/examples_cis.pdf",
       #width = 12, height = 2)
plotlist_trans <- purrr::map(ExperimentNames, \(e) {
  scer_clust <- sample(c(1, 2), size = 1)
  spar_clust <- setdiff(c(1, 2), scer_clust)
  gene_idxs <- finaldf |> 
    dplyr::filter(cer == scer_clust & 
                    par == spar_clust &
                    hyc == hyp &
                    experiment == e) |> 
    select(gene_name) |> pull()
  p <- plotGenes(gene_idxs, .normalization = "scale", .quartet = TRUE, .experiment_name = e, .plot_titles = paste0(e, " ", length(gene_idxs), " genes"))
  return(p)
})
ggarrange(plotlist = plotlist_trans,
          nrow = 1, ncol = 6)
#ggsave("paper_figures/CisTrans/examples_trans.pdf",
       #width = 12, height = 2)
```


Do those genes with 


## Cross-parent tests to compare the cis basis of level divergence vs plasticity divergence

Is having the "correct" parental allele more informative than having the "incorrect" parental allele? It should be, at least if there is cis-regulatory divergence

Here we compare the frequency with which a hybrid allele has the same direction of level difference (+ or - log2 fold change) or plasticity (same cluster 0, 1, or 2) as a parental allele (either the "correct" parent of the "cross" parent) versus randomly assigned hybrid-parent pairs of different genes.

8 permutation distributions total: correct vs incorrect parent x level vs plasticity x diverged vs conserved = 2^3 = 8

```{r getPlasticityCrossSpeciesDataset}
getPlasticityCrossSpeciesDataset <- function(.df, .separate_species = FALSE, .nIter = 10000) {
  nGenes <- nrow(.df)
  cer_clusters <- .df$cer
  par_clusters <- .df$par
  hyc_clusters <- .df$hyc
  hyp_clusters <- .df$hyp
  if (!.separate_species) {
    observed_same <- sum(c(cer_clusters, par_clusters) == c(hyc_clusters, hyp_clusters))/(nGenes*2)
    observed_cross <- sum(c(cer_clusters, par_clusters) == c(hyp_clusters, hyc_clusters))/(nGenes*2)
    null_same <- purrr::map(c(1:.nIter), \(i) {
      scrambled_hybrid_clusters <- sample(c(hyc_clusters, hyp_clusters), size = nGenes*2, replace = FALSE)
      return(sum(c(cer_clusters, par_clusters) == scrambled_hybrid_clusters)/(nGenes*2))
    }) |> unlist()
    null_cross <- purrr::map(c(1:.nIter), \(i) {
      scrambled_hybrid_clusters <- sample(c(hyp_clusters, hyc_clusters), size = nGenes*2, replace = FALSE)
      return(sum(c(cer_clusters, par_clusters) == scrambled_hybrid_clusters)/(nGenes*2))
    }) |> unlist()
    return(list(observed_samespecies = observed_same,
                observed_crossspecies = observed_cross,
                null_samespecies = null_same,
                null_crossspecies = null_cross))
  }
  if (.separate_species) {
    observed_cerhyc <- sum(cer_clusters == hyc_clusters)/nGenes
    observed_cerhyp <- sum(cer_clusters == hyp_clusters)/nGenes
    observed_parhyc <- sum(par_clusters == hyc_clusters)/nGenes
    observed_parhyp <- sum(par_clusters == hyp_clusters)/nGenes
    null_cerhyc <- purrr::map(c(1:.nIter), \(i) {
      scrambled_hyc_clusters <- sample(hyc_clusters, size = nGenes, replace = FALSE)
      return(sum(cer_clusters == scrambled_hyc_clusters)/nGenes)
    }) |> unlist()
    null_cerhyp <- purrr::map(c(1:.nIter), \(i) {
      scrambled_hyp_clusters <- sample(hyp_clusters, size = nGenes, replace = FALSE)
      return(sum(cer_clusters == scrambled_hyp_clusters)/nGenes)
    }) |> unlist()
    null_parhyc <- purrr::map(c(1:.nIter), \(i) {
      scrambled_hyc_clusters <- sample(hyc_clusters, size = nGenes, replace = FALSE)
      return(sum(par_clusters == scrambled_hyc_clusters)/nGenes)
    }) |> unlist()
    null_parhyp <- purrr::map(c(1:.nIter), \(i) {
      scrambled_hyp_clusters <- sample(hyp_clusters, size = nGenes, replace = FALSE)
      return(sum(par_clusters == scrambled_hyp_clusters)/nGenes)
    }) |> unlist()
    return(list(observed_samespecies_Scer = observed_cerhyc,
                observed_samespecies_Spar = observed_parhyp,
                observed_crossspecies_Scer = observed_cerhyp,
                observed_crossspecies_Spar = observed_parhyc,
                null_samespecies_Scer = null_cerhyc,
                null_samespecies_Spar = null_parhyp,
                null_crossspecies_Scer = null_cerhyp,
                null_crossspecies_Spar = null_parhyc))
  }
}
# tests for getPlasticityCrossSpeciesDataset
# hard to test this one, so we'll just use the separate species option to verify that results are not different for Scer versus Spar
test <- getPlasticityCrossSpeciesDataset(.df = filter(finaldf, plasticity == "diverged"), # change btween conserved/diverged plasticity or filter for different environments
                                         .nIter = 100, .separate_species = TRUE)
# Scer, same species
p_scer_same <- plotPermutationTestOn0to1Scale(.observed_value = test$observed_samespecies_Scer,
                                   .null_vector = test$null_samespecies_Scer,
                                   .xlab = "fraction of Hyc genes in same cluster as Scer")
# Scer, cross species
p_scer_cross <- plotPermutationTestOn0to1Scale(.observed_value = test$observed_crossspecies_Scer,
                                    .null_vector = test$null_crossspecies_Scer,
                                    .xlab = "fraction of Hyp genes in same cluster as Scer")
# Spar, same species
p_spar_same <- plotPermutationTestOn0to1Scale(.observed_value = test$observed_samespecies_Spar,
                                   .null_vector = test$null_samespecies_Spar,
                                   .xlab = "fraction of Hyp genes in same cluster as Spar")
# Spar, cross species
p_spar_cross <- plotPermutationTestOn0to1Scale(.observed_value = test$observed_crossspecies_Spar,
                                    .null_vector = test$null_crossspecies_Spar,
                                    .xlab = "fraction of Hyc genes in same cluster as Spar")
ggarrange(p_scer_same, p_scer_cross, p_spar_same, p_spar_cross, nrow = 2, ncol = 2)
```

Repeat for checking the direction of level differences

```{r getLevelCrossSpeciesDataset}
getLevelCrossSpeciesDataset <- function(.df, .nIter = 10000) {
  nGenes <- nrow(.df)
  parental_signs <- .df$effect_size_species |> sign()
  hybrid_signs <- .df$effect_size_allele |> sign()
  observed_same <- sum(parental_signs == hybrid_signs)/nGenes
  observed_cross <- sum(parental_signs == -hybrid_signs)/nGenes
  null_same <- purrr::map(c(1:.nIter), \(i) {
    scrambled_hybrid_signs <- sample(hybrid_signs, size = nGenes, replace = FALSE)
    return(sum(parental_signs == scrambled_hybrid_signs)/nGenes)
  }) |> unlist()
  null_cross <- purrr::map(c(1:.nIter), \(i) {
    scrambled_hybrid_signs <- sample(-hybrid_signs, size = nGenes, replace = FALSE)
    return(sum(parental_signs == scrambled_hybrid_signs)/nGenes)
  }) |> unlist()
  return(list(observed_samespecies = observed_same,
              observed_crossspecies = observed_cross,
              null_samespecies = null_same,
              null_crossspecies = null_cross))
}
# tests for getLevelCrossSpeciesDataset
test <- getLevelCrossSpeciesDataset(.df = filter(finaldf, level == "diverged"), # change btween conserved/diverged plasticity or filter for different environments
                                         .nIter = 100)
# Same species
p_same <- plotPermutationTestOn0to1Scale(.observed_value = test$observed_samespecies,
                              .null_vector = test$null_samespecies,
                              .xlab = "fraction of hybrid alleles with same l2fc direction as same parent")
# Cross species
p_cross <- plotPermutationTestOn0to1Scale(.observed_value = test$observed_crossspecies,
                                    .null_vector = test$null_crossspecies,
                                    .xlab = "fraction of hybrid alleles with same l2fc direction as cross parent")
ggarrange(p_same, p_cross, nrow = 1, ncol = 2)
```

#### Permutation plots for supplemental figure

Collecting permutation datasets

```{r cross-species-permutation-datasets, cache=TRUE}
# level, conserved
lev_cons_df <- getLevelCrossSpeciesDataset(.df = filter(finaldf, level == "conserved"))
# level, diverged
lev_div_df <- getLevelCrossSpeciesDataset(.df = filter(finaldf, level == "diverged"))
# plasticity, conserved
plast_cons_df <- getPlasticityCrossSpeciesDataset(.df = filter(finaldf, plasticity == "conserved"))
# plasticity, diverged
plast_div_df <- getPlasticityCrossSpeciesDataset(.df = filter(finaldf, plasticity == "diverged"))
```

Plots for pdf:

```{r pdf-cross-species-permutation}
# Conserved level
# Same species
p_lev_cons_same <- plotPermutationTestOn0to1Scale(.observed_value = lev_cons_df$observed_samespecies,
                                                  .null_vector = lev_cons_df$null_samespecies,
                                                  .xlab = "same parent, conserved level")
# Cross species
p_lev_cons_cross <- plotPermutationTestOn0to1Scale(.observed_value = lev_cons_df$observed_crossspecies,
                                                  .null_vector = lev_cons_df$null_crossspecies,
                                                  .xlab = "cross parent, conserved level")
# Diverged level
# Same species
p_lev_div_same <- plotPermutationTestOn0to1Scale(.observed_value = lev_div_df$observed_samespecies,
                                                  .null_vector = lev_div_df$null_samespecies,
                                                  .xlab = "same parent, diverged level")
# Cross species
p_lev_div_cross <- plotPermutationTestOn0to1Scale(.observed_value = lev_div_df$observed_crossspecies,
                                                  .null_vector = lev_div_df$null_crossspecies,
                                                  .xlab = "cross parent, diverged level")
# Conserved plasticity
# Same species
p_plast_cons_same <- plotPermutationTestOn0to1Scale(.observed_value = plast_cons_df$observed_samespecies,
                                                  .null_vector = plast_cons_df$null_samespecies,
                                                  .xlab = "same parent, conserved plasticity")
# Cross species
p_plast_cons_cross <- plotPermutationTestOn0to1Scale(.observed_value = plast_cons_df$observed_crossspecies,
                                                     .null_vector = plast_cons_df$null_crossspecies,
                                                     .xlab = "cross parent, conserved plasticity")
# Diverged plasticity
# Same species
p_plast_div_same <- plotPermutationTestOn0to1Scale(.observed_value = plast_div_df$observed_samespecies,
                                                   .null_vector = plast_div_df$null_samespecies,
                                                   .xlab = "same parent, diverged plasticity")
# Cross species
p_plast_div_cross <- plotPermutationTestOn0to1Scale(.observed_value = plast_div_df$observed_crossspecies,
                                                    .null_vector = plast_div_df$null_crossspecies,
                                                    .xlab = "cross parent, diverged plasticity")
ggarrange(p_lev_cons_same, p_lev_cons_cross, 
          p_lev_div_same, p_lev_div_cross,
          p_plast_cons_same, p_plast_cons_cross, 
          p_plast_div_same, p_plast_div_cross, nrow = 4, ncol = 2)
ggsave(filename = "paper_figures/CisTrans/cross_species_tests.pdf", width = 10, height = 7)
```



## Maybe: Direction Tests to compare cis basis of level divergence vs plasticity divergence

Level direction test: fraction of Scer and Spar hybrid genes that have the same direction of expression between parents and hybrid alleles

Plasticity direction test: fraction of Scer and Spar hybrid genes that are in the same cluster between parents and hybrid alleles

## Log2 fold changes between parents and hybrid are correlated

```{r level-explained, warning=FALSE}
# R squared for effect sizes between parents and hybrid for all genes
mod_full <- lm(effect_size_allele ~ effect_size_species, data = finaldf)
summary(mod_full)
# R squared genes with consistent effect size and diverged level in at least one environment
levdiv_genes <-  filter(finaldf, level == "diverged") |> 
  dplyr::select(gene_name) |> pull() |> unique()
plotdf <- finaldf |> # filter(gene_name %in% levdiv_genes) |> 
  group_by(gene_name) # |> summarise(n_lev_dir = length(sign(effect_size_species))) |> 
  #filter(n_lev_dir == 1)
mod <- lm(effect_size_allele ~ effect_size_species, data = filter(finaldf, gene_name %in% levdiv_genes))
summary(mod) # more correlated for genes with the most consistent level divergence across envrironments

# looping through each experiment
library(ggExtra)
plotlist <- vector(mode = "list", length = length(ExperimentNames))
names(plotlist) <- ExperimentNames
for (e in ExperimentNames) {
  e_color <- colordf[colordf$limits == e,]$type
  plotdf <- finaldf |> filter(experiment == e) # & level == "diverged")
  mod_e <- lm(effect_size_allele ~ effect_size_species, data = plotdf)
  plotlist[[e]] <- ggMarginal(ggplot(plotdf, aes(x = effect_size_species,
                                 y = effect_size_allele)) +
               geom_vline(xintercept = 0, alpha = 0.5) +
               geom_hline(yintercept = 0, alpha = 0.5) +
               geom_abline(slope = 1, intercept = 0, alpha = 0.5) +
               geom_point(aes(color = level == "diverged"), alpha = 0.25, shape = 19) +
               # geom_text(x = -2, y = 4.5, 
               #           label = paste0("R^2 = ", round(summary(mod_e)$r.squared,
               #                                          digits = 3))) +
               scale_color_discrete(limits = c(TRUE, FALSE),
                                                     type = c("black", e_color)) +
               xlim(c(-5, 5)) +
               ylim(c(-5, 5)) +
               xlab("") +
               ylab("") +
               ggtitle(paste(LongExperimentNames[which(ExperimentNames == e)],
                       paste0(",\nR^2 = ", round(summary(mod_e)$r.squared,
                                                digits = 3)))) +
               theme_classic() +
               theme(legend.position = "none"),
             groupColour = TRUE, groupFill = TRUE)
}
```

Outputting plots to PDF:

```{r level-explained-pdf}
pdf("paper_figures/CisTrans/l2fc_eachEnvironment.pdf",
    width = 16, height = 3)
ggarrange(plotlist = plotlist, nrow = 1, ncol = 6)
dev.off()
ggarrange(plotlist = plotlist, nrow = 3, ncol = 2)
```

Lumping all environments into one figure for main paper:

```{r level-explained-all-environments, warning=FALSE}
  plotdf <- finaldf |> slice_sample(n = 10000)
  mod <- lm(effect_size_allele ~ effect_size_species, data = finaldf)
  plotdf$diverged <- if_else(plotdf$level == "diverged",
                             true = "diverged", false = "conserved")
  p <- ggMarginal(ggplot(plotdf, aes(x = effect_size_species,
                                     y = effect_size_allele)) +
                    geom_vline(xintercept = 0, alpha = 0.5) +
                    geom_hline(yintercept = 0, alpha = 0.5) +
                    geom_abline(slope = 1, intercept = 0, alpha = 0.5) +
                    geom_point(aes(shape = diverged, color = diverged)) +
                    scale_shape_manual(values = c("diverged" = 8, 
                                                  "conserved" = 1)) +
                    scale_color_manual(values = c("diverged" = "black", 
                                                  "conserved" = "grey80")) +
                    xlim(c(-5, 5)) +
                    ylim(c(-5, 5)) +
                    xlab("log2 fold change between parental species") +
                    ylab("log2 fold change between hybrid alleles") +
                    ggtitle(paste0("R^2 = ", round(summary(mod)$r.squared,
                                                      digits = 3))) +
                    theme_classic() +
                    theme(legend.position = "none"),
             groupColour = TRUE, groupFill = TRUE)
  p
```

outputting to PDF

```{r level-explained-all-environments-pdf}
pdf("paper_figures/CisTrans/l2fc_allEnvironments.pdf",
    width = 4, height = 4)
p
dev.off()
```

## Less of the plasticity divergence observed between parents is explained by the hybrid, using R-squared

```{r plasticity-unexplained, warning=FALSE}
# R squared for effect sizes between parents and hybrid for all genes
mod_full <- lm(cor_hybrid ~ cor_parents, data = finaldf)
summary(mod_full)
plotdf <- finaldf

# looping through each experiment
library(ggExtra)
plotlist <- vector(mode = "list", length = length(ExperimentNames))
names(plotlist) <- ExperimentNames
for (e in ExperimentNames) {
  e_color <- colordf[colordf$limits == e,]$type
  plotdf <- finaldf |> filter(experiment == e)
  mod_e <- lm(cor_hybrid ~ cor_parents, data = plotdf)
  plotlist[[e]] <- ggMarginal(ggplot(plotdf, aes(x = cor_parents,
                                                 y = cor_hybrid)) +
                                geom_vline(xintercept = 0, alpha = 0.5) +
                                geom_hline(yintercept = 0, alpha = 0.5) +
                                geom_abline(slope = 1, intercept = 0, alpha = 0.5) +
                                geom_point(aes(color = plasticity == "diverged"), alpha = 0.25, shape = 19) +
                                # geom_text(x = -0.5, y = 0.8, 
                                #           label = paste0("R^2 = ", round(summary(mod_e)$r.squared,
                                #                                          digits = 3))) +
                                scale_color_discrete(limits = c(TRUE, FALSE),
                                                     type = c("black", e_color)) +
               xlim(c(-1, 1)) +
               ylim(c(-1, 1)) +
               xlab("") +
               ylab("") +
               ggtitle(paste(LongExperimentNames[which(ExperimentNames == e)],
                       paste0(",\nR^2 = ", round(summary(mod_e)$r.squared,
                                                digits = 3)))) +
               theme_classic() +
               theme(legend.position = "none"),
             groupColour = TRUE, groupFill = TRUE)
}
```
```{r plasticity-unexplained-pdf}
pdf("paper_figures/CisTrans/cor_eachEnvironment.pdf",
    width = 16, height = 3)
ggarrange(plotlist = plotlist, nrow = 1, ncol = 6)
dev.off()
ggarrange(plotlist = plotlist, nrow = 3, ncol = 2)
```

Lumping all environments into one figure for main paper:

```{r plasticity-unexplained-all-environments, warning=FALSE}
  plotdf <- finaldf |> slice_sample(n = 10000)
  mod <- lm(cor_hybrid ~ cor_parents, data = finaldf)
  plotdf$diverged <- if_else(plotdf$plasticity == "diverged",
                             true = "diverged", false = "conserved")
  p <- ggMarginal(ggplot(plotdf, aes(x = cor_parents,
                                     y = cor_hybrid)) +
                    geom_vline(xintercept = 0, alpha = 0.5) +
                    geom_hline(yintercept = 0, alpha = 0.5) +
                    geom_abline(slope = 1, intercept = 0, alpha = 0.5) +
                    geom_point(aes(shape = diverged, color = diverged)) +
                    scale_shape_manual(values = c("diverged" = 8, 
                                                  "conserved" = 1)) +
                    scale_color_manual(values = c("diverged" = "black", 
                                                  "conserved" = "grey80")) +
                    xlim(c(-1, 1)) +
                    ylim(c(-1, 1)) +
                    xlab("correlation between parental species") +
                    ylab("correlation between hybrid alleles") +
                    ggtitle(paste0("R^2 = ", round(summary(mod)$r.squared,
                                                      digits = 3))) +
                    theme_classic() +
                    theme(legend.position = "none"),
             groupColour = TRUE, groupFill = TRUE)
  p
```

outputting to PDF

```{r plasticity-unexplained-all-environments-pdf}
pdf("paper_figures/CisTrans/cor_allEnvironments.pdf",
    width = 4, height = 4)
p
dev.off()
```

## Alternative measurement of plasticity similarity: Area Between Curves

Repeating parents vs hybrid R^2 plot with a different measure of plasticity conservation: area between curves. Should better account for lowly varying but conserved genes

```{r area-between-curves}
# R squared for effect sizes between parents and hybrid for all genes
mod_full <- lm(ABC_hybrid ~ ABC_parents, data = finaldf)
summary(mod_full)
plotdf <- finaldf

# looping through each experiment
library(ggExtra)
plotlist <- vector(mode = "list", length = length(ExperimentNames))
names(plotlist) <- ExperimentNames
for (e in ExperimentNames) {
  e_color <- colordf[colordf$limits == e,]$type
  plotdf <- finaldf |> filter(experiment == e)
  mod_e <- lm(ABC_hybrid ~ ABC_parents, data = plotdf)
  plotlist[[e]] <- ggMarginal(ggplot(plotdf, aes(x = ABC_parents,
                                                 y = ABC_hybrid)) +
                                geom_vline(xintercept = 0, alpha = 0.5) +
                                geom_hline(yintercept = 0, alpha = 0.5) +
                                geom_abline(slope = 1, intercept = 0, alpha = 0.5) +
                                geom_point(aes(color = plasticity == "diverged"), alpha = 0.25, shape = 19) +
                                # geom_text(x = -0.5, y = 0.8, 
                                #           label = paste0("R^2 = ", round(summary(mod_e)$r.squared,
                                #                                          digits = 3))) +
                                scale_color_discrete(limits = c(TRUE, FALSE),
                                                     type = c("black", e_color)) +
               #xlim(c(0, 1)) +
               #ylim(c(0, 1)) +
               xlab("") +
               ylab("") +
               ggtitle(paste(LongExperimentNames[which(ExperimentNames == e)],
                       paste0(",\nR^2 = ", round(summary(mod_e)$r.squared,
                                                digits = 3)))) +
               theme_classic() +
               theme(legend.position = "none"),
             groupColour = TRUE, groupFill = TRUE)
}
ggarrange(plotlist = plotlist, nrow = 2, ncol = 3)
```

```{r area-between-curves-unexplained-pdf}
pdf("paper_figures/CisTrans/abc_eachEnvironment.pdf",
    width = 16, height = 3)
ggarrange(plotlist = plotlist, nrow = 1, ncol = 6)
dev.off()
```

### Area between curves Hyc-Hyp vs Hyc-Scer and Hyp-Spar

Do hybrid alleles have a smaller area between curves to each other than to parent? 

Rationale: The average expression shows hybird alleles tend to fall together in the middle of the parental expression. But it is possible that this is because half the time they match one parent and half the time they match the other parent. Is that true? Or are they truly in the middle?

If it is true, the Hyb vs Hyc-Scer plot will show a group of genes that have lower abc for Hyc-Scer than they do for Hyb (Hyc-Hyp) a different set of genes will show the same for Hyp-Spar

```{r abc-hyb-vs-allele}

```


## Average expression examples

Examples of Diauxic Shift environment for main paper:

```{r diauxic-level-avg-expr}
plotdf <- finaldf |> 
  filter(level == "diverged") |> 
  mutate(is_trans_upcer = (effect_size_species > trans_lev_thresh) &
           (abs(effect_size_allele) < eff_thresh),
         is_trans_uppar = (effect_size_species < -trans_lev_thresh) &
           (abs(effect_size_allele) < eff_thresh))
plotdf |> mutate(is_trans = is_trans_upcer | is_trans_uppar) |> 
  group_by(experiment) |> 
  summarise(pct_trans = sum(is_trans)/n(),
            pct_rest = sum(!is_trans)/n())
# upcer, both cis and trans
gene_idxs <- filter(plotdf, experiment == "HAP4" & cer == 1 & par == 1 & effect_size_species > 0) |> 
  dplyr::select(gene_name) |> pull()
p <- annotate_figure(plotGenes(gene_idxs, .quartet = TRUE,
                               .experiment_name = "HAP4"), 
                     top = paste(length(gene_idxs), "genes"))
p
pdf("paper_figures/CisTrans/level_all_HAP411_upcer.pdf", width = 2.5, height = 3)
p
dev.off()

# upcer, trans
gene_idxs <- filter(plotdf, experiment == "HAP4" & cer == 1 & 
                      par == 1 & is_trans_upcer) |> 
  dplyr::select(gene_name) |> pull()
p <- annotate_figure(plotGenes(gene_idxs, .quartet = TRUE,
                               .experiment_name = "HAP4"), 
                     top = paste(length(gene_idxs), "genes"))
p
pdf("paper_figures/CisTrans/level_trans_HAP411_upcer.pdf", width = 3, height = 3)
p
dev.off()
  
# upcer, cis
gene_idxs <- filter(plotdf, experiment == "HAP4" & cer == 1 & 
                      par == 1 & !is_trans_upcer & !is_trans_uppar &
                        effect_size_species > 0) |> 
  dplyr::select(gene_name) |> pull()
p <- annotate_figure(plotGenes(gene_idxs, .quartet = TRUE,
                               .experiment_name = "HAP4"), 
                     top = paste(length(gene_idxs), "genes"))
p
pdf("paper_figures/CisTrans/level_cis_HAP411_upcer.pdf", width = 3, height = 3)
p
dev.off()
```

```{r diauxic-plasticity-avg-expr}
plotdf <- finaldf |> 
  filter(plasticity == "diverged") |> 
  mutate(is_cis = cor_hybrid < cis_plast_thresh)
plotdf |> group_by(experiment) |> 
  summarise(cis = sum(is_cis, na.rm = TRUE)/n(),
            trans = sum(!is_cis, na.rm = TRUE)/n())
# both cis and trans
gene_idxs <- filter(plotdf, experiment == "HAP4" & cer == 2 & 
                        par == 1) |> 
    dplyr::select(gene_name) |> pull()
p <- annotate_figure(plotGenes(gene_idxs, .quartet = TRUE,
                                    .experiment_name = "HAP4"), 
                          top = paste(length(gene_idxs), "genes"))
p
pdf("paper_figures/CisTrans/plasticity_all_HAP421.pdf",
        width = 2.5, height = 3)
p
dev.off()

# cis
gene_idxs <- filter(plotdf, experiment == "HAP4" & cer == 2 & 
                        par == 1 & is_cis) |> 
    dplyr::select(gene_name) |> pull()
p <- annotate_figure(plotGenes(gene_idxs, .quartet = TRUE,
                                    .experiment_name = "HAP4"), 
                          top = paste(length(gene_idxs), "genes"))
p
pdf("paper_figures/CisTrans/plasticity_cis_HAP421.pdf",
        width = 3, height = 3)
p
dev.off()

# trans
gene_idxs <- filter(plotdf, experiment == "HAP4" & cer == 2 & 
                        par == 1 & !is_cis) |> 
    dplyr::select(gene_name) |> pull()
p <- annotate_figure(plotGenes(gene_idxs, .quartet = TRUE,
                                    .experiment_name = "HAP4"), 
                          top = paste(length(gene_idxs), "genes"))
p
pdf("paper_figures/CisTrans/plasticity_trans_HAP421.pdf",
        width = 3, height = 3)
p
dev.off()
```

All environments for Supplement:

```{r avg-expression-all-environments-level, message=FALSE}
# UpScer
plotdf <- finaldf |> filter(level == "diverged" & effect_size_species > 0)
plotlist <- vector(mode = "list", length = 6)
names(plotlist) <- ExperimentNames
for (e in ExperimentNames) {
  gene_idxs <- plotdf |> filter(experiment == e) |> 
  dplyr::select(gene_name) |> pull()
  plotlist[[e]] <- annotate_figure(plotGenes(.gene_idxs = gene_idxs,
                                             .quartet = TRUE,
                                             .experiment_name = e),
                                   top = paste(length(gene_idxs), "genes"))
}
ggarrange(plotlist = plotlist, nrow = 1, ncol = 6)
pdf("paper_figures/CisTrans/avg_expr_all_environments_upScer.pdf",
    width = 16, height = 3)
ggarrange(plotlist = plotlist, nrow = 1, ncol = 6)
dev.off()

# UpSpar
plotdf <- finaldf |> filter(level == "diverged" & effect_size_species < 0)
plotlist <- vector(mode = "list", length = 6)
names(plotlist) <- ExperimentNames
for (e in ExperimentNames) {
  gene_idxs <- plotdf |> filter(experiment == e) |> 
  dplyr::select(gene_name) |> pull()
  plotlist[[e]] <- annotate_figure(plotGenes(.gene_idxs = gene_idxs,
                                             .quartet = TRUE,
                                             .experiment_name = e),
                                   top = paste(length(gene_idxs), "genes"))
}
ggarrange(plotlist = plotlist, nrow = 1, ncol = 6)
pdf("paper_figures/CisTrans/avg_expr_all_environments_upSpar.pdf",
    width = 16, height = 3)
ggarrange(plotlist = plotlist, nrow = 1, ncol = 6)
dev.off()
```
```{r avg-expression-all-environments-plasticity, message=FALSE}
# 1-2
plotdf <- finaldf |> filter(cer == 1 & par == 2)
plotlist <- vector(mode = "list", length = 6)
names(plotlist) <- ExperimentNames
for (e in ExperimentNames) {
  gene_idxs <- plotdf |> filter(experiment == e) |> 
  dplyr::select(gene_name) |> pull()
  plotlist[[e]] <- annotate_figure(plotGenes(.gene_idxs = gene_idxs,
                                             .quartet = TRUE,
                                             .experiment_name = e,
                                             .normalization = "scale"),
                                   top = paste(length(gene_idxs), "genes"))
}
ggarrange(plotlist = plotlist, nrow = 1, ncol = 6)
pdf("paper_figures/CisTrans/avg_expr_all_environments_12.pdf",
    width = 16, height = 3)
ggarrange(plotlist = plotlist, nrow = 1, ncol = 6)
dev.off()

# 2-1
plotdf <- finaldf |> filter(cer == 2 & par == 1)
plotlist <- vector(mode = "list", length = 6)
names(plotlist) <- ExperimentNames
for (e in ExperimentNames) {
  gene_idxs <- plotdf |> filter(experiment == e) |> 
  dplyr::select(gene_name) |> pull()
  plotlist[[e]] <- annotate_figure(plotGenes(.gene_idxs = gene_idxs,
                                             .quartet = TRUE,
                                             .experiment_name = e,
                                             .normalization = "scale"),
                                   top = paste(length(gene_idxs), "genes"))
}
ggarrange(plotlist = plotlist, nrow = 1, ncol = 6)
pdf("paper_figures/CisTrans/avg_expr_all_environments_21.pdf",
    width = 16, height = 3)
ggarrange(plotlist = plotlist, nrow = 1, ncol = 6)
dev.off()
```


## Compensatory changes are more common in plasticity than in level

Quick quantification of number of genes with large expression divergence (level or plasticity) in the hybrid despite being conserved in parents.

```{r compensatory-level}
### level
plotdf <- finaldf |> filter(level == "conserved") |> 
  dplyr::select(gene_name, experiment, effect_size_species, effect_size_allele)
# establishing parental range (2-tailed)
#low_bounds <- mean(plotdf$effect_size_species) - 2*sd(plotdf$effect_size_species)
#high_bounds <- mean(plotdf$effect_size_species) + 2*sd(plotdf$effect_size_species)
low_bounds1 <- -eff_thresh
high_bounds1 <- eff_thresh
low_bounds2 <- -eff_thresh*1.5
high_bounds2 <- eff_thresh*1.5
low_bounds3 <- -eff_thresh*2
high_bounds3 <- eff_thresh*2
# choosing low and high bounds
low_bounds <- low_bounds3
high_bounds <- high_bounds3
plot_filename <- "paper_figures/CisTrans/compensatory_level3.pdf"

plotdf$level_compensatory <- plotdf$effect_size_allele < low_bounds |
  plotdf$effect_size_allele > high_bounds
plot_counts <- plotdf |> 
  group_by(experiment, level_compensatory) |> 
  summarise(nComp_Scer = sum(effect_size_allele > 0),
            nComp_Spar = sum(effect_size_allele < 0),
            n = n())
LongExperimentNamesMod <- map(ExperimentNames, \(e) {
  nGenes <- plot_counts |> filter(experiment == e) |> dplyr::select(n) |> pull() |> sum()
  return(paste0(LongExperimentNames[which(ExperimentNames == e)], "\nn = ", nGenes))})
p <- ggplot(plotdf, aes(x = effect_size_species, fill = experiment)) +
  geom_density() +
  geom_vline(data = filter(plotdf, level_compensatory),
             aes(xintercept = effect_size_allele), color = "black") +
  facet_wrap(~factor(experiment, levels = ExperimentNames,
                     labels = LongExperimentNamesMod)) +
  scale_fill_discrete(limits = colordf$limits[colordf$scheme == "experiment"],
                      type = colordf$type[colordf$scheme == "experiment"]) +
  scale_color_discrete(limits = colordf$limits[colordf$scheme == "experiment"],
                       type = colordf$type[colordf$scheme == "experiment"]) +
  # nCompensatory, higher in Spar allele
  geom_label(data = filter(plot_counts, level_compensatory),
             x = -3, y = 0.75, size = 5, 
            aes(label = paste0("n = ", nComp_Spar))) +
  # nCompensatory, higher in Scer allele
  geom_label(data = filter(plot_counts, level_compensatory), 
             x = 3, y = 0.75, size = 5, 
            aes(label = paste0("n = ", nComp_Scer))) +
  theme_classic() +
  theme(legend.position = "none",
        strip.text = element_text(size = 14)) +
  xlim(c(-5, 5))
p
pdf(plot_filename,
    width = 7, height = 4)
p
dev.off()
```

```{r compensatory-plasticity}
# need to drop these genes before quantifying
sum(is.na(finaldf$cor_hybrid)) 
sum(is.na(finaldf$cor_parents))
### plasticity
plotdf <- finaldf |> filter(!is.na(cor_hybrid) & !is.na(cor_parents)) |> 
  filter(plasticity == "conserved" & cor_parents > 0.75) |> # using a threshold to make it more comparable to level
  dplyr::select(gene_name, experiment, cor_parents, cor_hybrid)
# establishing parental range (1-tailed)
#low_bounds <- mean(plotdf$cor_parents) - 2*sd(plotdf$cor_parents)
low_bounds1 <- 0.75
low_bounds2 <- 0.625
low_bounds3 <- 0.5
# choosing bounds
low_bounds <- low_bounds3
plot_filename <- "paper_figures/CisTrans/compensatory_plasticity3.pdf"

plotdf$plasticity_compensatory <- plotdf$cor_hybrid < low_bounds
plot_counts <- plotdf |> 
  group_by(experiment, plasticity_compensatory) |> 
  summarise(n = n())
LongExperimentNamesMod <- map(ExperimentNames, \(e) {
  nGenes <- plot_counts |> filter(experiment == e) |> dplyr::select(n) |> pull() |> sum()
  return(paste0(LongExperimentNames[which(ExperimentNames == e)], "\nn = ", nGenes))})
p <- ggplot(plotdf, aes(x = cor_parents, fill = experiment)) +
  geom_density() +
  geom_vline(data = filter(plotdf, plasticity_compensatory),
             aes(xintercept = cor_hybrid), color = "black", alpha = 0.25) +
  facet_wrap(~factor(experiment, levels = ExperimentNames,
                     labels = LongExperimentNamesMod)) +
  scale_fill_discrete(limits = colordf$limits[colordf$scheme == "experiment"],
                      type = colordf$type[colordf$scheme == "experiment"]) +
  scale_color_discrete(limits = colordf$limits[colordf$scheme == "experiment"],
                       type = colordf$type[colordf$scheme == "experiment"]) +
  # nCompensatory
  geom_label(data = filter(plot_counts, plasticity_compensatory),
             x = -0.5, y = 6, size = 6,
            aes(label = paste0("n = ", n))) +
  theme_classic() +
  theme(legend.position = "none",
        strip.text = element_text(size = 14)) +
  xlim(c(-1.25, 1.25))
p
pdf(plot_filename,
    width = 7, height = 4)
p
dev.off()
```

Repeating with Area Between Curves

```{r compensatory-plasticity-area-between-curves}
# need to drop these genes before quantifying
sum(is.na(finaldf$ABC_hybrid)) 
sum(is.na(finaldf$ABC_parents))
### plasticity
plotdf <- finaldf |> filter(!is.na(ABC_hybrid) & !is.na(ABC_parents)) |> 
  filter(plasticity == "conserved" & ABC_parents < 0.25) |> # using a threshold to make it more comparable to level
  dplyr::select(gene_name, experiment, ABC_parents, ABC_hybrid)
# establishing parental range (1-tailed)
#low_bounds <- mean(plotdf$ABC_parents) - 2*sd(plotdf$ABC_parents)
high_bounds <- 0.5
plotdf$plasticity_compensatory <- plotdf$ABC_hybrid > high_bounds
plot_counts <- plotdf |> 
  group_by(experiment, plasticity_compensatory) |> 
  summarise(n = n())
LongExperimentNamesMod <- map(ExperimentNames, \(e) {
  nGenes <- plot_counts |> filter(experiment == e) |> dplyr::select(n) |> pull() |> sum()
  return(paste0(LongExperimentNames[which(ExperimentNames == e)], "\nn = ", nGenes))})
p <- ggplot(plotdf, aes(x = ABC_parents, fill = experiment)) +
  geom_density() +
  geom_vline(data = filter(plotdf, plasticity_compensatory),
             aes(xintercept = ABC_hybrid), color = "black", alpha = 0.25) +
  facet_wrap(~factor(experiment, levels = ExperimentNames,
                     labels = LongExperimentNamesMod)) +
  scale_fill_discrete(limits = colordf$limits[colordf$scheme == "experiment"],
                      type = colordf$type[colordf$scheme == "experiment"]) +
  scale_color_discrete(limits = colordf$limits[colordf$scheme == "experiment"],
                       type = colordf$type[colordf$scheme == "experiment"]) +
  # nCompensatory
  geom_label(data = filter(plot_counts, plasticity_compensatory),
             x = 0.5, y = 6, size = 6,
            aes(label = paste0("n = ", n))) +
  theme_classic() +
  theme(legend.position = "none",
        strip.text = element_text(size = 14))
p
pdf("paper_figures/CisTrans/compensatory_plasticityABC.pdf",
    width = 7, height = 4)
p
dev.off()
```

## Hybrid allele cor relative to min(cor_Scer, cor_Spar)

```{r cor-hyb-vs-scer-spar}
plotdf <- finaldf |> dplyr::select(cor_hybrid, cor_scer, cor_spar, experiment)
plotdf$min_cor <- pmin(plotdf$cor_scer, plotdf$cor_spar)
plotdf$var_in_cis <- plotdf$cor_hybrid < plotdf$min_cor
ggplot(plotdf, aes(x = min_cor, y = cor_hybrid)) +
  geom_point(aes(color = experiment)) + facet_wrap(~experiment)
```


# %%%%%%%%%%%%%%%%% probably archive from here down %%%%%%%%%%%%%%%%% 

## Hybrids have more correlated plasticity between alleles than parental orthologs

```{r plasticity-cor}
#### Distribution of parent/hybrid expression correlation ####
plotlist <- vector(mode = "list", length = length(ExperimentNames))
names(plotlist) <- ExperimentNames
# density plots of parental vs hybrid cor in each environment
LongExperimentNamesMod <- gsub("Hydroxyurea Shock", "Hydroxyurea\nShock", LongExperimentNames)
for (e in ExperimentNames) {
  plotdf <- finaldf |> filter(experiment == e & plasticity == "diverged") |> 
    dplyr::select(gene_name, experiment, cor_parents, cor_hybrid) |> 
    pivot_longer(cols = c("cor_parents", "cor_hybrid"), 
                 names_to = "parent_or_hybrid",
                 values_to = "cor")
  plotlist[[e]] <- ggplot(plotdf, aes(x = cor)) + 
    geom_density(data = filter(plotdf, parent_or_hybrid == "cor_parents"), 
                 aes(fill = experiment), alpha = 1, color = "white") +
    geom_density(data = filter(plotdf, parent_or_hybrid == "cor_hybrid"), 
                 aes(fill = experiment), alpha = 0.5, color = "black") +
    scale_fill_discrete(limits = colordf[colordf$scheme == "experiment",]$limits,
                        type = colordf[colordf$scheme == "experiment",]$type) +
    xlim(c(-1, 1)) +
    xlab("") +
    ylab("") +
    ggtitle(LongExperimentNamesMod[which(ExperimentNames == e)]) +
    theme_classic() +
    theme(legend.position = "none")
}
p_null <- ggplot(plotdf) + theme_void()
pdf("paper_figures/CisTrans/cor_density_eachEnvironment.pdf",
    width = 7.5, height = 4)
ggarrange(plotlist[[1]], plotlist[[2]],
          plotlist[[3]], plotlist[[4]],
          p_null, p_null, 
          plotlist[[5]], plotlist[[6]], nrow = 2, ncol = 4)
dev.off()
ggarrange(plotlist[[1]], plotlist[[2]],
          p_null, p_null, plotlist[[3]], plotlist[[4]],
          plotlist[[5]], plotlist[[6]], nrow = 2, ncol = 4)
```

## Level divergence tends to be in cis

What better way to show this than that looking at the small fraction of genes that are diverging in level in trans the most---that are diverging in level in the parents and have the smallest log2 fold change between hybrid alleles.

```{r level-pie-chart}
# pct genes in "transest level" group for figure
plotdf <- finaldf |> 
  filter(level == "diverged") |> 
  mutate(is_trans_upcer = (effect_size_species > trans_lev_thresh) &
           (abs(effect_size_allele) < eff_thresh),
         is_trans_uppar = (effect_size_species < -trans_lev_thresh) &
           (abs(effect_size_allele) < eff_thresh))
plotdf |> mutate(is_trans = is_trans_upcer | is_trans_uppar) |> 
  group_by(experiment) |> 
  summarise(pct_trans = sum(is_trans)/n(),
            pct_rest = sum(!is_trans)/n())
pctdf <- plotdf |> mutate(is_trans = is_trans_upcer | is_trans_uppar) |> 
  group_by(experiment) |> 
  summarise(trans = sum(is_trans)/n(),
            cis = sum(!is_trans)/n()) |> 
  pivot_longer(cols = c("cis", "trans"), names_to = "mode",
               values_to = "percent")
p <- ggplot(pctdf, aes(x="", y=percent, fill=mode)) +
  scale_fill_discrete(limits = c("cis", "trans"),
                      type = c("salmon", "aquamarine")) +
  geom_bar(stat="identity", width=1) +
  coord_polar("y", start=0) +
  theme_classic() +
  theme(axis.ticks = element_blank(),
        axis.text = element_blank(),
        legend.position = "bottom") +
  xlab("") +
  ylab("") +
  facet_wrap(~factor(experiment, levels = ExperimentNames,
                     labels = LongExperimentNames), nrow = 2, ncol = 3)
pdf("paper_figures/CisTrans/level_pcts.pdf",
    width = 8, height = 4)
p
dev.off()
p
```

## Plasticity divergence tends to be in trans

```{r plasticity-pie-chart}
plotdf <- finaldf |> 
  filter(plasticity == "diverged") |> 
  mutate(is_cis = cor_hybrid < cis_plast_thresh)
plotdf |> group_by(experiment) |> 
  summarise(cis = sum(is_cis, na.rm = TRUE)/n(),
            trans = sum(!is_cis, na.rm = TRUE)/n())
pctdf <- plotdf |> group_by(experiment) |> 
  summarise(cis = sum(is_cis, na.rm = TRUE)/n(),
            trans = sum(!is_cis, na.rm = TRUE)/n()) |> 
  pivot_longer(cols = c("cis", "trans"), names_to = "mode",
               values_to = "percent")
p <- ggplot(pctdf, aes(x="", y=percent, fill=mode)) +
  scale_fill_discrete(limits = c("cis", "trans"),
                      type = c("salmon", "aquamarine")) +
  geom_bar(stat="identity", width=1) +
  coord_polar("y", start=0) +
  theme_classic() +
  theme(axis.ticks = element_blank(),
        axis.text = element_blank(),
        legend.position = "bottom") +
  xlab("") +
  ylab("") +
  facet_wrap(~factor(experiment, levels = ExperimentNames,
                     labels = LongExperimentNames), nrow = 2, ncol = 3)
p
pdf("paper_figures/CisTrans/plasticity_pcts.pdf",
    width = 8, height = 4)
p
dev.off()
```
Average expression of Diauxic Shift example:

```{r plasticity-avg-expr}
plotdf <- finaldf |> 
  filter(plasticity == "diverged") |> 
  mutate(is_cis = cor_hybrid < cis_plast_thresh)
plotdf |> group_by(experiment) |> 
  summarise(cis = sum(is_cis, na.rm = TRUE)/n(),
            trans = sum(!is_cis, na.rm = TRUE)/n())
# cis
gene_idxs <- filter(plotdf, experiment == "HAP4" & cer == 2 & 
                        par == 1 & is_cis) |> 
    dplyr::select(gene_name) |> pull()
p <- annotate_figure(plotGenes(gene_idxs, .quartet = TRUE,
                                    .experiment_name = "HAP4"), 
                          top = paste(length(gene_idxs), "genes"))
p
pdf("paper_figures/CisTrans/plasticity_cis_HAP421.pdf",
        width = 3, height = 3)
p
dev.off()

# trans
gene_idxs <- filter(plotdf, experiment == "HAP4" & cer == 2 & 
                        par == 1 & !is_cis) |> 
    dplyr::select(gene_name) |> pull()
p <- annotate_figure(plotGenes(gene_idxs, .quartet = TRUE,
                                    .experiment_name = "HAP4"), 
                          top = paste(length(gene_idxs), "genes"))
p
pdf("paper_figures/CisTrans/plasticity_trans_HAP421.pdf",
        width = 3, height = 3)
p
dev.off()
```

## Permutation tests (probably archive)
Testing whether hybrid has more genes with level and plasticity divergence in common with parents than expected by chance.

Putting these at the bottom of the script because, as expected, they do, but there's also lots of non-overlap.

Permutation test for level, to compare to plasticity below:

```{r level-explained-permutation}
getLevelDivergedGeneSets <- function(.environment, .df = finaldf) {
  diverged_genes_parents <- .df |> filter(experiment == .environment &
                                            level == "diverged") |> 
    dplyr::select(gene_name) |> pull()
  real_diverged_genes_hybrids <- .df |> filter(experiment == .environment &
                                                 finaldf$padj_allele < 0.05 &
                                                 abs(effect_size_allele) > log2(1.5)) |> 
    dplyr::select(gene_name) |> pull()
  return(list(parents = diverged_genes_parents, hybrids = real_diverged_genes_hybrids))
}
getLevelPermutationDataset <- function(.environment, .df = finaldf,
                                            .nIter = 10000) {
  gene_sets <- getLevelDivergedGeneSets(.environment = .environment, .df = .df)
  # measuring observed jaccard index
  observed <- getJaccardIndex(gene_sets$parents, gene_sets$hybrids)
  # creating a null distribution of jaccard indices
  null_set <- purrr::map(c(1:.nIter), \(i) {
    fake_diverged_genes_hybrids <- .df |> filter(experiment == .environment) |> 
      dplyr::select(gene_name) |> pull() |> 
      sample(size = length(gene_sets$hybrids), replace = FALSE)
    return(getJaccardIndex(gene_sets$parents, fake_diverged_genes_hybrids))
  }) |> unlist()
  return(list(observed_jaccard = observed, null_jaccard = null_set))
}
# tests for getPlasticityPermutationDataset
test_dataset <- getLevelPermutationDataset(.environment = "HAP4", .nIter = 10)
unlist(test_dataset)
```
Collecting permutation datasets for each environment:

```{r level-permutation-datasets, cache=TRUE}
df_list_level <- purrr::map(ExperimentNames, \(e) {
  cat(e, "\n")
  getLevelPermutationDataset(.environment = e, .nIter = 10000)
})
```

```{r plotting-level-permutation-tests}
plotlist <- purrr::map(df_list_level, \(df) {
  plotPlasticityPermutationTest(.observed_value = df$observed_jaccard, .null_distribution = df$null_jaccard, .xlab = "Jaccard Index ([A & B]/[A or B])")
})
ggarrange(plotlist = plotlist, ncol = 2, nrow = 3)
```

Accompanying Venn Diagrams:

```{r venn-level}
plotlist_venn <- purrr::map(ExperimentNames, \(e) {
  gene_sets <- getLevelDivergedGeneSets(.environment = e)
  ggVennDiagram(gene_sets) + ggtitle(e)
})
ggarrange(plotlist = plotlist, ncol = 2, nrow = 3, legend = "none")
```

## New version: Less of the plasticity divergence observed between parents is explained by the hybrid, tested using permuted labels

```{r plasticity-unexplained-permutation}
getPlasticityDivergedGeneSets <- function(.environment, .df = finaldf,
                                            .just_reversals = FALSE) {
  if (.just_reversals) {
    diverged_genes_parents <- .df |> filter(experiment == .environment &
                                              plasticity == "diverged" &
                                              cer != 0 & par != 0) |> 
      dplyr::select(gene_name) |> pull()
    real_diverged_genes_hybrids <- .df |> filter(experiment == .environment &
                                                   hyc != hyp &
                                                   hyc != 0 & hyp != 0) |> 
      dplyr::select(gene_name) |> pull()
  }
  if (!.just_reversals) {
    diverged_genes_parents <- .df |> filter(experiment == .environment &
                                              plasticity == "diverged") |> 
      dplyr::select(gene_name) |> pull()
    real_diverged_genes_hybrids <- .df |> filter(experiment == .environment &
                                                   hyc != hyp) |> 
      dplyr::select(gene_name) |> pull()
  }
  return(list(parents = diverged_genes_parents, hybrids = real_diverged_genes_hybrids))
}
getPlasticityPermutationDataset <- function(.environment, .df = finaldf,
                                            .just_reversals = FALSE,
                                            .nIter = 10000) {
  gene_sets <- getPlasticityDivergedGeneSets(.environment = .environment,
                                             .df = .df, .just_reversals = .just_reversals)
  # measuring observed jaccard index
  observed <- getJaccardIndex(gene_sets$parents, gene_sets$hybrids)
  # creating a null distribution of jaccard indices
  null_set <- purrr::map(c(1:.nIter), \(i) {
    fake_diverged_genes_hybrids <- .df |> filter(experiment == .environment) |> 
      dplyr::select(gene_name) |> pull() |> 
      sample(size = length(gene_sets$hybrids), replace = FALSE)
    return(getJaccardIndex(gene_sets$parents, fake_diverged_genes_hybrids))
  }) |> unlist()
  return(list(observed_jaccard = observed, null_jaccard = null_set))
}
# tests for getPlasticityPermutationDataset
test_dataset1 <- getPlasticityPermutationDataset(.environment = "HAP4", .nIter = 10)
test_dataset2 <- getPlasticityPermutationDataset(.environment = "HAP4", .nIter = 10, .just_reversals = TRUE) # should have lower jaccard b/c smaller set
rbind(unlist(test_dataset1), unlist(test_dataset2))
```
Collecting permutation datasets for each environment:

```{r permutation-datasets, cache=TRUE}
df_list_plasticity <- purrr::map(ExperimentNames, \(e) {
  cat(e, "\n")
  getPlasticityPermutationDataset(.environment = e, .nIter = 10000)
})
```
Plotting null distributions and observed Jaccard index for each environment:

```{r plotting-plasticity-permutation-tests}
plotlist <- purrr::map(df_list_plasticity, \(df) {
  plotPlasticityPermutationTest(.observed_value = df$observed_jaccard, .null_distribution = df$null_jaccard, .xlab = "Jaccard Index ([A & B]/[A or B])")
})
ggarrange(plotlist = plotlist, ncol = 2, nrow = 3)
```
Accompanying Venn diagram:

```{r venn-plasticity}
plotlist <- purrr::map(ExperimentNames, \(e) {
  gene_sets <- getPlasticityDivergedGeneSets(.environment = e)
  ggVennDiagram(gene_sets) + ggtitle(e)
})
ggarrange(plotlist = plotlist, ncol = 2, nrow = 3, legend = "none")
```
