---
title: "Exploring plasticity divergence with yestract-defined regulons"
---

Regulators are rows
Targets are columns
i,j = 1 means i regulates j (could be positive or negative or both depending on regulatory matrix used)
i,j = 0 means there is no evidence that i regulates j

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r package-loading}
load("data_files/FinalDataframe3Disp.RData")
load("data_files/Cleaned_Count_Data.RData")
load("data_files/Cleaned_Counts.RData")
load("data_files/Cleaned_Counts_Allele.RData")
sapply(c("Matrix", "stringr", "ComplexHeatmap"), require, character.only=TRUE)
source("functions_for_figure_scripts.R")
library(dplyr)
library(igraph)
library(lmtest)
```

### Before beginning, check that commonly-named functions are associated with the expected package:

```{r package-issues}
environmentName(environment(select)) # should be dplyr
environmentName(environment(degree)) # should be igraph
```


Making the gene list we'll use as the regulation targets we put into yeastract (the regulators are all Scer regulators)

```{r gene-list}
# Making one shared gene idx list
idxs <- rownames(counts)
grep("/", idxs, value = TRUE) # paralog pairs with too similar sequence 
# need to have separate rows for yeastract:
idxs <- map(idxs, \(g) {
  if (grepl("/", g)) {
    g <- strsplit(g, split = "/") |> unlist()
  }
  return(g)
}) |> unlist() |> sort()
write_delim(tibble(gene_name = idxs),
            delim = "\n", file = "data_files/yeastract_genes.txt",
            col_names = FALSE)
```

Table downloaded from yeastract containing names of genes in common and systematic format, used to convert between formats below:

```{r lookup}
# reading in table to convert from common to systematic names
yeastract_lookup <- read_delim("data_files/downloaded_genomes_and_features/yeastract_orftogene.csv",
                               delim = ";", col_names = c("common_protein", 
                                                          "systematic", 
                                                          "common"))
yeastract_lookup$description <- dplyr::select(yeastract_lookup, 
                                       setdiff(colnames(yeastract_lookup),
                                               c("common_protein", 
                                                 "systematic", 
                                                 "common"))) |> 
  apply(MARGIN = 1, \(x) purrr::reduce(x, .f = paste0)) |> as.character()
yeastract_lookup <- dplyr::select(yeastract_lookup,
                           common_protein,
                           systematic,
                           common,
                           description)
# removing duplicate rows
yeastract_lookup[duplicated(yeastract_lookup$common_protein),] # 2 duplicates
yeastract_lookup <- yeastract_lookup[!duplicated(yeastract_lookup$common_protein),]
```

## Reading in regulatory matrices from yeastract

Regulators are rows
Targets are columns
i,j = 1 means i regulates j (could be positive or negative or both depending on regulatory matrix used)
i,j = 0 means there is no evidence that i regulates j

Function to streamline reading in each matrix, given its file location

```{r readInRegulatoryMatrix}
# reading in matrices
readInRegulatoryMatrix <- function(.file_name) {
  matdf <- read_delim(.file_name,
                      delim = ";")
  mat <- matdf[, -1] |> as.matrix()
  colnames(mat) <- left_join(tibble(common = colnames(mat)),
                             yeastract_lookup, by = "common") |> 
    dplyr::select(systematic) |> pull()
  rownames(mat) <- left_join(tibble(common_protein = unlist(matdf[,1])),
                             yeastract_lookup, by = "common_protein") |> 
    dplyr::select(systematic) |> pull()
  return(mat)
}
```

Reading in the un-stressed regulator matrix:

```{r ypd-regmat}
# setting file suffix to determine which matrix type
# file_name_suffix <- "_directOrExpression.csv" # Binding OR expression regulatory matrices
# file_name_suffix <- "_binding.csv" # Binding, with or without expression
file_name_suffix <- "_bindingANDexpression.csv" # Binding AND expression regulatory matrices
# file_name_suffix <- "_Activators_directAndExpression.csv" # Binding AND expression regulatory matrices, positive results on expression only
# No-stress control matrix
ypd <- readInRegulatoryMatrix(.file_name = paste0("data_files/yeastract/Control",
                                                  file_name_suffix))
unique(as.vector(ypd)) # binary matrix
```

Reading in all environment-specific regmats:

```{r env-regmats, cache=TRUE, include=FALSE}
# Diauxic Shift
hap4 <- readInRegulatoryMatrix(.file_name = paste0("data_files/yeastract/HAP4",
                                                   file_name_suffix))
# Cell Cycle
cc <- readInRegulatoryMatrix(.file_name = paste0("data_files/yeastract/CC",
                                                 file_name_suffix))
# Low Nitrogen
lown <- readInRegulatoryMatrix(.file_name = paste0("data_files/yeastract/LowN",
                                                   file_name_suffix))
# Low Phosphate
lowpi <- readInRegulatoryMatrix(.file_name = paste0("data_files/yeastract/LowPi",
                                                    file_name_suffix))
# Heat Shock
heat <- readInRegulatoryMatrix(.file_name = paste0("data_files/yeastract/Heat",
                                                   file_name_suffix))
# Cold Shock
cold <- readInRegulatoryMatrix(.file_name = paste0("data_files/yeastract/Cold",
                                                   file_name_suffix))
```

Our full list of regmats:

```{r regmat-list}
regmats <- list("YPD" = ypd, "HAP4" = hap4,
                # "HAP4_up" = hap4_activators,
                # "HAP4_down" = hap4_inhibitors,
                "CC" = cc, "LowN" = lown, "LowPi" = lowpi, 
                "Heat" = heat, "Cold" = cold)
```

### Filtering out unused regulators

```{r unused-regulators}
# filtering out any regulators (rows) with zero counts in all
# environments and not included in the targets (columns)
zero_count_elements <- rownames(ypd)[rowSums(purrr::reduce(regmats, .f = cbind)) == 0]
sum(zero_count_elements %in% colnames(ypd))
sum(!(zero_count_elements %in% colnames(ypd)))
toRemove <- setdiff(zero_count_elements, colnames(ypd))
regmats <- map(regmats, .f = \(x) {x[setdiff(rownames(x), toRemove),]})
```

### Making regmats square

```{r square}
# filtering out zero count regulators made rows/columns nearly identical
# but there are still a few non-zero regulators that aren't found in columns

# checking for any genes missing from idxs that have network connections
setdiff(rownames(regmats$YPD)[rowSums(regmats$YPD) > 0], idxs) # YER109C = FLO8, YLR256W = HAP1, YOL028C = YAP7 all unannotated b/c of large sequence differences between Scer and other saccharomyces
# Number of genes that are recorded in yeastract as Scer regulators, but
# they are missing in our gene lists: 
sum(!(rownames(regmats$YPD) %in% colnames(regmats$YPD))) # number of regulators
rowSums(regmats$YPD[!(rownames(regmats$YPD) %in% colnames(regmats$YPD)),]) # number of targets for each regulator

### making regulator matrices square
regmats <- map(regmats, \(x) {
  common_idxs <- colnames(x)
  out_mat <- x[common_idxs, common_idxs]
  return(out_mat)
})
```

### Removing unconnected nodes

```{r unconnected-nodes}
# genes with no in or out edges in any environment
zero_count_rows <- rownames(regmats$YPD)[rowSums(purrr::reduce(regmats, .f = cbind)) == 0]
zero_count_cols <- colnames(regmats$YPD)[colSums(purrr::reduce(regmats, .f = rbind)) == 0]
zero_count_genes <- intersect(zero_count_rows, zero_count_cols)
length(zero_count_genes)
regmats <- map(regmats, \(x) {
  out_mat <- x[setdiff(rownames(x), zero_count_genes), 
               setdiff(colnames(x), zero_count_genes)]
  return(out_mat)
})
```

## Visualizing Networks

### How many regulatory connections are found in both each environment and YPD?

```{r regdf}
# creating regdf, each regulator (non-zero rowSum in at least one environment),
# sum and list of its targets, environments it's active in
regdf <- map2(regmats, names(regmats), .f = \(x, y) {
  x_regulators <- rownames(x)[rowSums(x) != 0]
  outdf <- map(x_regulators, \(r) {
    targets <- colnames(x)[x[r,] != 0]
    return(tibble(regulator = r,
                  target = targets,
                  environment = y))
  }) |> purrr::reduce(.f = bind_rows)
  return(outdf)
}) |> purrr::reduce(.f = bind_rows) |> 
  # adding regulator descriptions
  left_join(dplyr::select(yeastract_lookup, systematic, common, description),
                     by = c("target"="systematic")) |> 
  dplyr::rename(c("target_description"="description",
           "target_common"="common")) |> 
  left_join(dplyr::select(yeastract_lookup, systematic, common, description),
            by = c("regulator"="systematic")) |>
  dplyr::rename(c("regulator_description"="description",
           "regulator_common"="common"))

# saving Regdf as Supp table
write_csv(regdf, file = "data_files/Regulator-target.csv")
regdf
```

```{r ypd-overlap}
# How many regulatory connections are shared between each environment and YPD?
makeUpsetPlot(mutate(regdf, pair = paste(regulator, target, sep = "_")), 
              .group_name = "environment", 
              .group_members = names(regmats),
              .item_names = "pair", .min_comb_size = 0)
```

YPD is majority unique connections, but that's mainly because it has so many more than any environment-specific network

all environments except Cold have at least one unique regulator-target pair (Cold only has 2 regulators and 3 targets total)

Note: here is where we could remove env-specific regulatory connections from YPD, if that is deemed appropriate. Because genes have semi-overlapping environmental conditions in which they are active, it feels more biologically meaningful to retain that same partial overlap in their regulatory connections

### Visualizing environment-specific networks

Given the adjacency matrix, outputs the connected portion of the graph in igraph format, good for plotting:

```{r getConnectedGraph}
getConnectedGraph <- function(.adj) {
  colnames(.adj) <- left_join(tibble(systematic = colnames(.adj)),
                             yeastract_lookup, by = "systematic") |> 
    dplyr::select(common) |> pull()
  rownames(.adj) <- left_join(tibble(systematic = rownames(.adj)),
                             yeastract_lookup, by = "systematic") |> 
    dplyr::select(common) |> pull()
  regnames <- rownames(.adj)[rowSums(.adj) > 0]
  g <- graph_from_adjacency_matrix(.adj)
  sub_g <- subgraph(g, which(igraph::degree(g) > 0))
  V(sub_g)$isHub <- names(V(sub_g)) %in% regnames
  V(sub_g)$vertex_size <- if_else(V(sub_g)$isHub, true = 7, false = 1)
  V(sub_g)$color <- if_else(V(sub_g)$isHub, true = "steelblue", false = "orange")
  V(sub_g)$vertex_label <- if_else(V(sub_g)$isHub, true = names(V(sub_g)), false = "")
  return(sub_g)
}
```

Visualizations of each environment-specific network, randomly downsampled if they have too many connections for igraph to handle

```{r igraph-plots}
# plotting each environment
# downsampling to 2500 (there are 3500ish features in ypd) because igraph seems to have a problem plotting
# plot(getConnectedGraph(regmats$YPD)) # better as a heatmap
conn_thresh <- 1500 # number of edges to trigger downsample
down_thresh <- 2500 # number of nodes to downsample to
for (r in setdiff(names(regmats), "YPD")) {
  mat <- regmats[[r]]
  if (sum(mat) > conn_thresh) {
    downsample_idxs <- sample(c(1:nrow(mat)), down_thresh, replace = FALSE)
    g <- getConnectedGraph(mat[downsample_idxs, downsample_idxs])
  }
  else {
    g <- getConnectedGraph(mat)
  }
  # version that makes the actual plot, commented out once it is run once, b/c igraph randomizes plot each time:
  # pdf(paste0("paper_figures/Yeastract/",
  #            r, ".pdf"),
  #     width = 7, height = 7)
  # plot(g, vertex.size = V(g)$vertex_size,
  #      vertex.label = V(g)$vertex_label,
  #      edge.arrow.size = 0.1,
  #      main = r, layout = layout_nicely(g),
  #      frame = TRUE)
  # dev.off()
  # version for website:
  plot(g, vertex.size = V(g)$vertex_size,
       vertex.label = V(g)$vertex_label,
       edge.arrow.size = 0.1,
       main = r, layout = layout_nicely(g),
       frame = TRUE)
}
```

## Supplementary figure: stacked barplots

Proportions of incoming and outgoing connections by plasticity group. Barplots with mutually exclusive plasticity groups:

1) conserved plastic
2) conserved static
3) Scer-unique plastic
4) Spar-unique plastic
5) plasticity reversal
(YPD and env-specific connections in each)

```{r getPlasticityGroup}
# helper for plotting function
getPlasticityGroup <- function(.cer, .par) {
  if (.cer == 0 & .par == 0) {
    return("conserved static")
  }
  if (.cer == .par) {
    return("conserved plastic")
  }
  if (.cer == 0) {
    return("Spar-unique")
  }
  if (.par == 0) {
    return("Scer-unique")
  }
  else {
    return("reversal")
    }
}
```

Plotting (very large amount of dplyr-ing the plot dataframe):

```{r stacked-barplots}
plotdf <- map(ExperimentNames, \(.e) {
  expr_genes <- filter(finaldf, experiment == .e) |> 
    dplyr::select(gene_name) |> pull() |> 
    intersect(colnames(regmats$YPD))
  tibble(gene_name = expr_genes,
         in_degree_e = colSums(regmats[[.e]][expr_genes, expr_genes]),
         in_degree_YPD = colSums(regmats$YPD[expr_genes, expr_genes]),
         out_degree_e = rowSums(regmats[[.e]][expr_genes, expr_genes]),
         out_degree_YPD = rowSums(regmats$YPD[expr_genes, expr_genes])) |> 
    right_join(filter(finaldf, experiment == .e),
              by = "gene_name")
}) |> purrr::reduce(.f = bind_rows)
# NA degree means gene isn't in regmat 
# (likely never had any connections in any environment),
# so setting these to a degree of 0
plotdf$in_degree_e[is.na(plotdf$in_degree_e)] <- 0
plotdf$in_degree_YPD[is.na(plotdf$in_degree_YPD)] <- 0
plotdf$out_degree_e[is.na(plotdf$out_degree_e)] <- 0
plotdf$out_degree_YPD[is.na(plotdf$out_degree_YPD)] <- 0
# getting plasticity group by gene/environment
plotdf$plasticity <- map2(plotdf$cer, plotdf$par, getPlasticityGroup) |> unlist()
totalsdf <- plotdf |> group_by(experiment) |> 
  dplyr::summarise(total_in_e = sum(in_degree_e),
            total_in_YPD = sum(in_degree_YPD),
            total_out_e = sum(out_degree_e),
            total_out_YPD = sum(out_degree_YPD), 
            total_genes = n())
plotdf <- plotdf |> group_by(experiment, plasticity) |> 
  dplyr::summarise(n_in_e = sum(in_degree_e),
            n_in_YPD = sum(in_degree_YPD),
            n_out_e = sum(out_degree_e),
            n_out_YPD = sum(out_degree_YPD),
            n_genes = n()) |> 
  left_join(y =  totalsdf,
            by = "experiment") |> 
  mutate(prop_in_e = n_in_e/total_in_e,
         prop_in_YPD = n_in_YPD/total_in_YPD,,
         prop_out_e = n_out_e/total_out_e,
         prop_out_YPD = n_out_YPD/total_out_YPD,
         prop_genes = n_genes/total_genes) |>
  pivot_longer(cols = c("prop_in_e", "prop_in_YPD", "prop_genes",
                        "prop_out_e", "prop_out_YPD"),
               names_to = "prop_type", values_to = "proportion") |> 
  mutate(degree = if_else(grepl("in", prop_type),
                          true = "in", 
                          false = if_else(grepl("out", prop_type),
                                          true = "out", 
                                          false = "genes")),
         network = if_else(grepl("YPD", prop_type),
                           true = "YPD", 
                           false = if_else(grepl("genes", prop_type),
                                           true = "genes",
                                           false = "env")))
totalsdf <- pivot_longer(totalsdf,
                         cols = c("total_in_e", "total_in_YPD",
                                  "total_out_e", "total_out_YPD",
                                  "total_genes"),
                         names_to = "prop_type", values_to = "n") |> 
  mutate(degree = if_else(grepl("in", prop_type),
                          true = "in", 
                          false = if_else(grepl("out", prop_type),
                                          true = "out", 
                                          false = "genes")),
         network = if_else(grepl("YPD", prop_type),
                           true = "YPD", 
                           false = if_else(grepl("genes", prop_type),
                                           true = "genes",
                                           false = "env")))
plotdf$network <- factor(plotdf$network, levels = c("genes", "YPD", "env"),
                         labels = c("number of genes", "un-stressed network\nconnections",
                                    "environment-specific network\nconnections"))
totalsdf$network <- factor(totalsdf$network, levels = c("genes", "YPD", "env"),
                           labels = c("number of genes", "un-stressed network\nconnections",
                                      "environment-specific network\nconnections"))
# in degree
p_in <- ggplot(filter(plotdf, degree %in% c("in", "genes")), aes(x = network, y = proportion)) +
  geom_bar(aes(fill = plasticity), position = "stack", stat = "identity") +
  geom_text(data = filter(totalsdf, degree %in% c("in", "genes")), aes(x = network, label = n), y = 1.1) +
  ylim(c(0, 1.1)) +
  scale_fill_discrete(limits = colordf[colordf$scheme == "plasticity",]$limits,
                      type = colordf[colordf$scheme == "plasticity",]$type) +
  ylab("%") +
  xlab("") +
  ggtitle("incoming connections") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, hjust = 0.95, vjust = 0.2)) +
  facet_wrap(~factor(experiment, levels = ExperimentNames, labels = LongExperimentNames))

# out degree
p_out <- ggplot(filter(plotdf, degree %in% c("out", "genes")), aes(x = network, y = proportion)) +
  geom_bar(aes(fill = plasticity), position = "stack", stat = "identity") +
  geom_text(data = filter(totalsdf, degree %in% c("out", "genes")), aes(x = network, label = n), y = 1.1) +
  ylim(c(0, 1.1)) +
  scale_fill_discrete(limits = colordf[colordf$scheme == "plasticity",]$limits,
                      type = colordf[colordf$scheme == "plasticity",]$type) +
  ylab("%") +
  xlab("") +
  ggtitle("outgoing connections") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, hjust = 0.95, vjust = 0.2)) +
  facet_wrap(~factor(experiment, levels = ExperimentNames, labels = LongExperimentNames))

# plotting
ggarrange(p_in, p_out, nrow = 1, ncol = 2, common.legend = TRUE, legend = "bottom")
pdf("paper_figures/Supplement/propConnectionsBars.pdf",
    width = 9, height = 7)
ggarrange(p_in, p_out, nrow = 1, ncol = 2, common.legend = TRUE, legend = "bottom")
dev.off()
```

## Waffle plots of the number of targets per regulator

In the stacked bars, we saw that regulatory out degree had very unequal representation
of plasticity groups, particularly an over-representation of single-species plasticiy connections
Here we want to compare regulators and their regulons from each plasticity group
by visualizing first the regulator expression across environments
and then their regulon (or at least the most common plasticity pattern)

First making a dataframe of each regulator in each environment and the number of connections it has
```{r tfdf}
# first how many regulators from each environment are from each plasticity group?
# and how many connections do each of them have?
tfdf <- map(ExperimentNames, \(.e) {
  expr_genes <- filter(finaldf, experiment == .e) |> 
    dplyr::select(gene_name) |> pull() |> 
    intersect(colnames(regmats$YPD))
  tibble(gene_name = expr_genes,
         out_degree_e = rowSums(regmats[[.e]][expr_genes, expr_genes])) |> 
    right_join(filter(finaldf, experiment == .e),
               by = "gene_name")
}) |> purrr::reduce(.f = bind_rows)
tfdf$out_degree_e[is.na(tfdf$out_degree_e)] <- 0
tfdf <- tfdf |> filter(out_degree_e != 0)
tfdf$plasticity <- map2(tfdf$cer, tfdf$par, getPlasticityGroup) |> unlist()
```


### Number of regulators and targets in each environmental network

```{r numbers}
# getting numbers for figure
# nTargets
for (e in ExperimentNames) {
  cat(e, "nTargets:", sum(colSums(regmats[[e]]) != 0), "\n")
}
# nRegulators
tfdf |> group_by(experiment) |> summarise(nRegulators = n())
```

### Waffle plots
Note: waffle package uses plyr, which superseeds dplyr::summarise, causing errors if we don't unload it once the waffle plots are done

```{r waffle}
library(waffle)

plotdf <- tfdf |> 
  dplyr::count(plasticity, experiment, out_degree_e, gene_name) |> 
  arrange(experiment, desc(out_degree_e))
  #arrange(experiment, plasticity)

# manually setting x and y of out degree labels
# cause waffle plot doesn't currently allow labels
# NOTE: Flip=TRUE needs to be set on geom_waffle for this to work
nrows <- 5
ncols <- max(table(plotdf$experiment))/nrows
plotdf$x <- ((c(1:nrow(plotdf)) - rank(plotdf$experiment, ties.method = "min")) %% nrows) + 1
plotdf$y <- map(as.numeric(table(plotdf$experiment)), \(x) {
                   full_vec <- sapply(c(1:ncols), \(i) {rep(i, times = nrows)}) |> c()
                   return(full_vec[c(1:x)])
                 }) |> unlist()
p <- ggplot(plotdf, aes(fill = plasticity, values = n)) +
  geom_waffle(n_rows = nrows,
    size = 1, 
    colour = "white",
    flip = TRUE) +
  geom_text(aes(x = x, y = y, label = out_degree_e), color = "white") +
  scale_fill_discrete(
    type = colordf[colordf$scheme == "plasticity",]$type,
    limits = colordf[colordf$scheme == "plasticity",]$limits) +
  facet_wrap(~factor(experiment, levels = ExperimentNames, labels = LongExperimentNames),
             nrow = 1) +
  theme_classic() +
  theme(legend.position = "none",
        axis.ticks = element_blank(),
        axis.text = element_blank()) +
  xlab("") +
  ylab("") +
  ggtitle("targets per regulator")
p
pdf("paper_figures/Yeastract/regulator_waffle.pdf",
    width = 10.5, height = 2.5)
p
dev.off()

### Also creating YPD waffle plot where genes are colored by their plasticity group in each environment
nTargets_thresh <- 21
# using a common set of genes in all environments, so regulators can be in the same
# order in each waffle plot, by desc(nTargets)
ypd_genes <- intersect(x = names(table(finaldf$gene_name))[table(finaldf$gene_name) == length(ExperimentNames)],
                       y = colnames(regmats$YPD))
ypddf <- map(ExperimentNames, \(.e) {
  tibble(gene_name = ypd_genes,
         out_degree_ypd = rowSums(regmats[["YPD"]][ypd_genes, 
                                                   ypd_genes])) |> 
    right_join(filter(finaldf, experiment == .e & gene_name %in% ypd_genes),
               by = "gene_name")
}) |> purrr::reduce(.f = bind_rows)
ypddf$out_degree_ypd[is.na(ypddf$out_degree_ypd)] <- 0
ypddf <- ypddf |> filter(out_degree_ypd > nTargets_thresh) # filtering out lowly connected regulators
table(ypddf$experiment) # should still have same regulator set across environments
ypddf$plasticity <- map2(ypddf$cer, ypddf$par, getPlasticityGroup) |> unlist()
plotdf <- ypddf |> 
  dplyr::count(plasticity, experiment, out_degree_ypd, gene_name) |> 
  arrange(experiment, desc(out_degree_ypd))
# organizing count labels
nrows <- 6
ncols <- ceiling(max(table(plotdf$experiment))/nrows)
plotdf$x <- ((c(1:nrow(plotdf)) - rank(plotdf$experiment, ties.method = "min")) %% nrows) + 1
plotdf$y <- map(as.numeric(table(plotdf$experiment)), \(x) {
  full_vec <- sapply(c(1:ncols), \(i) {
    rep(i, times = nrows)
  }) |> c()
  return(full_vec[c(1:x)])
}) |> unlist()

#arrange(experiment, plasticity)
p <- ggplot(plotdf, aes(fill = plasticity, values = n)) +
  geom_waffle(n_rows = nrows,
              size = 1, 
              colour = "white",
              flip = TRUE) +
  geom_text(aes(x = x, y = y, label = out_degree_ypd), color = "white") +
  scale_fill_discrete(
    type = colordf[colordf$scheme == "plasticity",]$type,
    limits = colordf[colordf$scheme == "plasticity",]$limits) +
  facet_wrap(~factor(experiment, levels = ExperimentNames, labels = LongExperimentNames),
             nrow = 1) +
  theme_classic() +
  theme(legend.position = "none",
        axis.ticks = element_blank(),
        axis.text = element_blank()) +
  xlab("") +
  ylab("") +
  ggtitle("targets per regulator (unstressed networks)")
p
pdf("paper_figures/Yeastract/regulator_waffle_YPD.pdf",
    width = 11.5, height = 2.5)  
p
dev.off()

# unloading waffle
detach("package:waffle", unload=TRUE)
```
## Example regulator-regulon expression

### Example I: PHO4 regulon conserved in LowPi, diverged in HAP4

```{r plotHomeAndAltEnvironments}
plotHomeAndAltEnvironments <- function(.tf_common, 
                                       .reg_idxs = NULL,
                                       .e, .e2, .regulator_plotlims = NULL,
                                       .regulon_plotlims = NULL,
                                       .include_hybrid = TRUE,
                                       .regulator_normalization = "log2",
                                       .regulon_normalization = "scale") {
  e_name <- LongExperimentNames[ExperimentNames == .e]
  e2_name <- LongExperimentNames[ExperimentNames == .e2]
  ### Home environment
  # regulator
  reg_idx <- yeastract_lookup |> filter(common == .tf_common) |>
    dplyr::select(all_of("systematic")) |> pull()
  p_regulator <- annotate_figure(plotGenes(.gene_idxs = reg_idx, .quartet = .include_hybrid,
                                           .normalization = .regulator_normalization,
                                           .plotlims = .regulator_plotlims,
                                           .experiment_name = .e), top = .tf_common) 
  # regulon
  if (is.null(.reg_idxs)) {
    regulon_idxs <- filter(regdf, environment == .e & regulator == reg_idx) |> 
    dplyr::select(target) |> pull() |> 
    intersect(y = finaldf[finaldf$experiment == .e,]$gene_name)
  } else {
    regulon_idxs <- intersect(x = .reg_idxs,
                              y = finaldf[finaldf$experiment == .e,]$gene_name)
  }
  p_regulon <- annotate_figure(plotGenes(.gene_idxs = regulon_idxs, 
                                         .normalization = .regulon_normalization,
                                         .plotlims = .regulon_plotlims,
                                         .experiment_name = .e,
                                         .quartet = .include_hybrid),
                               top = paste0(.tf_common, " regulon (", length(regulon_idxs), " genes)"))
  p_home <- annotate_figure(ggarrange(p_regulator, p_regulon, nrow = 2, ncol = 1),
                            top = e_name)
  ### Alt environment
  # regulator
  p_regulator <- annotate_figure(plotGenes(.gene_idxs = reg_idx, .quartet = .include_hybrid,
                                           .normalization = .regulator_normalization,
                                           .plotlims = .regulator_plotlims,
                                           .experiment_name = .e2), top = .tf_common) 
  # regulon
  p_regulon <- annotate_figure(plotGenes(.gene_idxs = regulon_idxs, 
                                         .normalization = .regulon_normalization,
                                         .plotlims = .regulon_plotlims,
                                         .experiment_name = .e2,
                                         .quartet = .include_hybrid),
                               top = paste0(.tf_common, " regulon (", length(regulon_idxs), " genes)"))
  p_divEnv <- annotate_figure(ggarrange(p_regulator, p_regulon, nrow = 2, ncol = 1),
                              top = e2_name)
  return(ggarrange(p_home, p_divEnv, nrow = 1, ncol = 2))
}
```

```{r pho4, warning=FALSE, message=FALSE}
regulator_ylims <- c(4.5, 10)
regulon_ylims <- NULL
p <- plotHomeAndAltEnvironments(.tf_common = "PHO4", .e = "LowPi", 
                                .e2 = "HAP4", .regulator_plotlims = regulator_ylims,
                                .regulon_plotlims = regulon_ylims)
p
pdf(paste0("paper_figures/Yeastract/PHO4.pdf"),
    width = 6, height = 6)
p
dev.off()
```

### Example II: GCN4 conserved in HU Shock, diverged in Heat and LowPi

```{r gcn4, warning=FALSE, message=FALSE}
p <- plotHomeAndAltEnvironments(.tf_common = "GCN4", .e = "CC", 
                                .e2 = "Heat", .regulator_plotlims = c(8, 12))
p
pdf(paste0("paper_figures/Yeastract/GCN4.pdf"),
    width = 6, height = 6)
p
dev.off()
```

### Example III: HSF1 regulon conserved in Heat

```{r hsf1, warning=FALSE, message=FALSE}
p <- plotHomeAndAltEnvironments(.tf_common = "HSF1", .e = "Heat", 
                                .e2 = "LowPi", .regulator_plotlims = c(5, 8))
p
pdf(paste0("paper_figures/Yeastract/HSF1.pdf"),
    width = 6, height = 6)
p
dev.off()
```
RPN4 appears to have a lower YPD expression in Scer that is based in cis. Coupled to stronger Diauxic Shift-induced activation in Scer that is based in trans. Certain regulon members reverse plasticity in both directions.

```{r rpn4}
test_tf <- "RPN4"
test_cer <- 1
test_par <- 2
test_home <- "HAP4"
test_alt <- "LowN"
test_reversal_idxs <- finaldf |> filter(experiment == test_home & 
                                          cer == test_cer & par == test_par) |> 
  dplyr::select(gene_name) |> pull()
test_idxs <- regdf |> filter(regulator_common == test_tf & target %in% test_reversal_idxs) |> 
  dplyr::select(target) |> pull()
plotHomeAndAltEnvironments(.tf_common = test_tf, .e = test_home, .e2 = test_alt,
                           .reg_idxs = test_idxs, .regulon_normalization = "log2")
```

### Maybe supplemental figure: reversals have similar expression variance to conserved plasticity genes

```{r variance}
# might be needed as a supplemental figure to point out that plasticity reveresals
# and conserved plasticity have the highest expression variation 
# across both species and environments
load("data_files/Cleaned_Counts.RData")
cer_counts <- counts[, sample_info$organism == "cer" &
                       sample_info$experiment == "LowN"]
par_counts <- counts[, sample_info$organism == "par" &
                       sample_info$experiment == "LowN"]
plotdf <- bind_rows(x = tibble(gene_name = rownames(cer_counts),
                               mean_expr = rowMeans(cer_counts),
                               var_expr = rowVars(cer_counts),
                               organism = "cer"),
                    y = tibble(gene_name = rownames(par_counts),
                               mean_expr = rowMeans(par_counts),
                               var_expr = rowVars(par_counts),
                               organism = "par")) |> 
  pivot_wider(id_cols = "gene_name",
              names_from = "organism",
              values_from = c("mean_expr", "var_expr")) |> 
  right_join(finaldf, by = c("gene_name"))
plotdf$plasticityCategory <- map2(plotdf$cer, plotdf$par, getPlasticityGroup) |> unlist()
# Var/Mean by plasticity category, Scer
p1 <- ggplot(plotdf, aes(x = log2(mean_expr_cer))) + 
  geom_density(aes(fill = plasticityCategory), alpha = 0.5) +
  theme_classic() +
  scale_fill_discrete(limits = colordf[colordf$scheme == "plasticity",]$limits,
                      type = colordf[colordf$scheme == "plasticity",]$type) +
  xlab("Mean Expression (log2)")
p2 <- ggplot(plotdf, aes(x = log2(var_expr_cer))) + 
  geom_density(aes(fill = plasticityCategory), alpha = 0.5) +
  theme_classic() +
  scale_fill_discrete(limits = colordf[colordf$scheme == "plasticity",]$limits,
                      type = colordf[colordf$scheme == "plasticity",]$type) +
  xlab("Var Expression (log2)")
annotate_figure(ggarrange(p1, p2, common.legend = TRUE,
                          legend = "right", ncol = 1, nrow = 2), top = "Scer")
# Var/Mean by plasticity category, Spar
p1 <- ggplot(plotdf, aes(x = log2(mean_expr_par))) + 
  geom_density(aes(fill = plasticityCategory), alpha = 0.5) +
  theme_classic() +
  scale_fill_discrete(limits = colordf[colordf$scheme == "plasticity",]$limits,
                      type = colordf[colordf$scheme == "plasticity",]$type) +
  xlab("Mean Expression (log2)")
p2 <- ggplot(plotdf, aes(x = log2(var_expr_par))) + 
  geom_density(aes(fill = plasticityCategory), alpha = 0.5) +
  theme_classic()  +
  scale_fill_discrete(limits = colordf[colordf$scheme == "plasticity",]$limits,
                      type = colordf[colordf$scheme == "plasticity",]$type) +
  xlab("Var Expression (log2)")
annotate_figure(ggarrange(p1, p2, common.legend = TRUE,
                          legend = "right", ncol = 1, nrow = 2), top = "Spar")
```

## Regulon plasticity divergence across environments

In the three examples above, we see that the regulons have strongly conserved plasticity in their home environment and tend to have more divergent plasticity in alternative environments. How common is this?

### Fisher exact tests of regulatory interactions being enriched among conserved plasticity target genes

```{r fisher-exact}
# each row is one regulatory interaction between one target and one regulator
plotdf <- purrr::map2(names(regmats), regmats, \(nm, mat) {
  taget_names_diverged <- finaldf$gene_name[finaldf$experiment == nm & finaldf$plasticity == "diverged"]
  taget_names_conserved <- finaldf$gene_name[finaldf$experiment == nm & finaldf$plasticity == "conserved"]
  bind_rows(x = tibble(experiment = nm,
                       plasticity = "conserved",
                       nGenes = length(taget_names_conserved),
                       nInteractions = sum(mat[,colnames(mat) %in% taget_names_conserved])),
            y = tibble(experiment = nm,
                       plasticity = "diverged",
                       nGenes = length(taget_names_diverged),
                       nInteractions = sum(mat[,colnames(mat) %in% taget_names_diverged])))
}) |> purrr::reduce(.f = bind_rows)
plotlist <- vector(mode = "list", length = length(ExperimentNames))
names(plotlist) <- ExperimentNames
for (e in ExperimentNames) {
  plotdf_e <- plotdf |> filter(experiment == e)
  fisher_mat <- plotdf_e |> 
    dplyr::select(nGenes, nInteractions) |> as.matrix()
  rownames(fisher_mat) <- plotdf_e$plasticity
  fisher_test <- fisher_mat |> fisher.test()
  fisher_odds <- fisher_test$estimate
  fisher_pvalue <- fisher_test$p.value
  knitr::kable(fisher_mat, row.names = TRUE,
               caption = paste0("Experiment = ", e, 
                                "Fisher odds ratio = ", fisher_odds, 
                                "p-value = ", fisher_pvalue)) |> print()
  plotdf_e <- pivot_longer(plotdf_e, cols = c("nGenes", "nInteractions"))
  plotlist[[e]] <- ggplot(plotdf_e, aes(x = name, y = plasticity)) +
    geom_label(aes(label = value), size = 5) +
    geom_vline(xintercept = 1.5) +
    geom_hline(yintercept = 1.5) +
    theme_classic() +
    xlab("") +
    ggtitle(paste0(LongExperimentNames[ExperimentNames == e], 
                   "\nFisher odds ratio = ", round(fisher_odds, digits = 2), 
                   "\np-value = ", formatC(fisher_pvalue))) +
    theme(plot.title = element_text(size = 8))
}
```

2 x 2 fisher matrices for paper figure:

```{r fisher-plots}
ggarrange(plotlist = plotlist, nrow = 2, ncol = 3)
pdf(file = paste0("paper_figures/Yeastract/fisher.pdf"),
      width = 7, height = 4)
ggarrange(plotlist = plotlist, nrow = 2, ncol = 3)
dev.off()
```

### Histograms of how conserved genes tend to have more regulators in each environment

```{r n-regulators-per-target-histogram}
plotdf <- purrr::map2(names(regmats), regmats, \(nm, mat) {
  tibble(target_name = colnames(mat),
         experiment = nm,
         nRegulators = colSums(mat))
}) |> purrr::reduce(.f = bind_rows) |> 
  right_join(y = finaldf, by = c("target_name"="gene_name",
                                 "experiment")) |> 
  drop_na()
plot_table <- plotdf |> filter(nRegulators > 0) |> group_by(experiment, nRegulators) |> 
  summarise(n = n())
# plotdf_downsampled <- plotdf |> group_by(experiment, plasticity) |> 
#   slice_sample(n = min(plot_table$n))
p <- ggplot(filter(plotdf, nRegulators > 0), aes(x = nRegulators)) +
  # geom_density(aes(fill = plasticity, y = after_stat(count)), 
  #              alpha = 0.25, bw = 1) +
  geom_histogram(aes(fill = plasticity), stat = "count", position = "dodge") +
  facet_wrap(~factor(experiment,
                     levels = ExperimentNames,
                     labels = c("Diauxic Shift   ", "HU Shock", "Low N", "Low Pi", "Heat", "Cold")), 
                     nrow = 1, ncol = 6) +
  theme_classic() +
  theme(legend.position = "none") +
  xlab("number of regulators per target gene") +
  ylab("number of target genes") +
  geom_text(data = filter(plot_table, n <= 5), 
            aes(label = n, x = nRegulators, y = 30),
            size = 2, position = "dodge") +
  scale_x_continuous(breaks = c(1:7)) +
  scale_fill_discrete(limits = c("conserved", "diverged"),
                      type = c(colordf$type[colordf$limits == "conserved level and plasticity"],
                             colordf$type[colordf$limits == "diverged level and plasticity"]))
p
```

pdf for figure:
```{r histogram-each-environment-pdf}
pdf("paper_figures/Yeastract/histogram.pdf",
    width = 7, height = 2)
p
dev.off()
```


```{r histogram-all-environments}
p <- ggplot(filter(plotdf, nRegulators > 0), 
            aes(x = nRegulators)) +
  geom_histogram(aes(fill = plasticity), 
                 position = position_dodge(width = 0.5), stat="count") +
  theme_classic() +
  geom_label(stat = "count", aes(label=..count.., group = plasticity), 
             position = position_dodge(width = 0.5), label.size = NA,
             alpha = 0, vjust = -0.25) +
  xlab("number of regulators per target gene") +
  scale_x_continuous(breaks = c(0:7)) +
  scale_fill_discrete(limits = c("conserved", "diverged"),
                      type = c(colordf$type[colordf$limits == "conserved level and plasticity"],
                               colordf$type[colordf$limits == "diverged level and plasticity"])) +
  ylim(c(0, 1560))
p
pdf("paper_figures/Yeastract/histogram_allEnv.pdf",
    width = 3, height = 2)
p + theme(legend.position = "none")
dev.off()
```

### Fraction of regulator targets with conserved vs diverged plasticity

```{r pct-div-plast-per-TF-table}
plotdf <- regdf |> filter(environment != "YPD") |> 
  left_join(dplyr::select(finaldf, gene_name, experiment, plasticity),
                             by = c("target"="gene_name",
                                    "environment"="experiment")) |> 
  group_by(regulator_common, regulator) |> 
  summarise(nTargets = n(),
            nDivPlast = sum(plasticity == "diverged", na.rm = TRUE)) |> 
  filter(nTargets >= 20) |> 
  mutate(pctDivPlast = nDivPlast/nTargets) 
plotdf |> 
  arrange(desc(pctDivPlast))
div30regulators <- plotdf |> filter(pctDivPlast > 0.3) |> 
  dplyr::select(regulator) |> pull()
```

```{r pct-div-plast-per-TF-plot}
p <- ggplot(plotdf, aes(x = nTargets, y = pctDivPlast)) + 
  geom_point(alpha = 0.5, aes(color = pctDivPlast > 0.3)) +
  geom_text(data = filter(plotdf, pctDivPlast > 0.3),
            aes(label = regulator_common, color = pctDivPlast > 0.3), 
            check_overlap = FALSE, position = position_jitter(width = 0, height = 0.03,
                                                              seed = 4),
            size = 3) +
  geom_text(data = filter(plotdf, pctDivPlast <= 0.3),
            aes(label = regulator_common, color = pctDivPlast > 0.3), 
            check_overlap = TRUE, nudge_x = 0, nudge_y = 0.02, size = 3.5) +
  geom_hline(yintercept = 0.3, alpha = 0.5) +
  scale_color_manual(values = c("TRUE"="red", "FALSE"="black")) +
  theme_classic() +
  theme(legend.position = "none") +
  ylab("percent") +
  ggtitle("percent of regulatory targets with diverged plasticity")
p
```

Any cis-diverging regulators with more than 30% diverged targets?

```{r cis-regulators}
plotdf <- dplyr::select(finaldf, gene_name, experiment,
                              cor_scer, cor_spar, cor_hybrid,
                              effect_size_allele, padj_allele) |> 
  filter(gene_name %in% div30regulators) |> 
  group_by(gene_name) |> 
  summarise(nEnvLevDiv= sum(padj_allele < 0.05),
            levDir = mean(effect_size_allele))
```

```{r PDR3}
plotEnvironments("YBL005W", .quartet = TRUE)
```

```{r RPN4}
plotEnvironments("YDL020C", .quartet = TRUE)
```



pdf for figure:
```{r pct-div-plast-per-TF-pdf}
pdf("paper_figures/Yeastract/pctDivergedPlasticityPerTF.pdf",
    width = 12, height = 3)
p
dev.off()
```




## Regulons have a higher percentage of conserved plasticity genes in their home environment vs alternative environments

For each large regulon (top 5 in each environment, 2 from cold), we will look at the fraction of genes with conserved plasticity in each environment.

```{r regulon-pct-conserved-plasticity}
plotdf <- tfdf |> group_by(experiment) |> 
  mutate(rank = rank(-out_degree_e, ties.method = "first")) |> 
  ungroup() |> filter(rank <= 5) |> 
  dplyr::rename(c("home"="experiment", "tf"="gene_name"))
# adding fraction conserved plasticity in each environment
plotdf_pcts <- purrr::map2(plotdf$tf, plotdf$home, \(tf, h) {
  home_mat <- regmats[[h]]
  h_idxs <- finaldf |> filter(experiment == h) |> 
    dplyr::select(gene_name) |> pull()
  regulon_idxs <- intersect(colnames(home_mat)[home_mat[tf,] > 0], h_idxs)
  outdf <- purrr::map(ExperimentNames, \(e) {
    e_plast_genes <- finaldf |>
      filter(experiment == e & plasticity == "conserved") |>
      dplyr::select(gene_name) |> pull()
    e_genedf <- tibble(tf = tf,
                       home = h,
                       experiment = e,
                       n = length(regulon_idxs),
                       nConsPlast = sum(regulon_idxs %in% e_plast_genes))
    return(e_genedf)
  }) |> purrr::reduce(.f = bind_rows)
  return(outdf)
}) |> purrr::reduce(.f = bind_rows)
```

plotting:

```{r regulon-pct-conserved-plasticity-plotting}
gcn4 <- "YEL009C"
hsf1 <- "YGL073W"
pho4 <- "YFR034C"
p <- ggplot() +
  geom_histogram(data = filter(plotdf_pcts, home != experiment),
                 aes(x = nConsPlast/n), fill = "grey") +
  geom_vline(data = filter(plotdf_pcts, home != experiment),
             aes(xintercept = mean(nConsPlast/n)), color = "black", linetype = "dashed") +
  geom_vline(data = filter(plotdf_pcts, home == experiment),
             aes(xintercept = nConsPlast/n), color = "red") +
  # annotating example tfs
  # GCN4
  geom_vline(data = filter(plotdf_pcts, home == experiment &
                             home == "CC" & tf == gcn4), aes(xintercept = nConsPlast/n), color = "blue") +
  # HSF1
  geom_vline(data = filter(plotdf_pcts, home == experiment &
                             home == "Heat" & tf == hsf1), aes(xintercept = nConsPlast/n), color = "blue") +
  # PHO4
  geom_vline(data = filter(plotdf_pcts, home == experiment &
                             home == "LowPi" & tf == pho4), aes(xintercept = nConsPlast/n), color = "blue") +
  facet_wrap(~factor(home, 
                     levels = ExperimentNames, 
                     labels = LongExperimentNames),
             nrow = 1) +
  theme_classic() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  scale_y_continuous(breaks = c(1:10)) + 
  xlab("fraction of regulon with conserved plasticity")
p
```

Outputting to pdf:

```{r home-density-pdf}
pdf("paper_figures/Yeastract/histogram_regulons.pdf",
    width = 9, height = 2)
p
dev.off()
```

## Supplemental Table 3: regulators and their regulons in each environment

```{r supp3-regulator-regulon}
plotdf <- tfdf |> dplyr::rename(c("home"="experiment", "tf"="gene_name"))
# dataframe of fraction conserved plasticity in each environment (same starting code as from regulon-pct-conserved-plasticity)
consplastdf <- purrr::map2(plotdf$tf, plotdf$home, \(tf, h) {
  home_mat <- regmats[[h]]
  h_idxs <- finaldf |> filter(experiment == h) |> 
    dplyr::select(gene_name) |> pull()
  regulon_idxs <- intersect(colnames(home_mat)[home_mat[tf,] > 0], h_idxs)
  outdf <- purrr::map(ExperimentNames, \(e) {
    e_plast_genes <- finaldf |>
      filter(experiment == e & plasticity == "conserved") |>
      dplyr::select(gene_name) |> pull()
    e_genedf <- tibble(tf = tf,
                       home = h,
                       experiment = e,
                       n = length(regulon_idxs),
                       nConsPlast = sum(regulon_idxs %in% e_plast_genes))
    return(e_genedf)
  }) |> purrr::reduce(.f = bind_rows)
  return(outdf)
}) |> purrr::reduce(.f = bind_rows) |> 
  mutate(pct_cons_plast = nConsPlast/n) |> 
  dplyr::select(tf, home, experiment, pct_cons_plast) |> 
  pivot_wider(id_cols = c("tf", "home"),
              names_from = "experiment",
              values_from = "pct_cons_plast",
              names_prefix = "percent_conserved_plasticity_")
suppTable3 <- tfdf |> dplyr::rename(c("home"="experiment", "tf"="gene_name", 
                               "regulon_size"="out_degree_e", 
                               "log2foldChange_parents"="effect_size_species",
                               "log2foldChange_hybrid"="effect_size_allele",
                               "padj_parents"="padj_species",
                               "padj_hybrid"="padj_allele")) |> 
  left_join(y = dplyr::select(yeastract_lookup, 
                       common, systematic), 
            by = c("tf"="systematic")) |> 
  dplyr::select(home, tf, common, regulon_size,
         log2foldChange_parents,
         log2foldChange_hybrid,
         padj_parents, padj_hybrid, plasticity, 
         cor_parents, cor_hybrid, cor_scer, cor_spar) |> 
  left_join(y = consplastdf, by = c("tf", "home"))
```

Writing supp table to csv:

```{r supp3-csv}
write_csv(suppTable3, file = "data_files/Supplementary_Table3.csv",
          col_names = TRUE)
```

Archive: Do any TFs Granger-cause a difference in their targets' expression?

```{r testGrangerCause}
testGrangerCause <- function(.tf, .target, .environment, .allele, .parents_or_hybrid) {
  if (.parents_or_hybrid == "hybrid") {
    cts_mat <- collapsed_allele[[.allele]]
    info_df <- info_allele
  }
  if (.parents_or_hybrid == "parents") {
    cts_mat <- collapsed[[.allele]]
    info_df <- info
  }
  if (!(.tf %in% rownames(cts_mat)) | !(.target %in% rownames(cts_mat))) {
    return(NA)
  }
  if (.tf == .target) { # no self-feedback allowed
    return(NA)
  }
  test_df <- tibble("regulator" = cts_mat[.tf,
                                          info_df$experiment == .environment],
                    "target" = cts_mat[.target,
                                       info_df$experiment == .environment],
                    "timepoint" = info_df$time_point_num[info_df$experiment == .environment]) |> 
    arrange(timepoint)
  if (var(test_df$regulator) == 0 | var(test_df$target) == 0) { # grangertest returns something other than NA for these, we want it to return NA
    return(NA)
  }
  test <- tryCatch(expr = {grangertest(target ~ regulator, order = 2, data = test_df)},
                   error = function(e) {return(NA)})
  if (!is.na(test)[1]) {
    return(as.numeric(test$`Pr(>F)`[2]))
  }
  return(test)
}
# # tests for testGrangerCause
# test_regulator_idx <- regdf |> filter(regulator_common == "RPN4") |> dplyr::select(regulator) |> pull() |> unique()
# test_target <- regdf |> filter(regulator_common == "RPN4" & 
#                                  !(environment %in% c("YPD"))) |> dplyr::select(target, environment) |> 
#   slice_sample(n = 1)
# testGrangerCause(.tf = test_regulator_idx, .target = test_target$target,
#                  .environment = test_target$environment, .allele = "cer",
#                  .parents_or_hybrid = "parents")
```


