---
title: "Level and plasticity divergence patterns across all 6 environments"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r package-loading}
sapply(c("stringr", "RColorBrewer", "circlize", "ggridges",
         "ComplexHeatmap"), require, character.only=TRUE)
source("functions_for_figure_scripts.R")
load("data_files/FinalDataframe3Disp.RData")
load("data_files/Cleaned_Counts.RData")
load("data_files/Cleaned_Counts_Allele.RData")
load("data_files/CorrelationClustering3Disp.RData")
load(file = "data_files/GO_Slim.RData")
```

Plotting a number-line of sampling timepoints proportional to time ellapsed (min) for the Experimental Design figure

```{r plotSamplingTimepoints}
# following bgoldst suggestion on stackoverflow: https://stackoverflow.com/questions/32051890/r-plotting-points-with-labels-on-a-single-horizontal-numberline
plotSamplingTimepoints <- function(.timepoints, .labels = FALSE,
                                   .label_height = 20) {
  max_tp <- max(.timepoints)
  min_tp <- min(.timepoints)
  xlim <- c(min_tp, max_tp)
  ylim <- c(min_tp, max_tp)
  px <- .timepoints
  py <- rep(0, times = length(.timepoints))
  lx.buf <- 5
  lx <- seq(xlim[1] + lx.buf,
            xlim[2] - lx.buf, 
            len = length(px))
  ly <- .label_height
  xlim <- c(xlim[1] - 10,
            xlim[2] + 30)
  ## create basic plot outline
  par(xaxs = 'i', yaxs = 'i', mar = c(5, 1, 1, 1))# mar = c(b, l, t, r)
  plot(NA, xlim = xlim, ylim = ylim, axes = FALSE, ann = FALSE)
  #axis(1)
  
  ## plot elements
  points(px, py, pch = 16, xpd = NA)
  if (.labels) {
    segments(px, py, lx, ly)
    text(lx, ly, px, pos = 3)
  }
}
# tests for plotSamplingTimepoints
test_tps <- info[info$experiment == "LowPi",]$time_point_num |> unique()
plotSamplingTimepoints(.timepoints = test_tps)
plotSamplingTimepoints(.timepoints = test_tps, .labels = TRUE)
plotSamplingTimepoints(.timepoints = test_tps, .label_height = 100, .labels = TRUE)

label_lookup <- tibble(experiment = ExperimentNames,
                       label_height = c(150, 10, 40, 20, 5, 5))
```

```{r numberlines}
# generating plots
for (e in ExperimentNames) {
  e_tps <- info[info$experiment == e,]$time_point_num |> unique()
  labh <- label_lookup[label_lookup$experiment == e,]$label_height |> as.numeric()
  pdf(file = paste0("paper_figures/ExperimentOverview/timepoints_",
                    e, ".pdf"), width = 5, height = 2)
  print(plotSamplingTimepoints(.timepoints = e_tps, .label_height = labh))
  dev.off()
}
```
Visualizing cluster average expression for all environments:

```{r brewer-pallets}
display.brewer.all()
```
Color palettes we'll use for the 6 environments: 

```{r palettedf}
palettedf <- tibble(experiment = ExperimentNames,
                    palette = c("Greys", "YlGn", "Greens", "Purples", "YlOrBr", "BuPu"),
                    long_name = LongExperimentNames)
```

Wrapping cluster plots into a function so we don't have to repeatedly make the same plot

```{r plotClusterPatternByExperiment}
plotClusterPatternByExperiment <- function(.df, .experiment, .title = NULL) {
  plotdf <- summarise(group_by(.df, time_point_num, label),
                      mean_expr = mean(expr, na.rm = TRUE))
  plotdf$label <- as.factor(plotdf$label)
  color_plt <- palettedf |> filter(experiment == .experiment) |> dplyr::select(palette) |> pull()
  if (is.null(.title)) {
    .title <- palettedf |> filter(experiment == .experiment) |> dplyr::select(long_name) |> pull()
  }
  ggplot(plotdf, aes(x = time_point_num, y = log2(mean_expr + 1))) +
    geom_line(aes(group = label, color = label), linewidth = 4) +
    geom_point(color = "black", alpha = 0.4) +
    xlab("timepoint (min)") +
    ylab("expression (log2)") +
    scale_color_brewer(palette = color_plt, name = "cluster",
                       direction = -1) +
    theme_classic() +
    theme(legend.position = "right") +
    ggtitle(.title)
}
```

```{r cluster-lineplots}
p_blank <- ggplot() + theme_void()
pdf("paper_figures/ExperimentOverview/cluster_ref.pdf",
    width = 8, height = 8)
ggarrange(plotClusterPatternByExperiment(clusterdf_list$HAP4_2$df, .experiment = "HAP4"),
          p_blank,
          plotClusterPatternByExperiment(clusterdf_list$CC_2$df, .experiment = "CC"),
          plotClusterPatternByExperiment(clusterdf_list$LowN_2$df, .experiment = "LowN"),
          p_blank,
          plotClusterPatternByExperiment(clusterdf_list$LowPi_2$df, .experiment = "LowPi"),
          plotClusterPatternByExperiment(clusterdf_list$Heat_2$df, .experiment = "Heat"),
          p_blank,
          plotClusterPatternByExperiment(clusterdf_list$Cold_2$df, .experiment = "Cold"),
          nrow = 3, ncol = 3, common.legend = FALSE, widths = c(2, 0.5, 2))
dev.off()
```
## Similar proportions of genes are diverging in each environment

Stacked bars of 4 divergence categories in each environment:

```{r proportions-stacked-bars}
plotdf <- finaldf |> dplyr::select(gene_name, experiment, group4) |> 
  pivot_wider(id_cols = c("gene_name"), names_from = "experiment",
              values_from = "group4") |> 
  pivot_longer(cols = ExperimentNames, names_to = "experiment",
               values_to = "group5") # to add NA values for low expression
plotdf$group5 <- if_else(is.na(plotdf$group5), true = "lowly expressed",
                         false = plotdf$group5) |> 
  factor(levels = c("conserved level and plasticity",
                    "conserved level, diverged plasticity",
                    "diverged level, conserved plasticity",
                    "diverged level and plasticity",
                    "lowly expressed"))
plotdf$experiment <- factor(plotdf$experiment, levels = ExperimentNames)
p <- ggplot(plotdf, aes(x = experiment)) + 
  geom_bar_pattern(aes(fill = group5, pattern = group5), pattern_fill = "black") +
  scale_fill_discrete(type = colordf[colordf$scheme == "group4",]$type,
                      limits = colordf[colordf$scheme == "group4",]$limits) +
  scale_pattern_discrete(choices = colordf[colordf$scheme == "group4",]$pattern,
                         limits = colordf[colordf$scheme == "group4",]$limits) +
  scale_x_discrete(limits = ExperimentNames, labels = LongExperimentNames) +
  theme_classic() +
  theme(legend.title = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        legend.position = "bottom") +
  ylab("number of genes") +
  xlab("") +
  guides(fill = guide_legend(nrow = 3))
pdf("paper_figures/EnvironmentalPatterns/stacked_bars.pdf",
    width = 5, height = 6)
p
dev.off()
```

Ridgeline plot of log2 fold changes in each environment:

```{r proportions-ridgeline}
plotdf <- filter(finaldf, level == "diverged")
filter(finaldf, plasticity == "conserved" & level == "conserved") |> group_by(experiment) |> 
  summarise(n = n())
# accompanying percentages of upScer and upSpar genes
pctdf <- plotdf |> group_by(experiment) |> 
  summarise(pct_upcer = round(sum(effect_size_species > 0)/length(effect_size_species), digits = 2)*100,
            pct_uppar = round(sum(effect_size_species < 0)/length(effect_size_species), digits = 2)*100,
            n = length(effect_size_species)) |> 
  pivot_longer(cols = c("pct_upcer", "pct_uppar"),
               names_to = "direction", values_to = "pct") |> 
  mutate(x_pos = if_else(direction == "pct_upcer",
                         true = 5, false = -5))
pctdf$y_pos <- sapply(pctdf$experiment, \(e) which(ExperimentNames == e) + 0.5)
p <- ggplot(data = plotdf, aes(x = effect_size_species, y = experiment, fill = experiment)) +
  geom_density_ridges(data = plotdf) +
  geom_text(data = pctdf, aes(x = x_pos, y = y_pos, label = paste0(pct, "%"))) +
  theme_classic() +
  theme(legend.position = "none") +
  ylab("") +
  xlab("log2 fold change") +
  scale_fill_discrete(limits = colordf[colordf$scheme == "experiment",]$limits,
                      type = colordf[colordf$scheme == "experiment",]$type) +
  scale_y_discrete(limits = ExperimentNames, labels = LongExperimentNames)
p
pdf("paper_figures/EnvironmentalPatterns/ridgeline.pdf",
    width = 4, height = 3)
p
dev.off()
```

Heatmap of species-specific and reversals of expression plasticity

```{r plasticity-group-heatmap}
plotdf <- finaldf |> # filter(plasticity == "diverged") |> 
  group_by(experiment, cer, par) |> 
  summarise(n = n())
plotdf$long_experiment <- map(plotdf$experiment, \(e) {
  LongExperimentNames[which(ExperimentNames == e)]
}) |> factor(levels = LongExperimentNames)
plotdf$cer <- factor(plotdf$cer, levels = c("0", "1", "2"))
plotdf$par <- factor(plotdf$par, levels = c("2", "1", "0"))
plotdf$type <- map2(plotdf$cer, plotdf$par, \(x, y) {
  if (x == y) {
    if (x == 0 & y == 0) {
      return("conserved static")
    }
    return("conserved plastic")
  }
  if (x == 0) {
    return("Spar-unique")
  }
  if (y == 0) {
    return("Scer-unique")
  }
  else {
    return("reversal")
  }
}) |> unlist()
plotdf_sums <- plotdf |> group_by(long_experiment) |> 
  summarise(n_Exp = sum(n))
plotdf <- left_join(plotdf, plotdf_sums, 
                    by = "long_experiment")
plotdf$pct <- (plotdf$n/plotdf$n_Exp) |> formatC(digits = 1)
p <- ggplot(plotdf, aes(x = cer, y = par, fill = type)) +
  geom_tile() +
  geom_text(aes(label = pct,
                color = grepl(pattern = "conserved", x = type)), size = 3) +
  scale_color_discrete(limits = c(TRUE, FALSE),
                       type = c("grey40", "white")) +
  scale_fill_discrete(limits = colordf[colordf$scheme == "plasticity",]$limits,
                      type = colordf[colordf$scheme == "plasticity",]$type) +
   facet_wrap(~long_experiment) +
  theme_classic() +
  xlab("Scer plasticity cluster") +
  ylab("Spar plasticity cluster") +
  guides(fill=guide_legend(title="number of genes")) +
  theme(legend.position = "none",
        panel.spacing.y = unit(1, "cm"))
pdf("paper_figures/EnvironmentalPatterns/plasticity_counts_heatmap.pdf",
    width = 4, height = 3)
p
dev.off()
```
nGenes per experiment for paper figure:

```{r plasticity-group-heatmap-ngenes}
plotdf_sums
```

## Divergence category in one environment is a poor predictor of other environments

```{r discrete-heatmap, cache=TRUE}
plotdf <- finaldf |> dplyr::select(gene_name, experiment, group4) |> 
  # pivoting wider and longer again to add the genes that have been removed for low expression
  pivot_wider(id_cols = "gene_name", values_from = "group4", names_from = "experiment") |> 
  pivot_longer(cols = all_of(ExperimentNames),
               names_to = "experiment", values_to = "group4")
plotdf$group4[is.na(plotdf$group4)] <- "lowly expressed"
plot_mat <- plotdf |> 
  pivot_wider(id_cols = "gene_name",
              names_from = "experiment", 
              values_from = "group4")
colnames_plotmat <- plot_mat$gene_name
rownames_plotmat <- colnames(plot_mat)
plot_mat <- data.frame(plot_mat) |> t()
colnames(plot_mat) <- colnames_plotmat
rownames(plot_mat) <- rownames_plotmat
plot_mat <- plot_mat[rownames(plot_mat) != "gene_name",]
plot_mat <- plot_mat |> as.matrix()
clrs <- structure(colordf[colordf$scheme == "group4",]$type, 
                  names = colordf[colordf$scheme == "group4",]$limits)
majority_NA <- apply(plot_mat, 2, \(x) {return(sum(is.na(x)) > 2)})
sum(majority_NA)
plot_mat <- plot_mat[,!majority_NA]
# sorting cols by HAP4's order
col_order_vec <- order(factor(plot_mat["HAP4",], levels = colordf[colordf$scheme == "group4",]$limits))

# # tests for OrderGenesByGroup (moved to function script b/c we use it in multiple scripts now)
# # subset of plotmat
# orderGenesByGroup(.mat = plot_mat[, 1:100]) |> dim()

# ordering full plotmat
ordered_plot_mat <- orderGenesByGroup(.mat = plot_mat[ExperimentNames,])
rownames(ordered_plot_mat) <- LongExperimentNames
plotdf <- map(LongExperimentNames, \(e) {
  idx <- which(LongExperimentNames == e)
  return(tibble(gene_name = colnames(ordered_plot_mat),
                environment = e,
                group4 = ordered_plot_mat[idx,],
                x = c(1:ncol(ordered_plot_mat)),
                y = c(length(LongExperimentNames):1)[idx])) # descending from top
}) |> purrr::reduce(.f = bind_rows)
p <- ggplot(plotdf, aes(x = x, y = y, fill = group4)) +
  geom_tile_pattern(aes(fill = group4, pattern = group4), pattern_fill = "black",
                    pattern_density = 0.05, pattern_size = 0.05) +
  scale_fill_discrete(limits = colordf[colordf$scheme == "group4",]$limits,
                      type = colordf[colordf$scheme == "group4",]$type) +
  scale_pattern_discrete(limits = colordf[colordf$scheme == "group4",]$limits,
                         choices = colordf[colordf$scheme == "group4",]$pattern) +
  theme_void()
p
# plotting
pdf("paper_figures/EnvironmentalPatterns/discrete_heatmapLines.pdf",
    width = 14, height = 2)
p
dev.off()
```

### Level divergence is consistent across environments and mostly between strains

Genes with higher Log2 fold change in Scer: 

```{r heatmap-upScer}
plot_mat <- matrix(nrow = length(ExperimentNames),
                   ncol = length(ExperimentNames))
plot_counts <- matrix(nrow = length(ExperimentNames),
                      ncol = length(ExperimentNames))
# up in Scer
for (e_row in ExperimentNames) {
  for (e_col in ExperimentNames) {
    e_gene_idxs <- finaldf |> 
      filter(experiment == e_row & 
               level == "diverged" &
               sign(effect_size_species) == 1) |> 
      dplyr::select(gene_name) |> pull()
    avg_lfc <- finaldf |> 
      filter(experiment == e_col & 
               gene_name %in% e_gene_idxs) |> 
      dplyr::select(effect_size_species) |> 
      pull() |> mean()
    n_genes <- length(e_gene_idxs)
    plot_mat[which(ExperimentNames == e_row),
             which(ExperimentNames == e_col)] <- avg_lfc
    plot_counts[which(ExperimentNames == e_row),
                which(ExperimentNames == e_col)] <- n_genes
  }
}
colnames(plot_mat) <- LongExperimentNames
rownames(plot_mat) <- LongExperimentNames

col_fun = colorRamp2(c(-1, 0, 1), c("blue2", "white", "orange1"))
p <- Heatmap(plot_mat, col = col_fun,
        row_order = LongExperimentNames, column_order = LongExperimentNames,
        row_names_side = "left", heatmap_legend_param = list(title = ""),
        cell_fun = function(j, i, x, y, width, height, fill) {
          grid.text(sprintf("%.2f", plot_mat[i, j]), x, y, gp = gpar(fontsize = 10))
        },
        show_heatmap_legend = FALSE)
p
pdf("paper_figures/EnvironmentalPatterns/level_divergence_heatmap_up_scer.pdf",
    width = 4, height = 3)
p
dev.off()
```

Legend and number of genes for paper figure (should be all the same number of genes for each row):

```{r heatmap-upScer-ngenes}
plot_counts

lgd <- Legend(col_fun = col_fun, title = "avgerage log2 fold change",  direction = "horizontal") # need to run draw(lgd) in the console
```

Genes with higher Log2 fold change in Spar: 

```{r heatmap-upPar}
plot_mat <- matrix(nrow = length(ExperimentNames),
                   ncol = length(ExperimentNames))
plot_counts <- matrix(nrow = length(ExperimentNames),
                      ncol = length(ExperimentNames))
for (e_row in ExperimentNames) {
  for (e_col in ExperimentNames) {
    e_gene_idxs <- finaldf |> filter(experiment == e_row &
                                       level == "diverged" & 
                                       sign(effect_size_species) == -1) |> 
      dplyr::select(gene_name) |> pull()
    avg_lfc <- finaldf |> filter(experiment == e_col & 
                                   gene_name %in% e_gene_idxs) |> 
      dplyr::select(effect_size_species) |> pull() |> mean()
    plot_mat[which(ExperimentNames == e_row),
             which(ExperimentNames == e_col)] <- avg_lfc
    plot_counts[which(ExperimentNames == e_row),
                which(ExperimentNames == e_col)] <- length(e_gene_idxs)
  }
}
colnames(plot_mat) <- LongExperimentNames
rownames(plot_mat) <- LongExperimentNames

col_fun = colorRamp2(c(-1, 0, 1), c("blue2", "white", "orange1"))
p <- Heatmap(plot_mat, col = col_fun,
        row_order = LongExperimentNames, column_order = LongExperimentNames,
        row_names_side = "left", heatmap_legend_param = list(title = ""),
        cell_fun = function(j, i, x, y, width, height, fill) {
          grid.text(sprintf("%.2f", plot_mat[i, j]), x, y, gp = gpar(fontsize = 10))
        }, show_heatmap_legend = FALSE)
p
pdf("paper_figures/EnvironmentalPatterns/level_divergence_heatmap_up_spar.pdf",
    width = 4, height = 3)
p
dev.off()
```

Legend and number of genes for paper figure (should be all the same number of genes for each row):

```{r heatmap-upSpar-ngenes}
plot_counts

lgd <- Legend(col_fun = col_fun, title = "avgerage log2 fold change",  direction = "horizontal") # need to run draw(lgd) in the console
```

### Plasticity divergence is unique to each environment

Y axis: environment those 2-1 or 1-2 divergers were ID'd in
X axis: how their plasticity divergence looks in other environments
How do we measure plasticity divergence? Correlation of avg expr between species

```{r getplasticityCorr}
# Given gene idxs and experiment, 
# returns between-species correlation
# of average expression across timepoints
getplasticityCorr <- function(.gene_idxs, .experiment_name) {
  condition_vec <- info |> filter(experiment == .experiment_name) |> 
    dplyr::select(condition) |> pull()
  cer_vec <- collapsed$cer[.gene_idxs, condition_vec] |> 
    colMeans(na.rm = TRUE)
  par_vec <- collapsed$par[.gene_idxs, condition_vec] |> 
    colMeans(na.rm = TRUE)
  return(cor(cer_vec, par_vec))
}
# tests for getplasticity Corr
# HAP4 2-1, should be very uncorrelated in Sat Growth
gene_idxs <- finaldf |> filter(experiment == "HAP4" &
                                 cer == 2 & par == 1) |> 
  dplyr::select(gene_name) |> pull()
getplasticityCorr(gene_idxs, .experiment_name = "HAP4")
getplasticityCorr(gene_idxs, .experiment_name = "LowN")
getplasticityCorr(gene_idxs, .experiment_name = "LowPi")
getplasticityCorr(gene_idxs, .experiment_name = "CC")
getplasticityCorr(gene_idxs, .experiment_name = "Heat")
getplasticityCorr(gene_idxs, .experiment_name = "Cold")
```

Heatmap of species correlations for genes increasing in Spar and decreasing in Scer:

```{r 2-1-heatmap}
# 2-1 divergers
plot_mat <- matrix(nrow = length(ExperimentNames),
                   ncol = length(ExperimentNames))
plot_counts <- matrix(nrow = length(ExperimentNames),
                      ncol = length(ExperimentNames))
for (e_row in ExperimentNames) {
  for (e_col in ExperimentNames) {
    e_gene_idxs <- finaldf |> filter(experiment == e_row &
                                       cer == 2 & par == 1) |> 
      dplyr::select(gene_name) |> pull()
    e_cor <- getplasticityCorr(e_gene_idxs, .experiment_name = e_col)
    plot_mat[which(ExperimentNames == e_row),
             which(ExperimentNames == e_col)] <- e_cor
    plot_counts[which(ExperimentNames == e_row),
                which(ExperimentNames == e_col)] <- length(e_gene_idxs)
  }
}
colnames(plot_mat) <- LongExperimentNames
rownames(plot_mat) <- LongExperimentNames

# plotting
col_fun = colorRamp2(c(-1, 0, 1), c("red2", "white", "skyblue"))
p <- Heatmap(plot_mat, col = col_fun,
        row_order = LongExperimentNames, column_order = LongExperimentNames,
        row_names_side = "left", heatmap_legend_param = list(title = ""),
        cell_fun = function(j, i, x, y, width, height, fill) {
          grid.text(sprintf("%.2f", plot_mat[i, j]), x, y, gp = gpar(fontsize = 10))
        }, show_heatmap_legend = FALSE)
p
pdf("paper_figures/EnvironmentalPatterns/heatmap_21.pdf",
    width = 4, height = 3)
p
dev.off()
```

nGenes and legend for paper figure

```{r heatmap-2-1-ngenes}
plot_counts

lgd <- Legend(col_fun = col_fun, title = "avgerage expression correlation",  direction = "horizontal") # need to run draw(lgd) in the console
```

Heatmap of species correlations for genes decreasing in Spar and increasing in Scer:

```{r 1-2-heatmap}
# 1-2 divergers
plot_mat <- matrix(nrow = length(ExperimentNames),
                   ncol = length(ExperimentNames))
plot_counts <- matrix(nrow = length(ExperimentNames),
                      ncol = length(ExperimentNames))
for (e_row in ExperimentNames) {
  for (e_col in ExperimentNames) {
    e_gene_idxs <- finaldf |> filter(experiment == e_row &
                                       cer == 1 & par == 2) |> 
      dplyr::select(gene_name) |> pull()
    e_cor <- getplasticityCorr(e_gene_idxs, .experiment_name = e_col)
    plot_mat[which(ExperimentNames == e_row),
             which(ExperimentNames == e_col)] <- e_cor
    plot_counts[which(ExperimentNames == e_row),
                which(ExperimentNames == e_col)] <- length(e_gene_idxs)
  }
}
colnames(plot_mat) <- LongExperimentNames
rownames(plot_mat) <- LongExperimentNames

# plotting
col_fun = colorRamp2(c(-1, 0, 1), c("red2", "white", "skyblue"))
p <- Heatmap(plot_mat, col = col_fun,
        row_order = LongExperimentNames, column_order = LongExperimentNames,
        row_names_side = "left", heatmap_legend_param = list(title = ""),
        cell_fun = function(j, i, x, y, width, height, fill) {
          grid.text(sprintf("%.2f", plot_mat[i, j]), x, y, gp = gpar(fontsize = 10))
        }, show_heatmap_legend = FALSE)
p
pdf("paper_figures/EnvironmentalPatterns/heatmap_12.pdf",
    width = 4, height = 3)
p
dev.off()
```

nGenes counts for paper figure:

```{r heatmap-1-2-ngenes}
plot_counts

lgd <- Legend(col_fun = col_fun, title = "avgerage expression correlation",  direction = "horizontal") # need to run draw(lgd) in the console
```


All plasticity-divergers grouped together:

```{r getplasticityCorrAllClusters}
getplasticityCorrAllClusters <- function(.gene_idxs, .experiment_name) {
  condition_vec <- info |> filter(experiment == .experiment_name) |> 
    dplyr::select(condition) |> pull()
  
  clust_pairs <- finaldf |> filter(gene_name %in% .gene_idxs &
                                     experiment == .experiment_name) |> 
    group_by(cer, par) |> summarise(n = n()) |> ungroup()
  cors <- map2(clust_pairs$cer, clust_pairs$par, \(x, y) {
    clust_idxs <- finaldf |> filter(gene_name %in% .gene_idxs &
                                      experiment == .experiment_name &
                                      cer == x & par == y) |> 
      dplyr::select(gene_name) |> pull()
    cer_vec <- collapsed$cer[clust_idxs, condition_vec, drop = FALSE] |> 
      colMeans(na.rm = TRUE)
    par_vec <- collapsed$par[clust_idxs, condition_vec, drop = FALSE] |> 
      colMeans(na.rm = TRUE)
    return(cor(cer_vec, par_vec))
  }) |> unlist()
  return(as.numeric(sum(cors*clust_pairs$n)/sum(clust_pairs$n))) # weighted average
}

# tests for getplasticityCorrBoth
# LowPi 1-2 alone:
gene_idxs <- finaldf |> filter(experiment == "LowPi" &
                                 plasticity == "diverged" &
                                 cer == 1 & par == 2) |> 
  dplyr::select(gene_name) |> pull()
testcor12 <- getplasticityCorrAllClusters(gene_idxs, .experiment_name = "LowPi")
testcor12
getplasticityCorr(gene_idxs, .experiment_name = "LowPi") # should be same number
# LowPi 2-1 alone:
gene_idxs <- finaldf |> filter(experiment == "LowPi" &
                                 plasticity == "diverged" &
                                 cer == 2 & par == 1) |> 
  dplyr::select(gene_name) |> pull()
testcor21 <- getplasticityCorrAllClusters(gene_idxs, .experiment_name = "LowPi")
testcor21
getplasticityCorr(gene_idxs, .experiment_name = "LowPi") # should be same number
# weights
finaldf |> filter(experiment == "LowPi" &
                    plasticity == "diverged" &
                    cer != 0 & par != 0) |> 
  dplyr::select(cer, par) |> table()
# what both should be:
sum(testcor12*372, testcor21*795)/(372 + 795) # with weighted average
mean(c(testcor12, testcor21)) # without weighted average
# Both 1-2 and 2-1:
gene_idxs <- finaldf |> filter(experiment == "LowPi" &
                                 plasticity == "diverged" &
                                 cer != 0 & par != 0) |> 
  dplyr::select(gene_name) |> pull()
getplasticityCorrAllClusters(gene_idxs, .experiment_name = "LowPi") # should be same as manual calculation
# test2: HAP4/CC failed until we added drop = FALSE
gene_idxs <- finaldf |> filter(experiment == "HAP4" &
                                 plasticity == "diverged" &
                                 cer != 0 & par != 0) |> 
  dplyr::select(gene_name) |> pull()
finaldf |> filter(experiment == "CC" & gene_name %in% gene_idxs) |> 
  dplyr::select(cer, par) |> table() # b/c there's only one 1-0 gene
getplasticityCorrAllClusters(gene_idxs, .experiment_name = "CC")
```

```{r all-plasticity-cor-heatmap}
# plotting
plot_mat <- matrix(nrow = length(ExperimentNames),
                   ncol = length(ExperimentNames))
for (e_row in ExperimentNames) {
  for (e_col in ExperimentNames) {
    cat(e_row, e_col, "\n")
    e_gene_idxs <- finaldf |> filter(experiment == e_row &
                                         cer != 0 & par != 0 &
                                       plasticity == "diverged") |> 
      dplyr::select(gene_name) |> pull()
    e_cor <- getplasticityCorrAllClusters(.gene_idxs = e_gene_idxs,
                                        .experiment_name = e_col)
    plot_mat[which(ExperimentNames == e_row),
             which(ExperimentNames == e_col)] <- e_cor
  }
}
colnames(plot_mat) <- LongExperimentNames
rownames(plot_mat) <- LongExperimentNames

# plotting
col_fun = colorRamp2(c(-1, 0, 1), c("red", "white", "skyblue"))
p <- Heatmap(plot_mat, col = col_fun,
        row_order = LongExperimentNames, column_order = LongExperimentNames,
        row_names_side = "left", heatmap_legend_param = list(title = ""),
        cell_fun = function(j, i, x, y, width, height, fill) {
          grid.text(sprintf("%.2f", plot_mat[i, j]), x, y, gp = gpar(fontsize = 10))
        })
p
pdf("paper_figures/EnvironmentalPatterns/heatmap_plasticity_all_clusters.pdf",
    width = 5, height = 4)
p
dev.off()
```

### QC: getting plasticity cor after centering and scaling

```{r getplasticityCorrScaled}
# Given gene idxs and experiment, 
# returns between-species correlation
# of average expression across timepoints
getplasticityCorr <- function(.gene_idxs, .experiment_name) {
  condition_vec <- info |> filter(experiment == .experiment_name) |> 
    dplyr::select(condition) |> pull()
  cer_vec <- collapsed$cer[.gene_idxs, condition_vec] |> 
    colMeans(na.rm = TRUE)
  par_vec <- collapsed$par[.gene_idxs, condition_vec] |> 
    colMeans(na.rm = TRUE)
  return(cor(cer_vec, par_vec))
}
# tests for getplasticity Corr
# HAP4 2-1, should be very uncorrelated in Sat Growth
gene_idxs <- finaldf |> filter(experiment == "HAP4" &
                                 cer == 2 & par == 1) |> 
  dplyr::select(gene_name) |> pull()
getplasticityCorr(gene_idxs, .experiment_name = "HAP4")
getplasticityCorr(gene_idxs, .experiment_name = "LowN")
getplasticityCorr(gene_idxs, .experiment_name = "LowPi")
getplasticityCorr(gene_idxs, .experiment_name = "CC")
getplasticityCorr(gene_idxs, .experiment_name = "Heat")
getplasticityCorr(gene_idxs, .experiment_name = "Cold")
```

### Example expression profiles for figure

Genes increasing in Scer, decreasing in Spar, identified in Low Phosphate:

```{r LowPi-1-2}
gene_idxs <- finaldf |> filter(experiment == "LowPi" & cer == 1 & par == 2) |> 
  dplyr::select(gene_name) |> pull()
p <- plotExpressionProfilePair(.cts1 = collapsed$cer[gene_idxs,],
                          .cts2 = collapsed$par[gene_idxs,],
                          .info1 = info,
                          .info2 = info,,
                          .method = "line",
                          .show_points = FALSE,
                          .show_confidence_intervals = TRUE,
                          .normalization = "log2")
p
pdf("paper_figures/EnvironmentalPatterns/LowPi12.pdf",
    width = 12, height = 2)
p
dev.off()
```

Genes with higher expression in Scer, identified in the Low Nitrogen environment:

```{r LowN-upScer}
gene_idxs <- finaldf |> filter(experiment == "LowN" & level == "diverged" &
                                 sign(effect_size_species) == 1) |> 
  dplyr::select(gene_name) |> pull()
p <- plotExpressionProfilePair(.cts1 = collapsed$cer[gene_idxs,],
                          .cts2 = collapsed$par[gene_idxs,],
                          .info1 = info,
                          .info2 = info,,
                          .method = "line",
                          .show_points = FALSE,
                          .show_confidence_intervals = TRUE,
                          .normalization = "log2")
pdf("paper_figures/EnvironmentalPatterns/LowN_upcer.pdf",
    width = 12, height = 2)
p
dev.off()
```

### Following up on additional examples flagged in heatmap

```{r HAP4-1-2}
# following up on additional examples from heatmap
# Diauxic Shift 1-2, plasticity divergers have slight negative correlation Low Pi -0.2?
gene_idxs <- finaldf |> filter(experiment == "HAP4" & cer == 1 & par == 2) |> 
  dplyr::select(gene_name) |> pull()
p <- plotExpressionProfilePairBasic(.cts1 = collapsed$cer[gene_idxs,],
                          .cts2 = collapsed$par[gene_idxs,],
                          .info1 = info,
                          .info2 = info,
                          .normalization = "log2")
length(gene_idxs)
p
```

Plot for supplemental figure:

```{r HAP4-1-2-pdf}
pdf("paper_figures/Supplement/HAP4-1-2.pdf",
    width = 12, height = 2)
p
dev.off()
```

```{r LowN-1-2}
# following up on additional examples from heatmap
# Low Nitrogen 1-2, plasticity divergers have strongest negative correlation in Cold? -.97?
gene_idxs <- finaldf |> filter(experiment == "LowN" & cer == 1 & par == 2) |> 
  dplyr::select(gene_name) |> pull()
p <- plotExpressionProfilePairBasic(.cts1 = collapsed$cer[gene_idxs,],
                          .cts2 = collapsed$par[gene_idxs,],
                          .info1 = info,
                          .info2 = info,
                          .normalization = "log2")
length(gene_idxs)
p
```

Plot for supplemental figure:

```{r LowN-1-2-pdf}
pdf("paper_figures/Supplement/LowN-1-2.pdf",
    width = 12, height = 2)
p
dev.off()
```

Why does LowPi have correlation of 0.95 for these 52 genes? The average looks middling
Conclusion: it was a result of taking log2 before the average. We want to take average before log2, as log2 is just for visualization purposes not analysis
Leaving this chunk here for now in case it is helpful
```{r low-Pi95}
gene_idxs <- finaldf |> filter(experiment == "LowN" & cer == 1 & par == 2) |> 
  dplyr::select(gene_name) |> pull()
getplasticityCorr(.gene_idxs = gene_idxs, .experiment_name = "LowPi")
test_condition_vec <- info |> filter(experiment == "LowPi") |> 
    dplyr::select(condition) |> pull()
# means
test_cer_vec <- collapsed$cer[gene_idxs, test_condition_vec] |>
  colMeans(na.rm = TRUE)
test_par_vec <- collapsed$par[gene_idxs, test_condition_vec] |>
  colMeans(na.rm = TRUE)
test_cer_vec_scaled <- collapsed$cer[gene_idxs, test_condition_vec] |>
  t() |> 
  scale() |> 
  t() |> 
  colMeans(na.rm = TRUE)
test_par_vec_scaled <- collapsed$par[gene_idxs, test_condition_vec] |> 
  t() |> 
  scale() |> 
  t() |> 
  colMeans(na.rm = TRUE)
# sds
test_cer_sd <- collapsed$cer[gene_idxs, test_condition_vec] |>
  apply(MARGIN = 2, FUN = sd)
test_par_sd <- collapsed$par[gene_idxs, test_condition_vec] |>
  apply(MARGIN = 2, FUN = sd)
test_cer_sd_scaled <- collapsed$cer[gene_idxs, test_condition_vec] |>
  t() |> 
  scale() |> 
  t() |> 
  apply(MARGIN = 2, FUN = sd)
test_par_sd_scaled <- collapsed$par[gene_idxs, test_condition_vec] |>
  t() |> 
  scale() |> 
  t() |> 
  apply(MARGIN = 2, FUN = sd)
plotdf <- bind_rows(tibble(species = "cer",
                           expr = test_cer_vec,
                           scaled_expr = test_cer_vec_scaled,
                           sd = test_cer_sd,
                           scaled_sd = test_cer_sd_scaled,
                           timepoint = parse_number(test_condition_vec)),
                    tibble(species = "par",
                           expr = test_par_vec,
                           scaled_expr = test_par_vec_scaled,
                           sd = test_par_sd,
                           scaled_sd = test_par_sd_scaled,
                           timepoint = parse_number(test_condition_vec)))
#View(plotdf)
# scaled
p_scale <- ggplot(plotdf, aes(x = timepoint, y = scaled_expr)) + 
  geom_line(aes(color = species)) + 
  geom_ribbon(aes(ymax = scaled_expr + scaled_sd, 
                  ymin = scaled_expr - scaled_sd, 
                  fill = species), alpha = 0.3)
# log2
p_log <- ggplot(plotdf, aes(x = timepoint, y = log2(expr))) + 
  geom_line(aes(color = species)) +
  geom_ribbon(aes(ymax = log2(expr + sd/sqrt(length(gene_idxs))), 
                  ymin = log2(expr - sd/sqrt(length(gene_idxs))), 
                  fill = species), alpha = 0.3)
ggarrange(p_scale, p_log, nrow = 1, ncol = 2)
cor(test_cer_vec, test_par_vec) # should be 0.95
p_log <- plotExpressionProfilePairBasic(.cts1 = collapsed$cer[gene_idxs, info$experiment == "LowPi"],
                          .cts2 = collapsed$par[gene_idxs, info$experiment == "LowPi"],
                          .info1 = filter(info, experiment == "LowPi"),
                          .info2 = filter(info, experiment == "LowPi"),,
                          .normalization = "log2")
length(gene_idxs)
p_scalebasic <- plotExpressionProfilePairBasic(.cts1 = collapsed$cer[gene_idxs, info$experiment == "LowPi"],
                          .cts2 = collapsed$par[gene_idxs, info$experiment == "LowPi"],
                          .info1 = filter(info, experiment == "LowPi"),
                          .info2 = filter(info, experiment == "LowPi"),,
                          .normalization = "scale")
p_scale <- plotExpressionProfilePair(.cts1 = collapsed$cer[gene_idxs, info$experiment == "LowPi"],
                          .cts2 = collapsed$par[gene_idxs, info$experiment == "LowPi"],
                          .info1 = filter(info, experiment == "LowPi"),
                          .info2 = filter(info, experiment == "LowPi"),,
                          .normalization = "scale")
ggarrange(p_scale, p_scalebasic, nrow = 1, ncol = 2, common.legend = TRUE)
```


```{r HAP4-2-1}
# following up on additional examples from heatmap
# Diauxic Shift 2-1, plasticity divergers also have strongest negative correlation in Cold?
gene_idxs <- finaldf |> filter(experiment == "HAP4" & cer == 2 & par == 1) |> 
  dplyr::select(gene_name) |> pull()
p <- plotExpressionProfilePairBasic(.cts1 = collapsed$cer[gene_idxs,],
                                    .cts2 = collapsed$par[gene_idxs,],
                                    .info1 = info,
                                    .info2 = info,
                                    .normalization = "log2")
length(gene_idxs)
p
```

```{r HAP4-2-1-pdf}
pdf("paper_figures/Supplement/HAP4-2-1.pdf",
    width = 12, height = 2)
p
dev.off()
```

```{r CC-2-1}
# following up on additional examples from heatmap
# HUShock 2-1, plasticity divergers also have strongest negative correlation in LowPi?
gene_idxs <- finaldf |> filter(experiment == "CC" & cer == 2 & par == 1) |> 
  dplyr::select(gene_name) |> pull()
p <- plotExpressionProfilePairBasic(.cts1 = collapsed$cer[gene_idxs,],
                          .cts2 = collapsed$par[gene_idxs,],
                          .info1 = info,
                          .info2 = info,
                          .normalization = "log2")
length(gene_idxs)
p
```

```{r CC-2-1-pdf}
pdf("paper_figures/Supplement/CC-2-1.pdf",
    width = 12, height = 2)
p
dev.off()
```


# Supplemental figures

```{r getplasticityCorrEachGene}
# no averages. This is for a large heatmap
getplasticityCorrEachGene <- function(.gene_idxs, .experiment_name) {
  # filter for genes expressed high enough in this environment
  finaldf |> filter(experiment == .experiment_name &
                                         gene_name %in% .gene_idxs) |> 
    dplyr::select(gene_name, cor_parents) |> 
    right_join(y = tibble(gene_name = .gene_idxs),
               by = "gene_name") |> dplyr::select(cor_parents) |> pull()
}
# tests for getplasticityCorrEachGene
test_idxs <- finaldf |> filter(experiment == "HAP4" &
                                       plasticity == "diverged" & 
                                       cer != 0 & par != 0) |> 
      dplyr::select(gene_name) |> pull()
test_cors <- getplasticityCorrEachGene(.gene_idxs = test_idxs, .experiment_name = "HAP4")
test_plotmat_n <- ceiling(sqrt(length(test_cors)))
test_plotmat <- matrix(test_cors[1:(test_plotmat_n*test_plotmat_n)], 
                       nrow = test_plotmat_n, ncol = test_plotmat_n)
# Heat has enough lowly expressed genes that are diverging in HAP4 to result in a lower-dimension square matrix being ideal (18 vs 19)
test_cors2 <- getplasticityCorrEachGene(.gene_idxs = test_idxs, .experiment_name = "Heat")
test_plotmat_n2 <- ceiling(sqrt(length(test_cors2)))
test_plotmat2 <- matrix(test_cors2[1:(test_plotmat_n2*test_plotmat_n2)], 
                       nrow = test_plotmat_n2, ncol = test_plotmat_n2)

# plotting
test_plotmat_full <- cbind(test_plotmat, test_plotmat2)
col_fun = colorRamp2(c(-1, 0, 1), c("red2", "white", "skyblue"))
Heatmap(test_plotmat_full, col = col_fun,
        row_order = c(1:nrow(test_plotmat_full)),
        column_order =  c(1:ncol(test_plotmat_full)),
        na_col = "grey30")
```

Environment-specificity of plasticity divergence is driven by the strongest divergers:

```{r giant-plasticity-heatmap-strongestDiverged}
plotlist <- vector(mode = "list", length = length(ExperimentNames))
names(plotlist) <- ExperimentNames
plotlist_names <- vector(mode = "list", length = length(ExperimentNames))
names(plotlist_names) <- ExperimentNames
for (e_detect in ExperimentNames) {
  e_idxs <- finaldf |> filter(experiment == e_detect &
                                   plasticity == "diverged" &
                                   cor_parents < -0.5 &
                                   cer != 0 & par != 0) |> 
    dplyr::select(gene_name) |> pull()
  e_plotmat_n <- ceiling(sqrt(length(e_idxs)))
  plot_mat_row <- matrix(nrow = e_plotmat_n, ncol = 0)
  plotlist_names[[e_detect]] <- matrix(e_idxs[1:(e_plotmat_n*e_plotmat_n)], 
                                       nrow = e_plotmat_n, ncol = e_plotmat_n)
  for (e_measure in ExperimentNames) {
    e_cors <- getplasticityCorrEachGene(.gene_idxs = e_idxs, 
                                        .experiment_name = e_measure)
    e_plotmat <- matrix(e_cors[1:(e_plotmat_n*e_plotmat_n)], 
                           nrow = e_plotmat_n, ncol = e_plotmat_n)
    plot_mat_row <- cbind(plot_mat_row, e_plotmat)
  }
  plotlist[[e_detect]] <- plot_mat_row
}
# plotting
col_fun = colorRamp2(c(-1, 0, 1), c("red2", "white", "skyblue"))
for (e_detect in names(plotlist)) {
  print(Heatmap(plotlist[[e_detect]], col = col_fun,
                row_order = rownames(plotlist[[e_detect]]),
                row_labels = rownames(plotlist[[e_detect]]),
                column_order =  c(1:ncol(plotlist[[e_detect]])),
                show_heatmap_legend = FALSE,
                na_col = "grey30"))
  #print(plotlist_names[[e_detect]]) # to ID specific genes of interest
  }
```

Weaker diverged (cor parents >= -0.5):

```{r giant-plasticity-heatmap-weakerDiverged}
plotlist <- vector(mode = "list", length = length(ExperimentNames))
names(plotlist) <- ExperimentNames
plotlist_names <- vector(mode = "list", length = length(ExperimentNames))
names(plotlist_names) <- ExperimentNames
for (e_detect in ExperimentNames) {
  e_idxs <- finaldf |> filter(experiment == e_detect &
                                   plasticity == "diverged" &
                                   cor_parents >= -0.5 &
                                   cer != 0 & par != 0) |> 
    dplyr::select(gene_name) |> pull()
  e_plotmat_n <- ceiling(sqrt(length(e_idxs)))
  plot_mat_row <- matrix(nrow = e_plotmat_n, ncol = 0)
  plotlist_names[[e_detect]] <- matrix(e_idxs[1:(e_plotmat_n*e_plotmat_n)], 
                                       nrow = e_plotmat_n, ncol = e_plotmat_n)
  for (e_measure in ExperimentNames) {
    e_cors <- getplasticityCorrEachGene(.gene_idxs = e_idxs, 
                                        .experiment_name = e_measure)
    e_plotmat <- matrix(e_cors[1:(e_plotmat_n*e_plotmat_n)], 
                           nrow = e_plotmat_n, ncol = e_plotmat_n)
    plot_mat_row <- cbind(plot_mat_row, e_plotmat)
  }
  plotlist[[e_detect]] <- plot_mat_row
}
# plotting
col_fun = colorRamp2(c(-1, 0, 1), c("red2", "white", "skyblue"))
for (e_detect in names(plotlist)) {
  print(Heatmap(plotlist[[e_detect]], col = col_fun,
                row_order = rownames(plotlist[[e_detect]]),
                row_labels = rownames(plotlist[[e_detect]]),
                column_order =  c(1:ncol(plotlist[[e_detect]])),
                show_heatmap_legend = FALSE,
                na_col = "grey30"))
  #print(plotlist_names[[e_detect]]) # to ID specific genes of interest
}
```

Do those genes with a low parental cor in at least one other environment have more hybrid plasticity divergence?

```{r multi-environment-plasticity-divergence}
# TODO: pivot_wider on finaldf and sort genes that have strong plastictiy divergence
# in each environment (cor_parents < -0.5) and also strong plasticity divergence in at least
# one other environment. Then see if hybrid cor is lower for these genes vs
# ones that don't have that multi-environment divergence
plotdf <- finaldf |> filter(plasticity == "diverged" & 
                              cer != 0 & par != 0 & 
                              cor_parents < -0.5) |> 
  dplyr::select(gene_name, experiment) |> 
  dplyr::rename(c("detection_environment"="experiment")) |> 
  left_join(y = finaldf |> dplyr::select(gene_name, experiment, cor_parents, cor_hybrid, cer, par) |> 
              pivot_wider(id_cols = "gene_name", names_from = "experiment",
                          values_from = c("cor_parents", "cor_hybrid", "cer", "par")),
            by = "gene_name")
plotdf$nEnv_plastdiv <- purrr::map(c(1:nrow(plotdf)), \(i) {
  nEnv <- 1 # all these genes have plasticity divergence in at least one environment
  e_detect <- plotdf$detection_environment[i]
  for (e_test in setdiff(ExperimentNames, e_detect)) {
    cor_test <- plotdf[i,] |> dplyr::select(all_of(paste0("cor_parents_", e_test)))
    cer_test <- plotdf[i,] |> dplyr::select(all_of(paste0("cer_", e_test)))
    par_test <- plotdf[i,] |> dplyr::select(all_of(paste0("par_", e_test)))
    if (!is.na(cor_test)) {
      if (cor_test < -0.5 & (cer_test != par_test) & (cer_test != 0) & (par_test != 0)) {
        nEnv <- nEnv + 1
      }
    }
  }
  return(nEnv)
}) |> unlist()
plotdf$detection_cor_parents <- purrr::map(c(1:nrow(plotdf)), \(i) {
  e <- plotdf$detection_environment[i]
  return(dplyr::select(plotdf[i,], all_of(paste0("cor_parents_", e))))
}) |> unlist()
plotdf$min_cor_hybrid <- purrr::map(c(1:nrow(plotdf)), \(i) {
  out_mean <- dplyr::select(plotdf[i,], all_of(paste0("cor_hybrid_", ExperimentNames))) |> unlist() |> min(na.rm = TRUE)
  return(out_mean)
}) |> unlist()
plotdf |> mutate(high_cor_hybrid =  min_cor_hybrid > 0) |> 
  dplyr::select(nEnv_plastdiv, high_cor_hybrid) |> table()
```

Regardless of detection environment, do the genes with multi-environment plasticity divergence have higher or lower cor_hybrid?
```{r multi-environment-plasticity-divergence-plotting}
ggplot(unique(dplyr::select(plotdf, gene_name, nEnv_plastdiv, min_cor_hybrid)), 
       aes(x = factor(nEnv_plastdiv), y = min_cor_hybrid)) +
  geom_violin(aes(fill = factor(nEnv_plastdiv))) +
  geom_jitter()
```

If anything, hybrid alleles have stronger correlation for the genes that are diverging in plasticity in multiple environments.

What are these 6 genes?

```{r multienv-plasticity-avg-expr, warning=FALSE}
gene_idxs <- plotdf |> filter(nEnv_plastdiv == 3) |> dplyr::select(gene_name) |> pull() |> unique()
#plotlist <- vector(mode = "list", length = length(gene_idxs))
#names(plotlist) <- gene_idxs
for (g in gene_idxs) {
  #plotlist[[g]] <- plotEnvironments(.gene_idxs = g, .quartet = TRUE)
  print(plotEnvironments(.gene_idxs = g, .quartet = TRUE))
}
```

YKL127W (PGM1) is the only one that looks to have cis-diverging plasticity in multiple environments:

```{r PGM1-YKL127W}
plotEnvironments(.gene_idxs = "YKL127W", .quartet = TRUE)
```

### Exploration: Diagonal is still the most uncorrelated when taking average cor of each ortholog pair

```{r getplasticityCorrPerGene}
# (as opposed to one cor of average expr of all divergent genes)
# But now the correlations are much weaker magnitude b/c most single genes are weakly correlated moving the average towards 0
getplasticityCorrPerGene <- function(.gene_idxs, .experiment_name) {
  # filter for genes expressed high enough in this environment
  good_expr_genes <- finaldf |> filter(experiment == .experiment_name) |> 
    dplyr::select(gene_name) |> pull()
  .gene_idxs <- .gene_idxs[.gene_idxs %in% good_expr_genes]
  # calculating correlation of each gene btwn parental orthologs
  condition_vec <- info |> filter(experiment == .experiment_name) |> 
    dplyr::select(condition) |> pull()
  cer_mat <- collapsed$cer[.gene_idxs, condition_vec]
  par_mat <- collapsed$par[.gene_idxs, condition_vec]
  cor_vec <- map(.gene_idxs, \(g) {
    return(cor(cer_mat[g,], par_mat[g,]))
  }) |> unlist()
  return(mean(cor_vec, na.rm = TRUE)) # NAs can arise from one species' counts being all 0s
}
# tests for getplasticityCorrPerGene
# HAP4 all diverged plasticity
gene_idxs <- finaldf |> filter(experiment == "HAP4" &
                                 plasticity == "diverged" & 
                                 cer != 0 & par != 0) |> 
  dplyr::select(gene_name) |> pull()
getplasticityCorrPerGene(gene_idxs, .experiment_name = "HAP4")
getplasticityCorrPerGene(gene_idxs, .experiment_name = "LowN")
getplasticityCorrPerGene(gene_idxs, .experiment_name = "LowPi")
getplasticityCorrPerGene(gene_idxs, .experiment_name = "CC")
getplasticityCorrPerGene(gene_idxs, .experiment_name = "Heat")
getplasticityCorrPerGene(gene_idxs, .experiment_name = "Cold")

# all plasticity divergers
plot_mat <- matrix(nrow = length(ExperimentNames),
                   ncol = length(ExperimentNames))
for (e_row in ExperimentNames) {
  for (e_col in ExperimentNames) {
    e_gene_idxs <- finaldf |> filter(experiment == e_row &
                                       plasticity == "diverged" & 
                                       cer != 0 & par != 0) |> 
      dplyr::select(gene_name) |> pull()
    e_cor <- getplasticityCorrPerGene(e_gene_idxs, .experiment_name = e_col)
    plot_mat[which(ExperimentNames == e_row),
             which(ExperimentNames == e_col)] <- e_cor
  }
}
colnames(plot_mat) <- LongExperimentNames
rownames(plot_mat) <- LongExperimentNames

# plotting
col_fun = colorRamp2(c(0, 0.25, 0.5), c("red2", "white", "skyblue"))
Heatmap(plot_mat, col = col_fun,
        row_order = LongExperimentNames, column_order = LongExperimentNames,
        row_names_side = "left", heatmap_legend_param = list(title = ""),
        cell_fun = function(j, i, x, y, width, height, fill) {
          grid.text(sprintf("%.2f", plot_mat[i, j]), x, y, gp = gpar(fontsize = 10))
        })
```


### Do genes that are diverging in plasticity tend to have less similar expresison in the beginning of the experiment and more similar at the end?

```{r example-criss-cross, warning=FALSE}
# example gene first, 2-1 HAP4
gene_idx <- finaldf |> filter(experiment == "HAP4" & 
                                level == "conserved" &
                                cer == 2 & par == 1) |> 
  dplyr::select(gene_name) |> pull() |> sample(1)
# plotEnvironments(.gene_idxs = "YMR221C") # first example we found, strong CC level divergence
plotEnvironments(.gene_idxs = gene_idx)
```

```{r getTPAvgExpr}
getTPAvgExpr <- function(.gene_name, .experiment, .organism,
                          .first_or_last = "first") {
  if (.organism == "cer") {
    cts_mat <- collapsed$cer
    info_df <- info
  }
  if (.organism == "par") {
    cts_mat <- collapsed$par
    info_df <- info
  }
  if (.first_or_last == "first") {
    tp0_condition <- info_df |> filter(experiment %in% .experiment) |> 
      filter(time_point_num == min(time_point_num)) |> 
      dplyr::select(condition) |> pull()
    mean_expr <- cts_mat[.gene_name, tp0_condition] |> mean()
    return(mean_expr)
  }
  if (.first_or_last == "last") {
    tpLast_condition <- info_df |> filter(experiment == .experiment) |> 
      filter(time_point_num == max(time_point_num)) |> 
      dplyr::select(condition) |> pull()
    mean_expr <- cts_mat[.gene_name, tpLast_condition] |> mean()
    return(mean_expr)
  }
}
# tests for getTPAvgExpr
getTPAvgExpr(.gene_name = "YGR192C", .experiment = "HAP4",
              .organism = "cer")
getTPAvgExpr(.gene_name = "YGR192C", .experiment = "HAP4",
              .organism = "par")
getTPAvgExpr(.gene_name = "YGR192C", .experiment = setdiff(ExperimentNames, "HAP4"),
              .organism = "cer")
getTPAvgExpr(.gene_name = "YGR192C", .experiment = setdiff(ExperimentNames, "HAP4"),
              .organism = "par")
getTPAvgExpr(.gene_name = "YGR192C", .experiment = "HAP4",
              .organism = "cer", .first_or_last = "last")
getTPAvgExpr(.gene_name = "YGR192C", .experiment = "HAP4",
              .organism = "par", .first_or_last = "last")
```

Getting average expression of each plasticity-diverging gene in the environment they were detected in:

```{r tp0-tplast-home, cache=TRUE}
tpdiffdf <- finaldf |> filter(plasticity == "diverged" & level == "conserved")
tpdiffdf$tp0_cer <- purrr::map(c(1:nrow(tpdiffdf)), .f = \(i) {
  getTPAvgExpr(.gene_name = tpdiffdf$gene_name[i],
                .experiment = tpdiffdf$experiment[i],
                .organism = "cer",
                .first_or_last = "first")
}) |> unlist()
tpdiffdf$tp0_par <- purrr::map(c(1:nrow(tpdiffdf)), .f = \(i) {
  getTPAvgExpr(.gene_name = tpdiffdf$gene_name[i],
                .experiment = tpdiffdf$experiment[i],
                .organism = "par",
                .first_or_last = "first")
}) |> unlist()
tpdiffdf$tplast_cer <- purrr::map(c(1:nrow(tpdiffdf)), .f = \(i) {
  getTPAvgExpr(.gene_name = tpdiffdf$gene_name[i],
                .experiment = tpdiffdf$experiment[i],
                .organism = "cer",
                .first_or_last = "last")
}) |> unlist()
tpdiffdf$tplast_par <- purrr::map(c(1:nrow(tpdiffdf)), .f = \(i) {
  getTPAvgExpr(.gene_name = tpdiffdf$gene_name[i],
                .experiment = tpdiffdf$experiment[i],
                .organism = "par",
                .first_or_last = "last")
}) |> unlist()
```

Calling whether tp0 or tplast had the bigger difference in expression between species:

```{r diff-lessening-or-strengthening}
tpdiffdf$diff_tp0 <- abs(tpdiffdf$tp0_cer - tpdiffdf$tp0_par)
tpdiffdf$diff_tplast <- abs(tpdiffdf$tplast_cer - tpdiffdf$tplast_par)
sum(tpdiffdf$diff_tp0 == 0)
sum(tpdiffdf$diff_tplast == 0)
getFoldTPlast <- function(.diff_tp0, .diff_tplast) {
  if (.diff_tp0 == 0) {
    return(.diff_tplast)
  }
  if (.diff_tplast == 0) {
    return(0)
  }
  return(.diff_tplast/.diff_tp0)
}
tpdiffdf$fold_tplast <- purrr::map2(.x = tpdiffdf$diff_tp0,
                                  .y = tpdiffdf$diff_tplast, 
                                  .f = getFoldTPlast) |> 
  unlist()
tpdiffdf$diff_is_strengthening <- tpdiffdf$fold_tplast > 2
tpdiffdf$diff_is_lessening <- tpdiffdf$fold_tplast < 0.5
tpdiffdf |> 
  dplyr::select(diff_is_strengthening, diff_is_lessening) |> 
  table()
```
```{r saving-tpdiffdf}
tpdiffdf$diff_type = if_else(tpdiffdf$diff_is_strengthening,
                             true = "strengthen",
                             false = if_else(tpdiffdf$diff_is_lessening,
                                             true = "lessen",
                                             false = "neutral"))
save(tpdiffdf, file = "data_files//TP_FirstAndLast_Difference.RData")
```

Example plasticity-diverging gene where the difference between species lessens over time

```{r example-diff-lessening, warning=FALSE}
# example: HAP4
gene_idx <- tpdiffdf |> filter(experiment == "HAP4" &
                                diff_is_lessening) |> 
  dplyr::select(gene_name) |> pull() |> sample(1)
# plotEnvironments(.gene_idxs = "YMR221C") # first example we found, strong CC level divergence
plotGenes(.gene_idxs = gene_idx, .experiment_name = "HAP4")
```

Example plasticity-diverging gene where the difference between species strengthens over time

```{r example-diff-strengthening, warning=FALSE}
# example: HAP4
gene_idx <- tpdiffdf |> filter(experiment == "HAP4" &
                                diff_is_strengthening) |> 
  dplyr::select(gene_name) |> pull() |> sample(1)
# plotEnvironments(.gene_idxs = "YMR221C") # first example we found, strong CC level divergence
plotGenes(.gene_idxs = gene_idx, .experiment_name = "HAP4")
```


