---
title: 'QC: Signal to Noise vs Dispersion'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r environment-setup}
source("functions_for_figure_scripts.R")
library("pheatmap")
library("ggalluvial")
load("data_files/FinalDataframe3Disp.RData")
finaldf_Disp3 <- finaldf
rm(finaldf)
load("data_files/FinalDataframe_SNR30pct.RData")
finaldf_SNR30pct <- finaldf
rm(finaldf)
snrdf <- read_csv("data_files/QC_signal-to-noise.csv")
```
Discrete heatmaps of genes assigned to 0 cluster using different thresholds/criteria.

Comparing the main paper's cluster criterion of Disp = 3 versus each version of signal to noise to find the one that is most similar to Disp = 3 assignments

```{r discrete-heatmap-lowly-varying}
plot_mat <- finaldf_Disp3 |> dplyr::select(gene_name, experiment, cer, par) |> 
  pivot_longer(cols = c("cer", "par"),
               names_to = "species",
               values_to = "Disp3") |> 
  left_join(snrdf, by = c("gene_name"="gene_ID", "experiment", "species"))
colnames_plotmat <- paste(plot_mat$gene_name, plot_mat$species, 
                          plot_mat$experiment, sep = "_")
rownames_plotmat <- colnames(plot_mat)
plot_mat <- data.frame(plot_mat) |> t()
colnames(plot_mat) <- colnames_plotmat
rownames(plot_mat) <- rownames_plotmat
plot_mat <- plot_mat[!(rownames(plot_mat) %in% c("gene_name", "experiment", "species")),]
plot_mat <- plot_mat |> as.matrix()
mode(plot_mat) <- "numeric"
plot_mat[is.na(plot_mat)] <- 1 # currently only 1 NA value: YPR148C, Heat, cer, SNR 1pct
plot_mat[plot_mat != 0] <- 1
clrs <- structure(c("grey", "red"), 
                  names = c(1, 0))
# sorting cols by Disp3's order
col_order_vec <- order(factor(plot_mat["Disp3",], levels = c(1, 0)))

# ordering full plotmat
ordered_plot_mat <- orderGenesByGroup(.mat = plot_mat[c("Disp3", "SNR_30pct"),], .labels = c(1, 0))
rownames(ordered_plot_mat) <- c("Disp3", "SNR_30pct")
pheatmap(ordered_plot_mat, cluster_rows = FALSE, cluster_cols = FALSE, show_colnames = FALSE)
```

4-group comparison between Disp3 and signal-to-noise 30%:

```{r discrete-heatmap-group4, cache=TRUE}
plotdf <- finaldf_Disp3 |> dplyr::select(gene_name, experiment, group4, cer, par) |> 
  dplyr::rename(c("group4_disp3"="group4",
                  "cer_disp3"="cer",
                  "par_disp3"="par"))
plotdf <- left_join(plotdf, select(finaldf_SNR30pct, gene_name, experiment, group4, cer, par), by = c("gene_name", "experiment")) |> 
  dplyr::rename(c("group4_SNR30pct"="group4",
                  "cer_SNR30pct"="cer",
                  "par_SNR30pct"="par"))
plot_mat <- select(plotdf, group4_disp3, group4_SNR30pct) |> t() |> as.matrix()
ordered_plot_mat <- orderGenesByGroup(.mat = plot_mat)
rownames(ordered_plot_mat) <- c("Dispersion 3", "Signal-to-noise 30%")
plotdf <- map(c("Dispersion 3", "Signal-to-noise 30%"), \(e) {
  idx <- which(c("Dispersion 3", "Signal-to-noise 30%") == e)
  return(tibble(gene_name = colnames(ordered_plot_mat),
                environment = e,
                group4 = ordered_plot_mat[idx,],
                x = c(1:ncol(ordered_plot_mat)),
                y = c(length(LongExperimentNames):1)[idx])) # descending from top
}) |> purrr::reduce(.f = bind_rows)
p <- ggplot(plotdf, aes(x = x, y = y, fill = group4)) +
  geom_tile_pattern(aes(fill = group4, pattern = group4), pattern_fill = "black",
                    pattern_density = 0.05, pattern_size = 0.05) +
  scale_fill_discrete(limits = colordf[colordf$scheme == "group4",]$limits,
                      type = colordf[colordf$scheme == "group4",]$type) +
  scale_pattern_discrete(limits = colordf[colordf$scheme == "group4",]$limits,
                         choices = colordf[colordf$scheme == "group4",]$pattern) +
  theme_void()
p
```

What types of plasticity divergence are the most different between dispersion and SNR?

```{r plasticity-groups}
plotdf <- finaldf_Disp3 |> dplyr::select(gene_name, experiment, group4, cer, par) |> 
  dplyr::rename(c("group4_disp3"="group4",
                  "cer_disp3"="cer",
                  "par_disp3"="par"))
plotdf <- left_join(plotdf, select(finaldf_SNR30pct, gene_name, experiment, group4, cer, par), by = c("gene_name", "experiment")) |> 
  dplyr::rename(c("group4_SNR30pct"="group4",
                  "cer_SNR30pct"="cer",
                  "par_SNR30pct"="par"))
plotlist <- vector(mode = "list", length = length(ExperimentNames))
names(plotlist) <- ExperimentNames
for (e in ExperimentNames) {
  plot_mat_e <- plotdf |> filter(experiment == e) |> 
    mutate(disp3 = paste(cer_disp3, par_disp3, sep = "_"),
           SNR30pct = paste(cer_SNR30pct, par_SNR30pct, sep = "_")) |> 
    dplyr::select(disp3, SNR30pct) |> t() |> as.matrix()
  ordered_plot_mat_e <- orderGenesByGroup(.mat = plot_mat_e,
                                          .labels = unique(c(plot_mat_e[1,], plot_mat_e[2,])))
  rownames(ordered_plot_mat_e) <- c("Dispersion 3", "Signal-to-noise 30%")
  plotdf_e <- map(c("Dispersion 3", "Signal-to-noise 30%"), \(nm) {
    idx <- which(c("Dispersion 3", "Signal-to-noise 30%") == nm)
    return(tibble(gene_name = colnames(ordered_plot_mat_e),
                  method = nm,
                  clusters = ordered_plot_mat_e[idx,],
                  x = c(1:ncol(ordered_plot_mat_e)),
                  y = c(length(LongExperimentNames):1)[idx])) # descending from top
  }) |> purrr::reduce(.f = bind_rows)
  plotlist[[e]] <- ggplot(plotdf_e, aes(x = x, y = y, fill = clusters)) +
    geom_tile(aes(fill = clusters)) +
    theme_void()
}
ggarrange(plotlist = plotlist, nrow = 3, ncol = 2, common.legend = TRUE)
```

Alluvial plots of which of 4 plasticity categories are different between dispersion and signal to noise:
1) single 0s: 10, 01, 20, 02
2) conserved 0s: 00
3) reversals: 12, 21
4) conserved varying: 11, 22

```{r alluvial}
# helper function to collapse 9 plasticity combinations into 4 categories we are interested in here
getPlasticityCategory <- function(.cer, .par) {
  if (.cer == 0 & .par == 0) {
    return("ConsZero")
  }
  if (.cer == 0 | .par == 0) {
    return("SingleZero")
  }
  if (.cer  != .par) {
    return("Reversal")
  }
  if (.cer == .par) {
    return("ConsVar")
  }
}
# tests for getPlasticityCategory
getPlasticityCategory(0, 1)
getPlasticityCategory(0, 0)
getPlasticityCategory(0, 56)
getPlasticityCategory("bugs", 1)
getPlasticityCategory(1, 1)

plotdf$plasticity_category_disp3 <- purrr::map2(plotdf$cer_disp3, 
                                                plotdf$par_disp3, .f = getPlasticityCategory) |> unlist()
plotdf$plasticity_category_SNR30pct <- purrr::map2(plotdf$cer_SNR30pct, 
                                                   plotdf$par_SNR30pct, .f = getPlasticityCategory) |> unlist()
plotdf_counts <- plotdf |> 
  group_by(experiment, plasticity_category_disp3, plasticity_category_SNR30pct) |> 
  summarise(ngenes = n()) |> 
  ungroup()
category_counts <- plotdf |> 
  group_by(experiment, plasticity_category_disp3) |> 
  summarise(ncategory = n()) |> 
  ungroup()
plotdf_counts <- left_join(plotdf_counts, category_counts,
                           by = c("plasticity_category_disp3", "experiment"),
                           relationship = "many-to-one")
plotdf_counts$pctcategory <- plotdf_counts$ngenes/plotdf_counts$ncategory
is_alluvia_form(plotdf_counts, silent = TRUE)
p <- ggplot(filter(plotdf_counts, plasticity_category_disp3 != plasticity_category_SNR30pct),
       aes(y = ngenes, axis1 = plasticity_category_disp3, axis2 = plasticity_category_SNR30pct)) +
  geom_alluvium(aes(fill = plasticity_category_disp3), width = 1/12) +
  geom_stratum(width = 1/12, fill = "black", color = "grey") +
  geom_label(stat = "stratum", aes(label = after_stat(stratum))) +
  scale_x_discrete(limits = c("Disp 3", "SNR 30%"), expand = c(.05, .05)) +
  scale_fill_brewer(type = "qual", palette = "Set1") +
  facet_wrap(~experiment) +
  theme_classic() +
  ylab("number of genes")
p
```


