---
title: 'QC: comparing lowly-varying gene detection criteria, signal-to-noise versus dispersion'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Goal: output a table of cluster assignments that use SNR instead of dispersion as the criteria for lowly-varying genes. This table will be loaded into the non-SNR branch of the repository to compare clustering results to dispersion.

```{r environment-setup}
source("functions_for_figure_scripts.R")
load("data_files/FinalDataframe_SNR5pct.RData")
load("data_files/QC_CorrelationClustering_SignalToNoise.RData")
```

Making outdf with one column for each SNR threshold. Number of rows = nSpecies x nGenes x nEnvironments (in contrast to finaldf, where species have separate columns)

```{r outdf}
outdf <- list(clusterdf |> 
                pivot_longer(cols = c("cer", "par"),
                             names_to = c("species"),
                             values_to = c("SNR_1")),
              clusterdf_snr0.1 |> 
                pivot_longer(cols = c("cer", "par"),
                             names_to = c("species"),
                             values_to = c("SNR_0.1")),
              clusterdf_snr0.5 |> 
                pivot_longer(cols = c("cer", "par"),
                             names_to = c("species"),
                             values_to = c("SNR_0.5")),
              clusterdf_snr30pct |> 
                pivot_longer(cols = c("cer", "par"),
                             names_to = c("species"),
                             values_to = c("SNR_30pct")),
              clusterdf_snr5pct |> 
                pivot_longer(cols = c("cer", "par"),
                             names_to = c("species"),
                             values_to = c("SNR_5pct")),
              clusterdf_snr10pct |> 
                pivot_longer(cols = c("cer", "par"),
                             names_to = c("species"),
                             values_to = c("SNR_10pct"))
) |> purrr::reduce(.f = left_join, by = c("gene_ID", "experiment", "species"))
```

Saving

```{r saving}
write_csv(outdf, file = "data_files/QC_signal-to-noise.csv")
```


